%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                         M E M O    T E M P L A T E                         %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% RCS Information:
% $Id$


\documentclass[12pt]{article}
\pagestyle{plain}

\setlength{\topmargin}			{-20mm}
\setlength{\headsep}			{0mm}
\setlength{\textheight}			{245mm}
\setlength{\textwidth}			{170mm}
\addtolength{\oddsidemargin}		{-17mm}
\setlength{\parskip}			{0.8ex}
\setlength{\parindent}			{0.0ex}
\renewcommand{\baselinestretch}	{1.0}

\newenvironment{memohead}{
	\noindent
	\begin{list}%
	{ }
	{
	\setlength{\parskip}{0ex}
	\setlength{\partopsep}{0ex}
	\setlength{\topskip}{0ex}
	\setlength{\labelwidth}{5em}
	\setlength{\labelsep}{1em}
	\setlength{\leftmargin}{6em}
	}
}{\end{list}}
	
\def\undertext#1{$\underline{\smash{\hbox{#1}}}$}

\begin{document}

\noindent
\rule{\textwidth}{0.25mm}
\begin{center}
\bf DBCE INTERNAL MEMORANDUM
\end{center}
\begin{memohead}{}
\item[\bf To:\hfill]
Jeff Symons 
\item[\bf From:\hfill]
Hugh Blackburn
\item[\bf Subject:\hfill]
Capital Expenditure Request 
\item[\bf Date:\hfill]
31 August 1999
\end{memohead}
\noindent
\rule{\textwidth}{0.25mm}
\vspace*{1ex}

% -- Start text after this line.

\thispagestyle{empty}

Dear Jeff,

In our last round of project review for CZ30 `Complex Flow Simulation'
we discussed the \$25k capital approved in the project bid for a `new
workstation'.  You may recall we also discussed the future of high
performance computing within CSIRO, and the possible eventual
devolution of the associated costs to project level.  I believe that
the most likely future of high performance computing lies in
commodity-hardware parallel machines (e.g.\ networked PC-style
machines running Linux).  \textbf{I would like to divert the funds
approved for a workstation to buying a multinode machine of this
kind.}  Here are some supporting reasons:
\begin{enumerate}
\item
Machines of this kind represent the future, and we need to gain
experience with setting up and running them --- since they are
essentially supported by public-domain software, this is a significant
aspect of starting up;
\item
Andrew Brydon, currently working for us, has already set up such a
machine at Monash Department of Mathematics, so we have ready access
to appropriate expertise;
\item
With the increasing use of the CSIRO/BOM NEC SX-4 by this division for
running commercial codes within the SFT project, we are experiencing
greatly reduced share of the machine for LES-related research.  We can
significantly reduce our dependence on this machine for the
outlay of \$25k;
\item
The machine can be expanded in the future at low incremental costs.
\end{enumerate}

We have examined various alternative configurations and run some
benchmarks with my code on different machines.  We believe that the
base node selected (an Alpha-chipped motherboard) currently gives us
the best FLOP/\$ value --- the fact that it uses the latest chip
release implies we will be able to expand the machine over the next
few years if needs be.  I have attached a Capital Expenditure
Authorisation Request form and three quotes for similar machines.
Compaq's quote is the best, and we can buy 4 nodes for our \$25k
outlay.

Thank you,


\vspace{20mm}
\undertext{Hugh Blackburn}

\end{document}


