% $Id$
% CHAPTER FOUR
% 5977 wds at 29/4/99

\setcounter{endnote}{0}

\chapter{Australian Mainland 1835--1855}
\label{ch:mainland} \addtoendnotes{\protect\subsection*{Chapter \thechapter}}
\markboth{Chapter \thechapter. Australian Mainland}%
{Pioneering Irrigation in Australia}

At the beginning of this period, British settlement on the mainland
was confined to nineteen counties of New South Wales with 39
million\,ac (61\,000 sq mi)\,---\,not three times the area of
Tasmania\,---\,and ten counties of the smaller area occupied in Western
Australia.  From 1835 the initiatives at several places on the coast,
leading to permanent occupation with land sales, gave scope for
agriculture while vast inland tracts of New South Wales were being
taken for pastoral runs by numerous squatters without secure rights of
occupation.

Until 1851 the territory later known as Victoria was identified as the
Port Phillip District of New South Wales, the colony so affected by
wool production that its number of sheep increased to 13 million in
1850.  At that time, the South Australian colony, begun in 1836, held
only a million sheep.  Queensland was not separated from New South
Wales until 1859.\fn{\citet[p.\,137]{fitzpatrick1941}.}

Irrigation was a matter almost ignored by officials and landholders.
For pastoralists the main problems then were to secure herdsmen and
water for livestock; farmers in most districts were more concerned
with market prospects than with lack of rain for crops.  The colony to
record most instances of irrigation was South Australia where many
settlers were intent on farming yet found themselves in a tract of
land not well endowed with rainfall.

This chapter considers both the evidence of irrigation undertaken and
the views of prominent observers on the need and scope for irrigation.

\section*{New South Wales and Victoria}
\index{New South Wales} \index{Victoria}

There was movement inland by pastoralists from the nineteen counties
administered from Sydney \index{Sydney} and from Port Phillip
\index{Port Phillip} and other places on the south coast.  Sheep and
cattle were soon grazing in semi-arid country with palatable herbage
and major streams.  Graziers liked to select sites for their
headquarters close to rivers but above flood level and thus might
irrigate small areas by pumping.  The only person known to do this was
J.\,L.~Phelps \index{Phelps, J.\,L.} of Canally,
\index{Canally} a
pastoral station on the Murrumbidgee River \index{river!Murrumbidgee}
near its junction with the Murray. \index{river!Murray} His practice
was reported by Nehemiah Bartley \index{Bartley, N.} who spent a few
weeks at Canally in November 1853:
\begin{quote}
	Mr John Lecky Phelps of Canally\ldots was a man much in
	advance of his time.  While other people, for 150 miles
	around, had no vegetables, he cultivated a half-acre on the
	river bank, with potatoes, green peas, French beans, cabbages,
	and kept it irrigated by a very simple process for rain was
	uncertain in that far inland spot.  He had a Californian
	wooden pump, about six inches square, with its end fixed in
	the river, and about 150 feet of `Osnaburg' hose from it to
	the top of the garden which was, perhaps, three feet higher
	then the lower end by the river and, half an hour of hand
	pumping every morning, sent the water flowing zig zag,
	backwards and forwards, and in and out, through all the
	well-kept furrows and beds of the enclosure, and the
	vegetation was always fresh and green at Canally
	garden.\fn{\citet[p.\,62]{bartley1892}.}
\end{quote}

Irrigation was certainly used by the late 1840s at Buninyong,
\index{Buninyong} 56~miles
west of Melbourne.  This was the base for the large area held by the
Learmonth brothers \index{Learmonth Bros.} and at the time it was the
only inland township, though unofficial, south of the River Murray.
Among its facilities were an inn and a boarding school.

Arthur Cotton, \index{Cotton, A.} the irrigation engineer who made two
visits from India to Tasmania during 1838--43, came to Buninyong in
1848 as guest of the Learmonths, his brothers-in-law.  Wool
\index{wool} prices were then low and many pastoralists had turned to
producing tallow \index{tallow} by boiling down the carcases of their
sheep.  Some cauldrons used in this process could hold 300
sheep.\fn{\citet[p.\,205]{roberts1935}.}  Francis Peter Labilliere,
\index{Labilliere, F.\,P.} an
historian of early Victoria, described the irrigation carried out by
Cotton:
\begin{quote}
	The Messrs Learmonth at Buninyong\ldots annually `boiled down'
	so many of their sheep, that they found that it answered their
	purpose to have a place of their own, instead of sending their
	fat stock, as was generally done, to a public `boiling down'
	establishment.  The author well remembers visiting it a year
	or two before the gold discovery and seeing numbers of sheep
	prepared for the process.  No finer mutton could have existed.
	The workingmen came in and helped themselves to whatever
	joints or slices they liked to take out of the sheep, as they
	hung skinned and ready to be cut up, before being thrown into
	the melting vat.  This was of large size, and in it the meat
	was left till all the tallow was extracted from it.  The offal
	remaining consisted of flesh, and of rich liquid which would
	have made fine soup or jelly.  This was applied to the
	strangest purpose for which such a thing was ever used.
	Major, now Sir Arthur Cotton\,---\,being on furlough from India,
	on a visit to his brothers-in-law the Messrs Learmonth\,---\,tried
	an irrigation experiment at their place on a limited scale.
	He had a dam thrown across a small creek, and with some
	sluices and trenches to convey the water, was able to carry
	the soup or gravy and other offal from the boiling down
	establishment over some of the adjoining land.  Magnificent
	crops were thus produced on some of the fields and splendid
	vegetables and fruit in the
	garden.\fn{\citet[p.\,331]{labilliere1878}.}
\end{quote}

Pastoralists were prohibited from cultivating the land they held under
licence except for their own use, but it is doubtful if this condition
was strictly enforced.  It has been claimed that in the Port Phillip
District `definite records can be cited showing that pastoralists as
long ago as 1842 cultivated and irrigated by the furrow system
considerable areas of land'.\fn{\citet[p.\,166]{billis1930}.}

The probability of irrigation on pastoral stations is suggested also
by experience recorded by William Howitt, the author and traveller who
spent two years in Victoria at the height of the gold rushes.  William
Howitt \index{Howitt, W.} was relatively fortunate during his travels
to various goldfields in having support from Lieutenant-Governor La
Trobe \index{Lieut-Gov. La Trobe} in the form of introductions to
several pastoralists, some of whom supplied vegetables and fruit to
his party.  Prominent among these landholders were the Forlonges
\index{Forlonges} of
Seven Hills \index{Seven Hills} (Euroa) \index{Euroa} in north-eastern
Victoria.  Howitt called at their homestead in the late summer of 1853
when water and fodder for horses were exceedingly scarce along the
main road.  The Forlonges provided plentiful supplies of fruit and
vegetables:
\begin{quote}
	The garden lies on the other side of the creek, in the flat a
	quarter of a mile from the house. It is, I suppose, five or
	six acres in extent and is full of splendid fruit in
	season\,---\,loads of grapes and melons of all kinds
	especially.\fn{\citet[pp.\,146--7]{howitt1885}.}
\end{quote}

This description suggests some supply of water to the garden from the
creek but Howitt made no reference to irrigation there or elsewhere in
his travels.  In another episode, Howitt in October 1853 reported on a
pastoral station near Bendigo with `a fine market for all its
produce':
\begin{quote}
	I have seen a man from this station going through the diggings
	with a horse and cart, the cart piled with cabbages, and a
	cabbage stuck on a broomstick as a sign that they were for
	sale\ldots\fn{\citet[p.\,247]{howitt1885}.}
\end{quote}

The farming carried on in New South Wales, apparently without
irrigation, was undertaken at many places near the coast, from the
Hunter River \index{river!Hunter} in the north to Melbourne,
\index{Melbourne} Geelong, \index{Geelong} 
Portland \index{Portland} and Port Fairy
\index{Port Fairy} in the south, as well as inland at Bathurst
\index{Bathurst} and
Goulburn. \index{Goulburn} Most of the farms involved would have
received rainfall generally sufficient for crops of wheat and hay.

The apparent general lack of attention by landholders to irrigation
occurred despite reference to it in various publications circulating
in Australia.  The earliest of these were the tracts on colonisation
by E.\,G.~Wakefield~(1829, 1834), \index{Wakefield, E.\,G.} who drew
on his experience in Italy to recommend use of irrigation in
Australia.  Thomas Mitchell \index{Mitchell, T.} was the first to
foresee the opportunity for irrigation in eastern Australia, where he
conducted four expeditions.  He saw the scope for irrigation in the
Murray valley during his 1836 expedition from Sydney to Portland on
the Victorian coast. His conclusions, probably composed during a
period of leave in England, show enthusiasm for the use of irrigation:
\begin{quote}
	The Murray, fed by the lofty mountains on the east, carries to
	the sea a body of fresh water, sufficient to irrigate the
	whole country, which is, in general, so level, even to a great
	distance from its banks, that the abundant waters of the
	river, might probably be turned into canals, for the purpose,
	either of supplying deficiencies of natural irrigation at
	particular places, or of affording the means of transport,
	across the wide plains.\fn{\citet[vol.\,II,
	p.\,332]{mitchell1839}.}
\end{quote}

From the last of his expeditions out of Sydney, Mitchell offered
further advice on water conservation and irrigation in the colony of
New South Wales, including the southern portion of what is now
Queensland.  He concluded that rocky ranges he saw `afforded the means
of forming reservoirs of water' to be filled by the erratic but
abundant falls of rain.  `Irrigation\ldots has not been yet attempted;
the natural fertility of the soil has alone been relied on\ldots So
generally available is the structure of the country for the
reservation of water by dams\ldots the undulations of the land present
everywhere facilities for constructing reservoirs, which heavy showers
would fill, and thus afford means sufficient for the purpose of
irrigation were not labour now so scarce there\ldots' He also
considered that lands along the eastern coast, `under a lofty range
which supplies abundance of water for the purposes of irrigation, are
well adapted for the cultivation of cotton and
sugar'.\fn{\citet[pp.\,421--4]{mitchell1848}.}

It is interesting that Mitchell's earlier view of irrigation includes
reference to use of canals for transport but this aspect is ignored in
his later book with, instead, some consideration of railways.  This may
reflect Mitchell's experience of the fervour for canal construction in
England during the 1830s and the growing popularity of railways there
during the 1840s.

Peter Cunningham (1789--1864) \index{Cunningham, P.} spent years at
sea as a naval surgeon and also in the period 1825--1830 acquired more
than 1200\,ac in the Hunter valley.  His experience of Australia is
given in two books published in 1827 and 1841.  The first was issued
before the experience of the serious drought of 1826--29; the second
makes several references to droughts and to irrigation. His practical
interest is shown in the view:
\begin{quote}
	I have little doubt\ldots that many portions of Australia
	which seem to the eye incapable of irrigation by canals from
	the rivers will turn out to the contrary when levels are
	taken; these canals only requiring to be commenced at points
	up the rivers where their beds are higher than the land to be
	irrigated, such canals often running many miles (in irrigating
	countries) before reaching the
	latter.\fn{\citet[p.\,4]{cunningham1841}.}
\end{quote}

The \textsl{Sydney Gazette} referred a letter read to a meeting of the
Legislative Council on 31 August 1841 from a Tasmanian gentleman
concerning his irrigation of 2000\,ac and offered its view that there
was
\begin{quote}
	very little prospect of this or any other useful project being
	set on foot until such time as that element of agricultural
	prosperity is in full operation; we mean an `Agricultural
	Association' without which it is in vain to look for a change
	in the present defective agricultural system as practised in
	this colony.\fn{\textsl{Sydney Gazette}, 2 Sept.\ 1841.}
\end{quote}

No contribution to discussion of irrigation in Australia could have
had wider impact than the book by Strzelecki\index{Strzelecki, P.}  in
1845.  He spent four years (1839--43) in the country during which he
made an exploration including ascent of Australia's highest peak,
studied geology and mineralogy of eastern Australia and Tasmania, made
soil analyses and took an active interest in land use including
irrigation.  The distinction associated with his naming of Mount
Kosciusko \index{Mount Kosciusko} helped to make his book well-known
in Australia.

On return to New South Wales from Tasmania in September 1842,
Strzelecki renewed friendly association with James Macarthur,
\index{Macarthur, J.} his
companion on the exploration in the Australian Alps but then involved
with affairs of the Australian Agricultural Company \index{Australian
Agricultural Co.} which had ambitious plans for land settlement.  He
visited the district north of Sydney where the Company had secured a
large grant of land embracing the valley of the Karua, a stream
discharging near Port Stephens.  \index{Port Stephens} Strzelecki
stayed at Port Stephens with Phillip Parker King, \index{King, P.\,P.}
then in charge of the Company's operations, and for a few months
followed his interests in geology and agriculture before returning to
Sydney early in 1843.\fn{\citet{kaluski1985}.}

Strzelecki must have been impressed with the scope for irrigation in
this northern part of the colony, for he gave it prominence in his
views on irrigation:
\begin{quote}
	In New South Wales, the river Karua, \index{river!Karua} and
	the tributaries of the Hunter, \index{river!Hunter} afford a
	most extensive range for the introduction of irrigation: the
	whole country of Cumberland may also be laid out in irrigated
	lands by means of the Grose \index{river!Grose} and
	Warragambia, \index{river!Warragambia} Hawkesbury
	\index{river!Hawkesbury} and Nepean \index{river!Nepean}
	rivers, and with the aid of cheap wooden aqueducts.  The river
	Nepean for the county of Campden; the Wollondilly,
	\index{river!Wollondilly} for Argyleshire; the river Cox
	\index{river!Cox} for the Vale of Clywd; and the Campbell
	\index{river!Campbell} and Macquarie, \index{river!Macquarie}
	for Bathurst; all offer most valuable water-courses for
	reclaiming and for increasing the productiveness of the
	comparatively sterile lands. The lower portion of Gippsland,
	\index{Gippsland} sheltered as it is, to the northwest and
	westward by the Dividing range, and watered by five fine
	rivers, may be rendered, by irrigation, a most flourishing
	portion of the colony.\fn{\citet[p.\,446--7]{strzelecki1845}.}
\end{quote}

In London during 1853 he became managing director of the Peel River
Land and Mineral Company associated with the Australian Agricultural
Company and thus may have attempted the settlement of migrants on its
Peel River \index{river!Peel}
block.\fn{\citet[p.\,263]{paszkowski1997}.}

Quite apart from all these contributions by men with direct experience
of Australia, there is one instance of an initiative for irrigation
coming from British authorities.  This occurred in 1847, when the
Colonial Secretary, Earl Grey, \index{Grey, Earl} sent documents to
the Australian colonial governors concerning irrigation in Italy and
requested their circulation, apparently to encourage use of the
practice in Australia.  The Governor of New South Wales, Sir Charles
Fitzroy, \index{Fitzroy, Sir\,C.} apparently failed to make a positive
response to the documents.  However, the lieutenant-governors of South
Australia and Tasmania directed publication in government
gazettes.\fn{\citet[p.\,164]{lloyd1988}; SA Government Gazette, 28
Oct.\ 1847; \textsl{Hobart Town Gazette}, 28 Sept.\ 1847, cited by
\citet[p.\,114]{masoncox1994}.}

The initiative in securing the information from Italy \index{Italy}
was taken by Palmerston, then Minister for Foreign Affairs.  The
British Consul in Milan furnished several documents concerning
%various
types of
irrigation and drainage in northern Italy, together with references to
crops involved and the administrative arrangements.  All this material
covered the period 1804--1846.\fn{HRA I, vol.\,XXV, pp.\,457--79.}

Another contribution to the discussion appeared in the
\textsl{Australia Felix Mo\-nthly Magazine}, a short-lived periodical
published in Geelong in 1849.  This was the series of articles
attributed anonymously to `Delta' but actually the work of Arthur
Cotton during\index{Cotton, A.} his last visit to Australia when he
was based at Buninyong, \index{Buninyong} the site of his novel
experiment with irrigation.

Cotton's articles took up the interests of
pastoralists. \index{pastoralists} The first one, in June 1849,
described the lessons on water conservation as he knew it in India
\index{India} and
Tasmania \index{Tasmania} and considered the main needs of
pastoralists: water for livestock, sheep-washes, power supply, and
irrigation.  His only reference to local irrigation concerned its use
in the boiling-down industry\,---\,his own personal experience
mentioned above.  His next two articles, entitled `The Supply of Water
in the Basin of the Murray' were concerned mainly with detailed
measures needed to provide water for pastoralists: especially the
excavation of tanks (dams) and diversion of water from streams.  His
article referred to his own experience in the pastoral country where a
brother-in-law then held the Avoca and Wycheproof run. Irrigation was
not regarded as an immediate issue but the author envisaged that the
Murray Basin would be developed with irrigation of cereals.  This
aspect of water use was to be dealt with in a later contribution, but
there was no further publication of the magazine after October
1849.\fn{\citet[p.\,168]{billis1930}.}

There is at least one other echo of Tasmanian experience of irrigation
in the development of a satisfactory water supply for Melbourne during
the early 1850s.  By 1849, before the discovery of gold, its
population of 15\,000 depended on water taken from the Yarra river
\index{river!Yarra} alongside the town; this was no longer
satisfactory.  The new city surveyor appointed that year had
previously been involved in a company pumping river water; he was now
called on to find a way of improving the water supply.  His proposal
in January 1851 was to conduct water by an aqueduct \index{aqueduct}
from marshes at the headwaters of the Plenty River,
\index{river!Plenty} a tributary to the Yarra, thus ensuring an
unpolluted supply and avoiding the expense of pumping.  His modified
scheme, submitted later that year, incorporated a reservoir to be
located in a natural depression near the headwaters.  This scheme
would provide for a larger population in the future and envisaged
irrigation of several hundred acres from the aqueduct in the
meantime.\fn{Vic. Rept of Select Comm.\ Sewerage and Supply of Water,
1853.}

The surveyor, James Blackburn, \index{Blackburn, J.} had arrived in
Melbourne in 1849 after years of experience as an engineer and
architect in Tasmania, including a period of employment assisting Hugh
Cotton \index{Cotton, H.} with the irrigation survey in the Macquarie
River valley in 1843.  The adoption of his proposal led to
construction of the Yan Yean reservoir.  \index{Yan Yean reservior}
Blackburn's scheme of water conservation may have been inspired by the
basic proposition of Arthur Cotton in Tasmania that the headwaters of
many streams there presented swampy areas which could be turned into
reservoirs; his suggestion regarding irrigation shows an interest also
probably connected with his Tasmanian experience.  Water was brought
to Melbourne from the Yan Yean reservoir in December 1857.

\section*{South Australia} \index{South Australia}

After Charles Sturt \index{Sturt, C.}  explored the River Murray
\index{river!Murray} to its mouth in 1826, his reports encouraged the
establishment of a new distinctive settlement inspired by the views of
E.\,G.~Wakefield,
\index{Wakefield, E.\,G.} promoted by an
unofficial Colonisation Society, and supported by the South Australian
Land Company. \index{South Australian Land Co.} When plans for
settlement were finalised, the arrangements agreed to by the British
government were a departure from those of the penal settlements
elsewhere and were intended to facilitate production of primary
products by free settlers.  The first settlement was made in July 1836
on Kangaroo Island \index{Kangaroo Island} by the South Australian
Land Company to provide a base for whaling, but the lack of adequate
supplies of fresh water forced its abandonment a few years later by
settlers.

Meanwhile the advantages of alternative settlement on the eastern
margin of Gulf St Vincent \index{St Vincent's Gulf} had been
recognised by the official surveyors and the site for Adelaide
\index{Adelaide} had
been chosen after the River Torrens \index{river!Torrens} had been
discovered.  During inspections of the coast in this region, John
Morphett \index{Morphett, J.} in November 1836 recorded impressions of
the coastal plain and the mountain range further inland.  No doubt
influenced to some extent by his four year's residence in Egypt, he
ventured the opinion that most of the plain `is a rich light soil,
wanting nothing but irrigation, during the four or five hottest summer
months, to make it eminently productive all the year round.'  His
interest in irrigation was further shown in remarks on the inland
mountain range:
\begin{quote}
	Mount Lofty bears nearly east, and the whole of this side of
	the range is intersected with gullies, ravines, and water
	courses, of the deepest kind, bearing evident marks of being
	acted on by powerful torrents.  All the hilly country along
	the coast has a similar character, but in no place is it so
	conspicuous as here.  The facilities for damming up, and the
	creation of water power, are greater than I have seen in any
	country in an equal area, and as a probability exists that it
	will be advisable to irrigate during the summer, for the
	second and third crops, this is an inestimable
	advantage.\fn{\citet[p.\,12]{morphett1836}.}
\end{quote}

The River Torrens is very small in comparison with the Murray, but as
the major stream discharging westward from the Mount Lofty
\index{Mount Lofty} Range it
proved invaluable as the source of fresh water for residents of the
township on its banks.  It also provided water for the first
irrigation in the young colony.  This was undertaken not by the
observant John Morphett or his English colleagues but by about 200
German migrants who arrived in November 1838 with August Kavel,
\index{Kavel, A.} their
Lutheran pastor.  They came to South Australia under the patronage of
George Angas, \index{Angas, G.} a major supporter of the new colony
who advanced the money for their transport. On arrival at Port
Adelaide many were suffering from scurvy and all lacked proper shelter
and occupation. George Flaxman, \index{Flaxman, G.} chief clerk to
Angas, accompanied the party to South Australia and took
responsibility for the migrants.  They could not obtain sufficient
employment but Flaxman arranged their lease of an Angas property on
the bank of the Torrens, about 4~miles north-east of Adelaide, where
in December they started building houses and growing much-needed
vegetables and other food.  Their settlement came to be known as
Klemzig. \index{Klemzig}

The Germans at Klemzig were already in debt to Angas for their passage
money and now owed him for the rent of the land; they would also have
to pay dearly for imported cereals and potatoes.  However, their
settlement close to the river gave access to water for domestic use,
gardening, and livestock.  The Torrens flows strongly for only six
months of the year in response to high rainfall over its headwaters;
it then provided a good supply at Klemzig even in summer from `a
series of lakes connected by rivulets which in some places are very
narrow and in others shallow but which always ran briskly'.  At that
time Adelaide had a population greater than 5000 but Ferdinand Kavel,
\index{Kavel, F.} 
one of the Klemzig settlers, found that the English residents were not
agriculturalists and his compatriot August Fiedler \index{Fiedler, A.}
found that `The English had not cultivated any gardens because they
did not like the watering'.\fn{\textsl{South Australian Register}, 30
Jan.\ 1878;
\citet[p.\,96]{schubert1985}.}

Growing vegetables at Klemzig was a necessity to combat the scurvy
contracted during the voyage to Australia.  Whether they carried water
from the river up its steep bank or used a pump is uncertain, but
watering was a time-consuming task after sundown, understandably so
considering that in 1840 the gardens extended over almost 7\,ac. The
settlers were soon raising enough vegetables to sell in Adelaide;
although there was no abundant rain in 1839 until June, production was
maintained by the regular hand-watering.  The adequacy of river water
at Klemzig is indicated by the fact that apart from its use for
gardening, `the women did almost all the washing for the people in the
town and had earned a good deal of money from it'.  Some indication of
the variety of vegetables grown by the settlers in 1839 is given in
the \textsl{Southern Australian} which mentioned lettuces, potatoes,
cucumbers, French, broad, and scarlet beans, carrots, turnips, onions,
radishes, spinach, broccoli, cabbage, and green peas. Clearly these
German migrants were important initiators of irrigated market
gardening, a practice for which another group\,---\,the Chinese\,---\,later
achieved distinction throughout Australia.\fn{\citet[pp.\,88, 97, 113,
131]{schubert1985}.}

There was one German, Johann Menge, \index{Menge, J.} who made a plea
for irrigation in South Australia before the arrival of Kavel's people
and whose helpful association with them extended to their gardening at
Klemzig.  Menge has been described as eccentric, enigmatic and
visionary, but his help to the Klemzig group appears to have been
practical and timely.  In 1838, on leaving employment at Kangaroo
Island with the South Australian Land Company, of which Angas was
director, he sent a scathing letter to the resident manager in which
he recommended irrigation on the island.  He was to have been
responsible to Angas for establishing Kavel's group remote from
Adelaide near the junction of the Darling and the Murray rivers, but
this plan was abandoned.  However, Menge did accompany Pastor
Schurmann, of the Dresden Mission Society, on a visit in October 1838
to the Murray at North-west Bend.  During this or another visit
northerly from Adelaide, Menge was impressed with the quality of
country now identified as the Barossa Valley, and his proposal early
in 1839 for its occupation by Kavel's group was soon realised.  He
lived at Klemzig during 1839 with the Fiedler family, and supplied
vegetable seed gathered from his own garden on Kangaroo Island.  There
appears to be no evidence of Menge's involvement in the initiation of
the settlement at Klemzig, but it is improbable that Pastor Kavel and
Charles Flaxman, both newcomers to South Australia, would have made
the choice without special advice\,---\,and who better to give it than
Johann Menge?\fn{Letter by J.~Menge, 17 Oct.\ 1838, SRSA item\,986/9;
\citet[p.\,29]{oneill1988}; \citet[p.\,96]{schubert1985}.}

The first attempt at irrigation from the River Murray was made by
Edward John Eyre, \index{Eyre, E.\,J} who in 1841 became a settler on
1400\,ac at Moorundie, \index{Moorundie} 70~miles north-east of
Adelaide.  Previously he had worked on pastoral holdings in New South
Wales, made two overland trips with livestock from that colony to
Adelaide, and gained distinction as an explorer\,---\,notably for his
epic journey in 1841 along the coast to Albany in Western Australia.
Eyre's intention to resume life as a pastoralist had been demonstrated
in 1839 when he acquired land by the Murray; he was able to occupy the
holding after appointment by Governor Grey \index{Governor!Grey} in
September 1841 as Resident Magistrate on the Murray and Protector of
Aborigines, with a salary of \pounds300 a year.  Grey's decision
reflected concern at serious clashes earlier that year between
overlanders and Aborigines near the River Murray; it also recognised
Eyre's reputation for good relations with the natives and may have
been intended as a reward for his recent success in exploration at a
time when he had no income.

Eyre went to live on the river bank in October 1841. The nearest
European along this stream was then more than 60 miles south at
Wellington, \index{Wellington} but he was soon joined by E.\,B. Scott,
\index{Scott, E.\,B.} 
his friend and companion on some explorations, who also settled near
the river and gave important help to Eyre during residence there.
Early in 1842 Eyre set about preparing to grow wheat as well as
arranging for the sale of small allotments to provide a township.  Few
sales were completed and the projected town, to be named for Sturt the
explorer, was never realised.  However, that year found soldiers,
mounted police, and building workers numbering altogether more then 40
living at the settlement.\fn{\citet[p.\,152]{dutton1977}.}

Eyre's agricultural efforts suffered setbacks.  These derived partly
from his travels on duty as Protector of Aborigines and in the spring
of 1842 when he was employed to search for a missing settler on Eyre
Peninsula. \index{Eyre Peninsula} More significantly there was marked
variation in annual rainfall at Moorundie and a serious risk of
flooding by the river in springtime.  Reporting to Governor Grey in
January 1844, Eyre mentioned major difficulties in his farming: in
1842 rainfall was insufficient for cropping but in 1843 the wheat grew
luxuriantly after good rains only to be ruined when the river flooded
from September to November.\fn{\citet{eyre1985}.}

Towards the end of 1842, Eyre had the satisfaction of raising a good
crop of wheat from half an acre of land moistened by soakage when the
river level rose.  He proposed then to erect sluice dams across
channels normally allowing flood water entry to his river flats, so
that this flow could be regulated to provide for irrigation and for
retention of a supply behind the banks, yet guarding against serious
flooding of his land.\fn{\citet[p.\,14]{hawker1901}.}  The
necessary work was to be undertaken in 1843.  Eyre's 1844 report to
Grey also stated that:
\begin{quote}
	From previous experience it had been apparent that the river
	rose periodically several feet and usually overflowed many of
	the alluvial flats lying between the river bank and the fossil
	cliffs\,---\,considerable labor and expense were bestowed in
	damming up the entrances by which the water escaped from the
	river to lower levels and in digging canals for the purpose of
	irrigation\,---\,these were completely successful as long as the
	river did not attain a greater height than it had risen to in
	1841 and 1842 but upon its rising several feet beyond this
	level there were of course no impediments to its
	progress\,---\,the dam and ditches were all destroyed and the
	whole expanse of alluvial flats were again laid under water,
	in some places fully six feet deep and of course all
	cultivation was completely annihilated.
\end{quote}

This reverse did not deter Eyre; he advised that:
\begin{quote}
	the ensuing year will I trust see these difficulties fully
        conquered and good embankments thrown up at all the openings
        through the river bank\,---\,so as to effectively block out the
        highest flood\,---\,at the same time that a few shallow ditches
        cut across these lands intended for cultivation will afford
        the important and in Australia almost unknown power of
        completely irrigating at pleasure all such lands. Thus what
        has been inconsiderably deemed as an insuperable objection to
        the Valley of the Murray will I believe eventually prove its
        highest recommend\-ation.\fn{\citet[pp.\,67--9]{eyre1985}.}
\end{quote}

Apparently the embankments were renewed after 1843 but the river rose
high again in 1844 and Eyre expected that much of his crop would be
lost and the balance cut for hay.  This was Eyre's last year at
Moorundie\,---\,he resigned his official post and left for England in
December.  Further flooding in later years brought more damage to his
dams; in 1856 the Moorundie settlement was abandoned in favour of a
higher site, several kilometres upstream, where the township of
Blanchetown \index{Blanchetown} developed.\fn{Moorundie and
E.\,J.~Eyre, SRSA item 951/9, Letter from Eyre to E.\,B.~Scott, 1
Nov.\ 1844}

Eyre left no details of results of his irrigation.  Whether or not he
had real success with it, there is no doubt that he made every effort
to pioneer its use on this river.

There is no doubt about the practice of irrigation near Adelaide in
1843, using the discharge of the River Torrens \index{river!Torrens}
onto lowland near the coast, where occasional floods affected an area
known as the Reed Beds. \index{Reed Beds} By 1843 wheat-growing
\index{wheat} was
well established in South Australia.  Its cultivation then involved
more than 23\,000\,ac, mainly on the Adelaide \index{Adelaide} plains,
and exports of grain had begun.  The colony easily outstripped cereal
production in the area later known as Victoria, which then had less
than 5000\,ac under wheat and depended on imports of grain.  The
spring season in Adelaide during 1843 was relatively dry, with only
2.14\,in.\ in Sept.--Oct.\ compared with a long-term mean of 3.72\,in.
One unnamed farmer at the Reed Beds experimented with irrigation and
as a result the crop yield from the irrigated portion was one-third
better than from the rest.  This locality was subdivided soon after
settlement and by 1843 had many landholders, some with scores of
hectares devoted to wheatgrowing.\fn{\citet[p.\,104]{dutton1892};
Commonwealth Yearbook 1972; \textsl{Southern Australian}, 26 Dec.\
1843.}

Tasmanian irrigation also got attention in South Australia, when the
\textsl{Southern Australian} (29 Dec.\ 1843) included a lengthy report of
the lecture on irrigation given in Hobart during 1843 by Major Hugh
Cotton \index{Cotton, H.} who considered its use in India and
Tasmania.  The decision to reprint this material from the Hobart press
was made in view of topical interest following news of irrigation at
the Reed Beds.

Irrigation at the Reed Beds was repeated in 1844 when it was reported
that
\begin{quote}
	The beneficial effects of irrigation have never been more
	clearly demonstrated than by this season's crops at the Reed
	Beds.  Of fruit the produce is abundant, but the wheat crops,
	generally good, are on some sections extraordinary. At
	Wymondlybury (Dr Addison's farm) the wheat crop is unusually
	heavy, that of the `White Talavera' standing about 6 feet in
	height, the ears exceedingly well filled, and several we have
	seen (plucked indiscriminately) measure seven inches in
	length.  The produce is estimated by competent judges at about
	45 bushels per acre.\fn{\textsl{Southern Australian}, 31 Dec.\
	1844.}
\end{quote}

This farm, the property of Dr Joseph Addison, \index{Addison, J.}
covered 133\,ac on the north side of the River Torrens, the land lies
in the modern suburb of Fulham. \index{Fulham} Addison then had 42\,ac
under wheat. He was one of those medical men who gave up their
profession on migration to Australia.\fn{J.\,B.~Cleland, \textsl{Med.\
J.\,Aust.}, 29 Oct.\ 1938, p.\,732.}

Rainfall amounting in places to more than 40\,in.\ annually feeds
streams on both sides of the Mount Lofty range, and it was not long
before irrigation was taken up on its eastern side.  This use also
involved another who gave up medicine in favour of farming. John
Rankine, \index{Rankine, J.} regarded as the founder of Strathalbyn,
\index{Strathalbyn} 35 miles south-east of Adelaide, obtained land in
1839 on the Angas River \index{river!Angas} after arriving in Adelaide
with two brothers and their families together with other Scottish
migrants.  They settled in the Strathalbyn district in 1841 where John
Rankine's property, at the junction of the Angas River and Gould
Creek, was known as Blackwood.  \index{Blackwood} Before 1846,
Rankine, according to Robert Davenport, \index{Davenport, R.} diverted
water from the river `to irrigate his fertile garden and potato
crop'.\fn{\citet{stowe1973};
\citet[p.\,159]{dutton1892}.}

During the 1840s vineyards \index{vineyards} were established near
Adelaide and the first winery was built in 1845.  In 1850--52,
Alexander Boord \index{Boord, A.} established and irrigated a vineyard
at Freshford\,---\,near Athelstone \index{Freshford}
\index{Athelstone} seven miles from
Adelaide\,---\,using water from the River
Torrens.\fn{\citet{fornachon1958};
\citet[p.\,22]{bishop1977}.}

\closure
Irrigation failed to secure a firm place on the mainland where
resources\,---\,human and physical\,---\,were concentrated on
producing wool for export and essential foodstuffs: cereals and meat.
Only a few well-established men undertook irrigation and then only on
a limited scale; there is no indication that the practice became
firmly established anywhere as it had done in Tasmania.  Most
irrigation was recorded from the southern part of the continent,
particularly in South Australia.  However, greater acquaintance with
parts of the mainland previously unknown to Europeans brought an
appreciation of the value of irrigation in the future.

The discoveries of gold in New South Wales in 1851 led to great
dislocation of pastoral and agricultural production for a few years,
thus impeding further development of irrigation in that colony.

%\section*{References}
%1. B.Fitzpatrick, The British Empire In Australia, 1941, p.137.
%2. N.Bartley, Opals And Agates, 1892, p.62.
%3. S.H.Roberts, The Squatting Age In Australia 1835-1847,1964, p.205.
%4. F.P.Labilliere, Early History Of The Colony Of Victoria, 1878, vol.II,
%    p.331.
%5. R.V.Billis \& A.S.Kenyon, Pastures New, 1974, p.166.
%6. W.Howitt, Land, Labour And Gold, 1855/1972, p.146.
%7. W.Howitt, 1972, p.147. 
%8. W.Howitt, 1972, p.247.
%9. T.L.Mitchell, Three Expeditions Into The Interior Of Eastern Australia,
%    1839, vol.II.
%10. T.L.Mitchell, 1839, vol II, p.332.
%11. T.L.Mitchell, Journal Of An Expedition Into The Interior Of Tropical 
%      Australia, 1848, pp. 421-24.	
%12. P.Cunningham, Hints For Australian Migrants, 1841, p.4.
%13. Sydney Gazette, 2/9/1841.
%14. M.Kaluski, Sir Paul E. Strzelecki, 1985.
%15. P.E.de Strzelecki, Physical Description Of New South Wales. . ., 1845,
%      p.446-47.          		
%16. L.Paszkowski, Sir Paul Edmund de Strzelecki, 1997, p.263. 
%17. C.J.Lloyd, Either Drought Or Plenty, 1988, p.164.
%18. SA Government Gazette, 28/10/1847, \&  Hobart Town Gazette,
%      28/9/1847, cited by Margaret Mason-Cox, Lifeblood Of A   
%      Colony,1993,p.114.
%19. HRA I, vol XXV, pp. 457-79.
%20. R.V.Billis \& A.S.Kenyon, Pastoral Pioneers Of Port Phillip, 
%      1974,p.168.
%21. VicRept.Select Comm. Sewerage and Supply of Water, 1853.
%22. J.Morphett, South Australia, 1836/1962, p.12.
%23. South Australian Register, 30/1/1878.
%24. D.Schubert, Kavel's People, 1985, p.96.
%25. D.Schubert, 1985, p.113, 131.
%26. D.Schubert, 1985, p.97.
%27. Southern Australian, 1/5/1839, cited by Schubert 1985, p.88.
%28. Letter by J.Menge, 17/10/1838, SRSA. item 986/9.
%29. B.O'Neill, The German Experience Of Australia, 1988, p.29.
%30. D.Schubert, 1985, p.96.
%31. G.Dutton, Edward John Eyre, 1977, p.152.
%32. E.J.Eyre, Reports and letters to Governor Grey from E.J.Eyre at
%     Moorunde, 1985.
%33. J.C.Hawker, Early Experiences In South Australia, 1899/1975, 2nd
%      Series, p.14.
%34. E.J.Eyre, Reports And Letters. . .1985, pp.67-69.
%35. Moorundie and E.J.Eyre, SRSA item 951/9, Letter from Eyre
%       to E.B.Scott, 1/11/1844
%36. F.S.Dutton, South Australia And Its Mines, 1846, p.104, \& 
%      Commonwealth Yearbook 1972..
%37. Southern Australian, 26/12/1843.
%38. Southern Australian, 31/12/1844. 
%39. J.B.Cleland, Med.J.Aust. 29/10/1938, p.732. 
%40. H.J.Stowe, They Built Strathalbyn, 1973.
%41. F.S.Dutton, 1846, p.159.
%42. J.Fornachon, The Wine Industry Of South Australia, in R.J.Best(ed),
%      Introducing South Australia, 1958.		   
%43. G.C.Bishop, The Vineyards Of Adelaide, 1977, p.22.
