% $Id$
% CHAPTER EIGHT
% 4902 words at 1/5/99

\setcounter{endnote}{0}

\chapter{Experience of Irrigation Trusts}
\label{ch:trusts}\addtoendnotes{\protect\section*{Chapter \thechapter}}
\markboth{
\textsc{Chapter \thechapter. Irrigation Trusts}
}{
}

Irrigation trusts were established under legislation in different
colonies to manage irrigation systems in specified areas. Most of
these trusts were formed after 1885 in Victoria, where they were
regarded by government as important for its policy of extending the
use of irrigation by a major delegation of responsibility to a form of
local government.  A few trusts were established in New South Wales,
each in response to special legislation.  Two trusts were formed in
response to the financial collapse of the promoters of existing
irrigation settlements.

The trusts were important for development of integrated irrigation
schemes during at least twenty years but only two remained in 1920.
This chapter considers their experience and how well they helped to
extend the use of irrigation.

Incorporated trusts with responsibility for water supply in specified
localities had existed in Australia before legislation provided
specifically for irrigation trusts.  In Tasmania the River Clyde Act
of 1857 led to the appointment of trustees to provide a water supply
to the townships of Bothwell and Hamilton on that river.  The
legislation was altered in 1869 to empower the trust to levy rates and
to widen its responsibilities which then came to include control of
irrigation with Clyde River water.  The numerous waterworks trusts of
northern Victoria, referred to in the previous chapter, depended on
legislation in 1881.\fn{\citet[p.\,136]{masoncox1994}.}

\section*{Victorian irrigation trusts}

Provision for trusts to undertake irrigation was made by legislation
of the Victorian government in 1883, but with no immediate effect.
Applications to form them were made after amending legislation, which
liberalised the conditions and provided access to funds, and the
indications that the government was about to offer financial advances
to the trusts and to provide those headworks known as national works,
which would supply two or more trusts.  By the time of the major
enabling legislation, known as the Irrigation Act 1886, several
irrigation trusts had been formed---all in the lower part of the
Loddon River valley or adjacent channel country of the Murray valley.
More trusts were formed under that legislation in the next few years
but only two after 1890.  One more trust---the First Mildura
Irrigation Trust---was constituted later under different legislation
referred to in another chapter.\fn{VicWater Conservation Act no.\,859
of 18 Dec.\ 1885; VicPP no.\,53 of 1885, RC~Water Supply, Further
Progress Rept July 1885.}

When the Minister for Water Supply, Alfred Deakin, opened
parliamentary debate in June 1886 on comprehensive measures for
extending the use of irrigation, his estimate of the area available
for irrigation was approximately 3.2~million\,ac with about
1~million\,ac as the total to be watered each year.  A great expansion
of irrigation was expected in northern Victoria, where the total
irrigable area in nine districts was given as almost 3~million\,ac,
with more than two-thirds lying in the Goulburn, Loddon, Upper Murray,
and Western Wimmera districts.  The limited data then available on
water resources indicated that for all nine districts no more than
750\,000\,ac could be irrigated in winter or 370\,000\,ac in summer,
with more than a third dependent on the Goulburn River.\fn{VicPD
vol.\,51, 1886, pp.\,415--447;
\textsl{Leader}, 3 July 1886.}

The first irrigation trust to be formed in Victoria was the Leaghur
and Meering Trust in 1885.  It embraced a comparatively small area of
10\,300\,ac in the western part of the lower Loddon valley, where
landholders had previously been operating a cooperative system of
irrigating wheat crops.  More significant was the formation in 1886 of
the Tragowel Plains Trust for a large area east of the Loddon river.
The procedure for establishing an irrigation trust involved an
application to the water supply department by a majority of
landholders in an agreed area, together with plans and estimates of
expenditure.  If the formation of the trust was approved by the
government, which might seek independent advice, responsibility for
the necessary works was taken by elected members of the trust, who
were known as commissioners, assisted by the trust engineer.  The
trust was also charged with collection of revenue and it controlled
expenditure of loan money, for which no interest was to be charged by
government for the first five years.  As a result of amendments to the
original legislation, the distinction between waterworks trusts
serving country areas and the irrigation trusts became blurred, with
the irrigation trusts supplying stock and domestic supplies in
addition to irrigation and the waterworks trusts were permitted to
undertake irrigation.\fn{VicPP no.\,20 of 1896, RC~Water Supply,
Rept.}

Before the end of 1886 five irrigation trusts with a total area of
130\,000\,ac (52\,631\,ha) were established under the new Act in the
settled parts of northwestern Victoria: in the Loddon Valley and the
nearby area traversed by the Murray and its effluents.  In 1887 the
Swan Hill Irrigation Trust on the Murray River was the only one
formed, but in 1888 another five were created, mainly in the Loddon
Valley.  The peak year was 1889 when eight more were formed, including
the Rodney Irrigation Trust, the first in the Goulburn Valley, three
dependent on the Campaspe River and one in southern Victoria at
Bacchus Marsh.  Several more trusts, including one at Bairnsdale in
Gippsland, were formed before the last in 1893.  Altogether 30 trusts
were constituted.\fn{VicPP no.\,20 of 1896, RC~Water Supply, Rept.}

\subsection*{National headworks}

The first of the national headworks was to be a weir on the Goulburn
River to divert supplies for irrigation in the valley; it was
authorised in 1886 by the River Goulburn Weir Act with an
appropriation of \pounds20\,000.  Such a weir had been advocated in
the Grand Northwestern Canal scheme of the 1870s, and by the Water
Conservancy Board in its proposals for water supply and irrigation
between the Goulburn and Campaspe Rivers.  The initiative in securing
the weir had been taken earlier by the Echuca and Waranga Waterworks
Trust which wished to use Goulburn River water for irrigation in
addition to stock and domestic supply.  Delays in settling on a design
for the weir led the trust to install a pumping plant on the river in
1885 to supply its channels, only to experience serious losses by
seepage, evaporation and channel breakages.  An earthen embankment
across the nearby Waranga Swamp had been constructed to allow storage
of water there and to give the necessary elevation required for the
channel gradient; leakage from this section of the channel involved
serious expense to the trust and reduced the area to be
supplied.\fn{River Goulburn Weir Act, VicStatute no.\,598 of 16 Dec.\
1886; \citet[pp.\,35--50]{martin1955}.}

The weir to be constructed by the government on the Goulburn River was
expected to divert water to both sides of the valley.  Work started in
1887 with expectation of completion in 1889 but floods delayed its
opening until March 1890 and its first supply of water occurred next
year.  On its completion, which provided a storage capacity of
20\,000\,acre-foot, a new channel on the west side of the river was
made from the weir to connect with those from the pumping station
downstream and then parallel to the trust's Cussen channel except for
a deviation around the north end of Waranga Swamp.  This national
channel later became known as the Stuart Murray channel.  No use of
the weir to provide supplies for the eastern side of the valley was
made for many years.\fn{\citet{murray1892}; \citet{mead1914};
\citet{mccoy1988}.}

Conversion of Waranga Swamp into a reservoir was designed in 1892 but
work was delayed by the depression of the 90s and did not start until
the drought year of 1902.  The reservoir first received water from the
Goulburn River in October 1905, when its capacity was
197\,000\,acre-foot.  The lack of adequate storage for Goulburn River
water until 1905 seriously curtailed the extension of irrigation
dependent on the river.\fn{\citet[pp.\,106--8]{bossence1965};
\citet{mead1914}.}

Other national works commenced during Deakin's term as Minister for
Water Supply were intended to improve supplies in parts of the Loddon
Valley and the adjacent Gunbower district where wheatgrowers were
already making the greatest Victorian use of irrigation.  The
formation of trusts in these districts represented an enthusiastic
response to the new legislation and emphasised the need for an
adequate supply of irrigation water for landholders in parts of
Victoria with low annual rainfall.  One of the national works was the
Laanecoorie weir and dam in the hills on the upper reaches of the
Loddon River.  Work began in January 1889 and finished by 1892 when it
became useful.  Its purpose was `to regulate the river by storing
water during flood so as to maintain the flow during dry seasons'.  It
had an initial storage capacity of 14\,042\,acre-foot, less than the
Goulburn Weir storage and reduced later to 6650\,acre-foot by
siltation.  Its purpose was to serve irrigation trusts of the
mid-Loddon valley, including those near Boort and on the Tragowel
plains.\fn{VicPP 1884, Statistical Register, Production: Irrigation to
31 Mar.\ 1884; \citet[p.\,14]{murray1892}; \citet[p.\,17]{east1940}.}

Supplies to areas further downstream were to be augmented by diversion
of water from the Murray, taking advantage of natural flooding into
its Gunbower Creek anabranch, and by improving flow to the large Kow
Swamp, which would be made into a reservoir for a canal to deliver
supplies to the Loddon River near Kerang.  This westward trend of
overflow from the Murray River via Gunbower Creek had been used in the
past for supplies to landholders, as indicated by Gordon and Black in
their 1882 proposals for water supply works in the Gunbower district.
The national works in the district began in 1889 with improvements to
the outlet from the Murray to Gunbower Creek, and construction of a
channel several miles long from the outlet to Gunbower in order to
improve the flow westward to Kow Swamp.  Subsequently there was
embankment of this swamp to increase its capacity by 1894 to
40\,000\,acre-foot and construction of the Macorna Channel conveying
water from this storage basin across the Tragowel Plains to the Loddon
River.  Stuart Murray expected beneficial effects of some of these
modifications to begin in the summer of
1892--93.\fn{\citet[p.\,14]{murray1892};
\citet[pp.\,21--27]{mccoy1988}; VicPP no.\,35 of 1902--03, Interstate
RC~on the River Murray. Rept and MoE, evidence of A.\,S.~Kenyon.}

\subsection*{Irrigated agriculture}

The parts of northern Victoria intended for irrigation by the trusts
were used mainly for cereal growing and grazing, both of which would
benefit from water supply in times of low rainfall.  However, as
rainfall in these parts fluctuates from year to year, irrigation would
be no advantage to some landholders in wet years.  The greatest use of
irrigation in the years 1891 to 1895 was in 1895 when rainfall in
Victoria was generally well below average following relatively wet
years which must have provided the stored water needed in 1895.  In
the wetter years the demand for irrigation water declined and in a
sequence of dry years there was insufficient water to meet the demand.
In these circumstances the trusts had difficulty in obtaining income
to discharge debts to the government.\fn{VicPP no.\,20 of 1896,
RC~Water Supply; Comm.\ Bur.\ Meteorol., Results of Rainfall
Observations made in Victoria, 1937; J.\,C.~Foley, Comm.\ Bur.\
Meteorol.\ Bull no.\,43, 1957, Drought in Australia.}

While the areas of irrigable land and the available water resources
mentioned for the nine northern districts in the government proposals
of 1886 may indicate official hopes at the time, the proposals
represented by the various trusts showed a marked preference for
irrigation in the drier areas west of the Goulburn River together with
a minor response in southern Victoria---at Bacchus Marsh, Werribee,
and Bairnsdale.  The lack of interest then in irrigation to the east
of the Goulburn River is indicated by the failure of efforts by the
East Goulburn Irrigation League to form a trust despite active
encouragement in 1890 by the Water Supply
department.\fn{\citet[p.\,70]{martin1955}.}

The problems of the irrigation trusts were referred to by Deakin in
addressing the irrigators' conference of 1890, before completion of
any national works,
\begin{quote}
	There are at present constituted, or practically constituted,
	25 Irrigation Trusts in Victoria.  Of these 25, one fifth (or
	five Trusts) had some water to sell to their constituents last
	season, and only five.  All of those five drew their supplies
	from uncontrolled streams which, like the majority of our
	Australian streams, are not to be relied upon to contribute
	any considerable flow of water. Not one single Trust in
	Victoria could undertake to sell water to its constituents
	beforehand, because in not one case as yet have the necessary
	works been brought to such a condition as would enable them to
	be certain of being able to deliver what they had sold. Only
	five Trusts could sell any water, and they had only an
	uncertain quantity, and, for all they knew, no quantity at all
	to dispose of.  Of those five Trusts, four are among the
	smallest Trusts in the colony.\fn{A.~Deakin, Speech in Procs.\
	First Conf.\ Irrigationists Victoria, 1890, Govt Printer,
	Melb.}
\end{quote}

\subsection*{Five irrigation trusts}

Water to most of the irrigation trusts in Victoria was provided by
five rivers: the Murray, Goulburn, Campaspe, Loddon, and Wimmera.  The
Murray supplied six irrigation trusts, either directly by pumping or
by gravitation.  The Swan Hill Irrigation Trust was constituted in
1887 with an area of 15\,000\,ac , much of it previously under the
control of a waterworks trust providing stock and domestic supplies.
Before 1896 no more than 2669\,ac were irrigated, mainly for cereal
crops, by diversion of water when the river was at high level. There
were efforts to foster production from fruit trees and vines on
holdings of 10\,ac created by subdivision but in 1895 less than 20\,ac
were irrigated for horticulture by pumping from the Murray
River.\fn{VicPP no.\,20 of 1896, RC~Water Supply, Rept.}

The Goulburn River, the Victorian tributary of the Murray with annual
flow ten times greater than others, was intended by the Government to
supply by far the largest irrigated area in the colony.  Irrigation
had been sought in 1882 by two local governments with areas west of
the river; it began by using channels of a waterworks trust supplied
at first by pumping from the river but irrigators lacked supplies in
some years.  In 1889 the Rodney Irrigation Trust was formed by
excision of territory from the older waterworks trust and irrigation
of small blocks with fruit trees and vines was carried on notably at
Ardmona.  Even when more dependable supplies of water were provided
after completion of the Goulburn Weir, irrigation was used
consistently for horticultural and fodder crops, with little
involvement of cereals and pastures except when rainfall was
deficient, as in 1895.  The Rodney Trust incurred heavy debts because
of general failure by landholders to contribute to revenue while
supplies of water were not assured.\fn{VicPP no.\,20 of 1896, RC~Water
Supply, Rept Appendix~c, p.\,205.}

The major undertaking for irrigation using water from the Campaspe
River involved the Campaspe Irrigation Trust, established in 1889 with
an area of 18\,797\,ac some of which would be irrigated from a weir on
the river.  But although the works cost \pounds52\,959 the sale of
water in 1895 realised less than \pounds3. This failure was linked
with diversions involved in the Coliban scheme from headwaters of the
river to the gold-mining towns of Bendigo and Castlemaine.

Further west, in the Loddon Valley, several trusts were constituted,
one of the most extensive being the Tragowel Plains Irrigation Trust.
Early in 1884 an irrigation scheme for the district was put forward by
residents, leading to a deputation in March to the Minister for Water
Supply involving a local parliamentarian, Rev.~E.\,C.~De Garis and
another local resident, W.\,W.~Culcheth, an irrigation engineer and
member of the royal commission on water supply, was instructed to
report on the scheme and responded with statements in August 1884, and
March and April of 1885.  He found that almost all the area under
consideration for the proposed trust was irrigable.  In March 1886 the
Trust was approved and 15 Commissioners were appointed for three years
with De Garis as chairman.  Irrigation in the trust area of
192\,800\,ac was undertaken principally for cereals and pastures; in
1891 the area so treated was the largest of all the trusts in Victoria
and it remained so in the subsequent four years, reaching 25\,403\,ac
in 1895.\fn{\textsl{Leader}, 3 July 1886.}

The Wimmera River was the only stream of any consequence for
irrigation within the area of the West Wimmera Irrigation and Water
Supply Trust formed in 1888 to embrace 1\,643\,132\,ac. This trust
took over water supply channels from a Waterworks Trust as well as two
important weirs and the Wartook Reservoir in the Grampians mountains;
it provided stock and domestic supplies in an extensive wheatgrowing
district and water for irrigated areas near Horsham.

\subsection*{Trusts investigated}

The opening of the Goulburn Weir in 1890 marked the high point of
Deakin's association with irrigation in Victoria, for he ceased to be
Minister when his government was defeated later in the year.  The next
Minister, George Graham, was aware of troubles associated with water
supply in country areas but failed to remedy them before his
replacement during the financial debacle of 1893 by J.\,H.~McColl, son
of the notable Hugh McColl.  This Minister of Water Supply instituted
a Waterworks Enquiry Board in 1893 to report on the financial problems
of waterworks trusts.  Next year a new government appointed a Royal
Commission on Water Supply to enquire into `the financial position and
prospects of the various local bodies that have obtained loans from
the state for the construction of works of water supply'. The six men
appointed to the commission were all legislators and they gave their
only report in 1896 together with a long record of evidence.
Considerable attention was given to the irrigation trusts and their
administration by the Water Supply Department.  Many of the 410
witnesses examined were irrigators.\fn{\citet[p.\,45]{martin1955};
VicPP no.\,20 of 1896, RC~Water Supply, Rept; VicPP no.\,21 of 1896,
RC~Water Supply, MoE.}

The Royal Commission found that the cost of irrigation then exceeded
\pounds1.8\,million, made up mainly of loans to 30 trusts and the cost of
national headworks, together with liabilities transferred from
waterworks trusts and arrears of interest due to the State.  Trusts
had been constituted without attention to justification for proposed
expenditure or to adequate supply of water.  One of the many trusts in
the Loddon Valley had used a loan of \pounds165\,000 to construct
channels capable of watering 200\,000\,ac but the river could not
supply water for even one-fifth of that area.\fn{VicPP no.\,20 of
1896, RC~Water Supply, Rept, p.\,185.}

Expenditure of money lent by the government to the trusts was found to
have been beyond the control of the Water Supply Department and there
was evidence of poor internal management of funds by trusts.  Advances
to them were made upon recommendations of Alfred Deakin, then Minister
for Water Supply.  The most glaring examples of uncontrolled advances
to trusts occurred with the private irrigation trusts, which initially
were not entitled to borrow money from the State.  The arrangements
concerning one such trust, the Werribee Irrigation Trust, proved to be
a source of embarrassment later to Deakin when he attempted to
represent his departmental head as responsible for irregularity but
was finally obliged to admit his own fault.  Decisions to form trusts
in some cases involved persons who were not bona-fide owners of land
in the area involved.  This criticism applied particularly to private
irrigation trusts, which each involved one landholder---either an
individual or a syndicate.\fn{VicPD 18 Dec.\ 1896, p.\,4738.}

Irregularities concerning formation of trusts were found by the
Commission to include pressure from the government water supply
department to secure their formation.  Residents in different parts of
the Loddon Valley had complained of being forced to form trusts in
response to threats from the department that otherwise their existing
supply of irrigation water would be stopped.  In the relevant case of
the Leaghur and Meering Trust, its area had included cooperative
irrigation of wheat by means of works constructed for the purpose
prior to the formation of the trust in 1885.

There had been unfortunate clashes of interest between irrigation
trusts dependent on one source of water and also those between
irrigation trusts and water supply trusts constituted earlier.  In the
latter case there was the example of friction between the Loddon
United Waterworks Trust (1882) and the Tragowel Plains Irrigation
Trust whose area was transferred from the Loddon United Waterworks
Trust.\fn{\citet{sharland1971}.}


Inquiries about the character of soil and its suitability for
irrigation in trust areas were generally lacking.  At Swan Hill, the
trust area included marshy ground; its liability to flooding from the
Murray led to the erection of a levee bank at additional expense.  In
the Loddon Valley unsuitable ground classed as irrigable had been
alluded to by Stuart Murray:
\begin{quote}
	rough crab-holey nature of much of its surface---a feature
	that greatly detracts from its adaptability for irrigation.
	Much of the land that has been irrigated is so uneven that the
	crops are patchy, being over-watered in some spots and
	insufficiently watered in others.  These unevennesses can, and
	it may be presumed will, be reduced by surface levelling; but
	this will cost money and require time.\fn{\citet[p.\,13]{murray1892}.}
\end{quote}

Two aspects of the construction of works by the trusts attracted
criticism in the Commission's report.  Advisers to the trusts included
engineers attracted to Victoria by the prospect of employment but
without good experience of Australian conditions.  Some had been
appointed whose competence was in question and it emerged that the
Water Supply Department had not always made sufficient inquiries about
the qualifications of advisers.  There was also the serious
discrepancy between the estimated and actual cost of many works, with
some projects costing several times the estimated figure.  As well
there were the cases where irrigation channels had been provided far
in excess of the actual scope for irrigation.

Stuart Murray, head of the Water Supply Department, told the
Commission of his reservations about many procedures involving the
irrigation trusts, which occurred despite his advice.  He claimed that
the problems of the trusts were affected by political pressure.  His
evidence touched on the promotion of irrigation trusts in the
interests of land speculators, as part of the Commission's report
shows:
\begin{quote}
	Under the present law preliminary expenses in connexion with
	the formation of Irrigation Trusts are allowed to be paid out
	of loan.  This practice we think should be abandoned.  If
	Trust promoters had had in the past to pay preliminary
	expenses out of their own pockets many of the Trusts which are
	at present in existence would, we are convinced, not have been
	constituted.  The evidence of the Chief Engineer of Water
	Supply is to the effect that a great many of the irrigation
	schemes have been initiated in the interests, in the first
	place, of syndicates and dealers in land.  If this contention
	be correct, and from our inquiries we are inclined to believe
	that it is, it can be readily understood what an incentive the
	fact of being able to pay preliminary expenses out of State
	money was to speculators.\fn{VicPP no.\,20 of 1896,
	RC~Water Supply, Rept, p.\,191.}
\end{quote}

\subsection*{Termination of Victorian Irrigation Trusts}

Despite the findings of the 1894 Royal Commission, irrigation trusts
continued to operate and were given some relief from their burden of
debts.  Unfortunately these continued to mount though some progress
was made in extending irrigation.  In the dry years following 1895 the
deficiencies of Victorian water supply for irrigation and for stock
and domestic use became more evident.  Agricultural settlement of the
northwestern Mallee district had begun in the 1880s, requiring
considerable diversion of water for stock and domestic supply from the
Wimmera river system that might otherwise have been used for local
irrigation schemes.  In the Loddon valley there was insufficient water
to satisfy the competing demands.  As the drought reached its worst in
1902--03, the government faced a time of reckoning.  George Swinburne,
an engineer turned company director, was elected to parliament in
1902.  He accompanied the Premier, William Irvine, on a tour of
northern districts in 1903, became Minister for Water Supply in 1904
and quickly submitted his Water Bill aimed at major reforms.  The
Water Act of 1905 provided for a Commission independent of control and
with three members to take charge of all assets and liabilities of the
various rural trusts except that managing irrigation at Mildura.  At
that time storage capacity for country water supplies was
approximately 175\,000\,acre-foot, with 75\,000\,ac under
irrigation.\fn{\citet[p.\,34]{boorman1942};
\citet[p.\,90]{martin1955}.}

\section*{Trusts in New South Wales and South Australia}

Irrigation trusts were established in New South Wales from 1890.  The
Wentworth municipal government was constituted as an irrigation trust
by the Wentworth Irrigation Act of 1890, with control of more than
10\,000\,ac of low-level alluvial country east of Wentworth to be
irrigated by pumping from the Murray River.  The trust accepted the
recommendation of H.\,G.~McKinney that initially development should be
confined to 1500\,ac near Wentworth, with water to be lifted 25\,ft
from the river by pumping.  The financial collapse in 1893 made
progress impossible and members of the trust sought its dissolution.
This was agreed to by the government which by 1896 had called tenders
for the pumping plant and provision of distribution channels. The
scheme attracted few settlers at first---there were only about 400\,ac
leased by 1906 when the government undertook a publicity campaign to
attract settlers---with some success.  The Hay Irrigation Trust, also
under municipal control, was established in 1892 with an area of
17\,147\,ac but the area was reduced to 4000\,ac in 1893 and only
600\,ac was under irrigation in 1902, using water drawn from the
Murrumbidgee River.  A similar scheme for Balranald involving 2000\,ac
was started in 1893 but only 60\,ac were leased by
1906.\fn{H.\,G.~McKinney, The Wentworth Irrigation Scheme, Agr.\ Gaz.\
NSW 1896, \textbf{7}, pp.\,469--70; \citet[p.\,113]{jeffcoat1988};
\citet{williamson1968}.}


The only other irrigation trust in New South Wales before 1920 was the
shortlived Murrumbidgee Irrigation Trust set up under legislation in
1910 to control development of irrigation north of the Murrumbidgee
River using water gravitating from the Burrinjuk Dam then under
construction to supply 250\,000\,ac.  This trust had three members:
the government ministers for public works, lands, and agriculture; its
secretary and executive officer was the chief irrigation engineer.
The enabling legislation was repealed in 1912 with passage of the
Irrigation Act which gave control of developments in the Murrumbidgee
Irrigation Area (MIA) to the Commissioner for Water Conservation and
Irrigation, Mr L.\,A.\,B.~Wade.\fn{\citet{lloyd1990}.}

Although the Irrigation Act of 1912 made provision for establishment
of trusts to administer water supply districts and irrigation, there
was no use of it before 1920 to create irrigation trusts.

In South Australia the only irrigation trust was that established for
the Renmark irrigation scheme under the authority of the Renmark
Irrigation Trusts Act 1892.  The Renmark Irrigation Trust No.~1 was
controlled by a board of seven irrigators elected by those owning 10
or more acres, its inaugural meeting was in January 1894 after the
collapse of the Chaffey Brothers
companies.\fn{\citet[p.\,149]{wells1986}.}

\section*{Conclusion}

Formation of many irrigation trusts in Victoria led to provision of
various works necessary for extension of irrigation.  Five to ten
years later, the annual mean irrigated area under their control
reached more than 50\,000\,ac for the period 1891--95.  This mainly
concerned supplementary irrigation in northern Victoria of cereal and
fodder crops and pastures, all of which had been grown there widely
and successfully before the extension of irrigation except in the
occasional years of low rainfall.  Considerable debts incurred by the
irrigation trusts led to serious criticism by a royal commission of
the government administration involved.

Three irrigation trusts formed in drier parts of New South Wales led
to limited use of irrigation for horticultural crops near Wentworth
and for pasture or fodder crops near Hay and Balranald.  These trusts
were not involved with major financial problems as in Victoria.

Two trusts were created to provide satisfactory management of the
irrigation settlements, at Mildura in Victoria and Renmark in South
Australia, after the financial collapse of the promoting company.
Both trusts continue to function satisfactorily.

%\section*{References}
%1. Margaret Mason-Cox, Lifeblood Of A Colony, A History of Irrigation
%    In Tasmania, 1994, p, 136.
%2. VicWater Conservation Act No.859 of 18/12/1885, \& VicPP No.53
%    of 1885, R.C.Water Supply, Further Progress Rept. July 1885.
%3. VicPD vol.51, 1886, pp. 415-447.
%4. Leader, 3 July 1886.
%5. VicPP No.20 of 1896, R.C.Water Supply, Rept.
%6. River Goulburn Weir Act, VicStatute No.598 of 16/12/1886 \& 
%    C.S.Martin, Irrigation And Closer Settlement In The Shepparton District
%    1836-1906, 1955, p.35.
%7. C.S.Martin, 1955, pp49-50.
%8. S.Murray, Irrigation In Victoria: Its Position And Prospects, 1892,
%     Govt.Printer, Melbourne.
%9. E.Mead, Irrigation In Victoria, pp 255-268 in BAAS Handbook to
%    Victoria, 1914, \& C.G.McCoy, The Supply Of Water For Irrigation In
%    Victoria From 1881 To 1981, 1988.
%10. W.H.Bossence, Murchison, 1965,pp 106-108, \& E.Mead, 1914.
%11. VicPP 1884, Statistical Register, Production: irrigation to 31/3/1884.
%12. S.Murray, 1892, p.14.
%13. L.R.East, Erosion And Water Supply, in Soil Erosion In Victoria, 1940,
%      p.17.
%14. S.Murray, 1892, p.14, \& C.G.McCoy, 1988, pp. 21-27. 
%15. S.Murray 1892, \& VicPP No.35 of 1902-03, Interstate R.C. on the River
%      Murray. Rept and MoE, evidence of A.S.Kenyon.
%16. S.Murray, 1892, p.14.
%17. VicPP No.20 of 1896, R.C.Water Supply, \& Comm.Bur.Meteorol.
%      Results Of Rainfall Observations Made In Victoria, 1937, \& J.C.Foley,
%     Comm.Bur.Meteorol.Bull No.43,1957, Drought in Australia. 
%18, C.S.Martin, 1955, p.70.
%19. A.Deakin, Speech in Procs First Conf.Irrigationists Victoria,1890, 
%      Govt.Printer, Melb.
%20. VicPP No.20 of 1896, R.C.Water Supply, Rept.
%21. VicPP No.20 of 1896, R.C.Water Supply, Rept. Appendix c, p.205.
%22. Leader, 3/7/1886.
%23. C.S.Martin, 1955, p.45.
%24. VicPP No.20 of 1896, R.C.Water Supply. Rept.
%25. VicPP No 20 of 1896, R.C.Water Supply,Rept..\& VicPP No.21 of
%       1896, R.C.Water Supply,MoE.
%26. VicPP No. 20 of 1896, R.C.Water Supply Rept. p.185.
%27. VicPD 18/12/1896, p.4738.
%28. M.Sharland, These Verdant Plains, History Of East Loddon Shire, 1971.
%29. S.Murray, 1892, p.13.
%30. VicPP No.20 of 1896, R.C.Water Supply, Rept, p.191.
%31. H.L.Boorman, Irrigation And Water Supply Development In Victoria, 
%       1942, Govt.Printer, Melb., p.34, \& C.S.Martin 1955, p.90.
%32. H.G.McKinney, The Wentworth Irrigation Scheme, Agr.Gaz.NSW. 1896, 
%      vol.7, pp 469-470.
%33. K.Jeffcoat, More Precious Than Gold, An Illustrated History Of Water In   
%      New South Wales, 1988, p.113.
%34. W.H.Williamson, Water - From Tank Stream To Snowy Scheme, Ch. 2 in
%      A Century of Scientific Progress, Sydney 1968.
%35. C.J.Lloyd, L.A.G.Wade, 1990, ADB vol.12, p.342.
%36. S.Wells, Paddle Steamers To Cornucopia, The Renmark-Mildura 
%      Experiment of 1887, 1986, p.149.

