% $Id$
% CHAPTER NINE 
% At 1/5/99 effective length was 5948 words

\setcounter{endnote}{0}

\chapter{Irrigation Colonies at Mildura and Renmark}
\index{Mildura, Vic.}\index{Renmark, SA}\index{irrigation!colonies}
\label{ch:colonies}\addtoendnotes{\protect\subsection*{Chapter \thechapter}}
%\markboth{Chapter \thechapter. Mildura and Renmark}%
%{Pioneering Irrigation in Australia}

\fancyhead[RE]{\sffamily Chapter \thechapter.\ %
               Mildura and Renmark}

Many settlements for irrigation of relatively unoccupied land were
begun abroad late in the 19th century and became known as irrigation
colonies.  Those in southern California \index{California} were
promoted commercially, with sales of land and supply of water for
orchards and grape vines.  In the Punjab province of India,
\index{India} the British government built major irrigation canals and
made settlements, also known as canal colonies, for production of
wheat and cotton by colonists drawn from other
areas.\fn{O.\,O.~Winther, The Colony System of Southern California,
Agric.\ History, 1953, \textbf{27}, pp.\,94--102; \citet{harris1923}.}

The spread of irrigation in America encouraged one Australian in 1884
to propose an irrigation settlement on the Murray River in South
Australia but it was left to George and W.\,B.~Chaffey,
\index{Chaffey Bros} Canadian
\index{Canada} brothers with experience of Californian irrigation, to
launch the first irrigation colonies in 1887 at Mildura and Renmark on
that river.  These ventures provided models for other irrigation
schemes in Australia and therefore deserve detailed consideration.

\section*{Chaffey enterprises in California}

George Chaffey (1848--1932) and his brother William Benjamin
(1856--1926) had successfully laun\-ch\-ed two colonies in California
by 1884.  William and a younger brother left Ontario, Canada, in 1878
with parents who were persuaded by a fellow Canadian to follow his
example by acquiring land in the irrigation colony known as Riverside,
\index{Riverside, California}
then embracing about 9000\,acres of land 60 miles east of Los
Angeles. \index{Los Angeles} Founded in 1870, its orchards and vines
were then watered from canals supplied by gravity from the Santa Ana
\index{river!Santa Ana}
river.  In 1876 it gained a railway connection across the continent
and by 1880 Riverside was distinguished by its successful introduction
of Washington navel oranges to California.\fn{\citet{patterson1971}.}
As the sales of land were made by promoters also engaged in selling
water from their canal, settlers faced difficulties with water supply
as land sales improved.  Elwood Mead, \index{Mead, E.} the American
who later took charge of irrigation in Victoria, described the
situation in the region:
\begin{quote}
	the spirit of speculation in which Californian civilisation
	was born soon fastened upon irrigation and ran a mad race
	through Southern California. \ldots Men began to dream of a new
	race of millionaires, created by making merchandise of the
	melting snows, by selling `rights' to the `renting' of water,
	and collecting annual toll from a new class of society, to be
	known as `water tenants'.\fn{E.~Mead, Rise and future of
	irrigation in the United States, p.\,591 in USDA Yearbook of
	Agriculture 1899, Washington USA 1900.}
\end{quote}

George Chaffey, \index{Chaffey, G.} a mechanical engineer with
experience of steamboats on inland waterways in the United States and
Canada, visited Riverside in 1880, and decided to stay after finding
opportunities there and elsewhere in southern California for his
talents and initiative.  In 1882 the brothers began promotion of their
first irrigation colony, known as
Etiwanda. \index{irrigation!colony!Etiwanda, CA} They purchased
1000\,acres on the slopes at the foot of the lofty San Gabriel
mountains with rights to a limited water supply from mountain canyons.
Land was sold in blocks of 10\,acres on the basis that a buyer gained
shares in a mutual water company in proportion to acreage held and was
thus assured of water supply.  Its supply for irrigation was by
gravitation in concrete pipes, thus eliminating losses suffered with
earthen channels.

Encouraged by successful promotion of irrigation at Etiwanda, the
Chaffeys in 1883 laun\-ch\-ed a colony named Ontario
\index{irrigation!colony!Ontario, CA} on their purchase of
6216\,acres.  Its situation and water supply were similar to those of
Etiwanda and it was adjacent to a railway.  The appeal of growing
oranges was then important in promoting irrigation colonies in the
region, and within a few years more than 100 sales, mainly for
holdings of 10 or 20\,acres at Ontario, had been made to people from
various parts of the United States and
Canada.\fn{\citet{alexander1928}; R.~L. Gentilcore, Agric.\ History,
1960, \textbf{34}, pp.\,77--87.}

The use of concrete water pipes at Ontario was referred to in an
official Victorian report An additional feature was the establishment
of a College of Agriculture.  Ontario followed the example of
Riverside by concentration on fruit production from small
holdings.\fn{VicPP no.\,19A of 1885. RC~Water Supply, Further
Progress Rept; \citet[p.\,101]{vincent1889}.}

The success of Ontario became known to Australians visiting USA in
1885 to gain information for the Victorian Royal Commission on Water
Supply.\index{Vic RC Water Supply} Alfred Deakin, \index{Deakin, A.}
chairman of the Commission, had been advised while in San Francisco to
look at irrigation colonies in southern California.  At Los Angeles
Deakin together with two Melbourne journalists met George Chaffey, at
Ontario 35 miles distant there was William Chaffey to show them around
the settlement, and at Riverside Alfred Deakin dined with Luther Holt,
\index{Holt, L.} 
whose idea of a mutual water company had been used with advantage for
the Chaffey colonies.\fn{VicPP no.\,19 of 1885, RC~Water Supply.}

\section*{From California to Australia}

An off-hand suggestion at the Ontario colony by one of the Melbourne
journalists to William Chaffey that the brothers might do well in
Australia is the basis for claims that they were invited to develop
irrigation there.  George and W.\,B.~Chaffey independently were
tempted by the prospects in Australia and there was less reason for
remaining in California after the death of their father in 1884.  Even
before Deakin's visit to California, George had been influenced by
advice from a much-travelled Englishman, Stephen Cureton,
\index{Cureton, S.} who had
spent several months in Australia and claimed that land for irrigation
could be acquired there cheaply.  The brothers decided to send him to
Australia as their agent.  He reached Melbourne in the spring of
1885.\fn{\citet[p.\,91]{alexander1928}.}

A diary kept by Deakin indicates that Cureton called on him early in
October 1885 and had several meetings with him over three months.
Stephen Cureton, born June 1845 in Shropshire, England, gained
prominence in Victoria from his close association with the Chaffey
brothers.  Initially he was accepted in Victoria as an authority on
irrigation and horticulture, giving evidence to two royal commissions
and a public lecture later at Shepparton.  He soon advised George
Chaffey to travel to Melbourne.\fn{Diary 1885, Deakin papers 1540/2/4,
NL.}

George Chaffey reached Victoria on 13 February 1886 and began a series
of meetings with Alfred Deakin, \index{Deakin, A.} then Minister for
Water Supply, whose diary shows that they stretched over three
months. Chaffey sought a substantial grant of land on which to develop
irrigation as in California; Deakin knew the obstacles to land grants
for such a purpose but was optimistic that a deal could be arranged.
George Chaffey was sufficiently certain of the outcome to cable his
brother, advising him to come to Australia and authorising him to sell
their Californian assets.  The Chaffeys sold their Californian assets
in March 1886, possibly for the assessed value of \$310\,000
(\pounds62\,000) reported for Ontario in 1886; their expenditure to
mid 1884 was given as \$238\,500
(\pounds47\,700).\fn{\citet[pp.\,98--101]{alexander1928}; Diary 1886,
Deakin papers 1540/2/5 NL; R.\,L. Gentilcore 1960; \citet[p.\,98, re
equivalence of US\$5 to
\pounds1]{vincent1889}.}

\section*{Mildura chosen for irrigation}
\index{irrigation!colony!Mildura}\index{Mildura, Vic.}

Deakin advised George Chaffey to become familiar with the Murray
valley. \index{river!Murray} He travelled to Echuca \index{Echuca,
Vic.} and downstream as far as the Kulkyne pastoral station in the
mallee country.  There he learnt that the Mildura station, not far
distant to the northwest, might be suitable for irrigation.  It is
unlikely that this was his first information on Mildura as Cureton had
apparently called there on a trip by steamer down the Murray during
his earlier visit to Australia.  The extensive pastoral station or run
known as Mildura embraced approximately 64\,000\,acres and had been
occupied for many years by the Jamieson brothers, \index{Jamison Bros}
who until 1878 grazed thousands of sheep on its grassy country near
the river and in the adjacent mallee scrub.  Their homestead by the
river was known to many travellers along that stream who were
attracted by its garden and orchard irrigated from the river.  Rabbits
began infesting the district by 1878 and so reduced the capacity of
the current landholder, the Tapalin Pastoral Company, to graze
livestock that it became insolvent and the Commercial Bank of
Adelaide,
\index{Commercial Bank of Adelaide} its main
creditor, crashed in February
1886.\fn{\citet[pp.\,108--9]{alexander1928}; A.\,S.~Kenyon, The story
of the Mallee, Vic.\ Hist.\ Mag.\ 1914--1915;
\citet[p.\,40]{wells1986}.}

After returning to Melbourne and consulting Deakin about Mildura,
George Chaffey apparently travelled to Adelaide and then proceeded by
train and paddle steamer to Wentworth, \index{Wentworth, NSW} the busy
port at the junction of the Darling and Murray Rivers
\index{river!Murray}\index{river!Darling} and only 15 miles
from the Mildura homestead.  George Chaffey met William Paterson,
\index{Paterson, W.}  manager of the Mildura station, at Wentworth and
asked to be shown that pastoral run.  Paterson agreed only after
Chaffey produced a letter from a bank manager seeking payment from the
station lessee and identifying Chaffey as a probable buyer of the
station.  George Chaffey spent at least three autumn days at Mildura
examining pine-covered sandhills, plains covered with mallee eucalypts
and bluebush, and flats near the river covered with box eucalypts.  He
looked at the soils and subsoils, enquired about the rise and fall of
the river and the features of a particular billabong, and even got the
use of a theodolite to find the height of land above the river.  The
visitor was convinced of the suitability of this country for
irrigation and told the bewildered manager of his intention to buy the
place.  All this was told later by Paterson to Price Fletcher
\index{Fletcher, P.} (`Jethro
Tull'), the agricultural journalist of the
\textsl{Queenslander}.\fn{\citet[p.\,108]{alexander1928};
\textsl{Queenslander}, 1 Nov.\ 1890.

The timing and circumstances of these inspections by George Chaffey
\index{Chaffey, G.} of
land near the River Murray in Victoria and South Australia are
uncertain in view of Deakin's relevant diary entries for 1886 (Diary
1886, Deakin Papers 1540/2/5, NL.) which do
not indicate the appropriate long intervals between appointments.
These show:
\begin{tabbing}
        12345679012345\=123456789012345\kill
	14 February \> George Chaffey\\
	17 February \> Chaffey\\
	10 March    \> Chaffey back\\
	19 March    \> Chaffey \\
	22 March    \> Chaffey's proposal \\
	29 March    \> Chaffey \\
	 1 April    \> Chaffey \pounds500 \\
	15 April    \> Chaffey \\
	22 April    \> Chaffey \\
	23 December \> Chaffeys \\
	24 December \> Chaffeys.
\end{tabbing}
The interval of almost three weeks before 10 March should have given
time for Chaffey's inspection along the river from Echuca as far as
Kulkyne and his return to Melbourne, in which case a return visit to
Mildura via Adelaide during the first fortnight in April is
problematic considering that the Melbourne to Adelaide railway was
then incomplete \citep[p.\,100]{blake1976}.}

\section*{Negotiations for the Mildura project}

After his visit to Mildura, Chaffey started discussions with Dea\-kin
concerning transferring the Mildura land to the firm of Chaffey
Brothers.  There were many complications arising from the current
arrangements for use of the mallee \index{Mallee district} country in
Victoria.  That vast tract of public or crown land with low rainfall
and a dense cover of eucalypt scrub was divided into many large
leasehold blocks held by pastoralists, several of these made up the
Mildura station. \index{station!Mildura} The leases were now held by
the liquidators of the failed Adelaide bank and by some coincidence
the principal liquidator was a prominent South Australian politician,
Thomas Playford. \index{Playford, T.}  The only pieces of freehold
land in the region were those which could be bought to include a
station homestead and could not exceed 640\,acres.  There was such a
title for the land including the Mildura homestead on the river.  Any
arrangement to accommodate the Chaffey request would need to provide
compensation for cancellation of unexpired leases and make new
boundaries.

The availability of water for irrigation in this region of low
rainfall and hot summers was another important consideration.  Records
of the river level were available for Mildura, \index{Mildura, Vic.}
and monthly changes in the volume of river water for the past 20 years
were known for the upstream port of Echuca.  \index{Echuca, Vic.} The
Chaffeys expected to follow their Californian procedure of irrigating
fruit trees and vines, which needed abundant water in the hot dry
summer, a time of reduced river flow.  As a result of conferences in
1885 between the appropriate royal commissions in Victoria and New
South Wales, it was expected that although the River Murray does not
lie in Victoria its waters would be divided equally between the two
colonies.  That arrangement would have been prejudicial to South
Australian interests in river traffic, so the fact that a prominent
South Australian politician was now in control of the Mildura leases
meant that the Victorian government had to be careful in determining
how much water the Chaffeys should be allowed to pump from the river
for irrigation.

Another aspect of development at Mildura was transport.  The Victorian
railhead nearest to Mildura was then at Kerang, \index{Kerang, Vic.}
more than 150 miles distant by coach.  River transport upstream from
Mildura was subject to interruption when levels dropped in summer and
autumn but was then more reliable downstream thanks to seasonal inflow
from the Darling River at Wentworth.  A railway connection to
Melbourne would have involved laying an additional 160 miles of track
across unproductive mallee country.  Unless this extension was made,
the Mildura irrigation scheme would have to rely mainly on river
transport.

Negotiations between Deakin and Chaffey were completed in October
1886, when an announcement was made in the Victorian Parliament during
debate on the Waterworks Construction Encouragement Bill.  Their
agreement provided that under certain conditions the Chaffey Brothers
\index{Chaffey Bros} 
would receive a grant of 50\,000\,acres at Mildura, and the right to
acquire adjoining land, to the extent of 200\,000\,acres at a cost of
\pounds2 per acre.  The main conditions were to establish an irrigation
settlement on the area it was proposed to grant, and to make permanent
improvements on it such as irrigation channels, installation of
pumping equipment, roads and bridges, all involving expenditure to the
extent of \pounds300\,000 within 20 years, with specified outlays in
each five-year term.  As expenditure mounted, the Chaffeys would
gradually acquire freehold title to an increasing portion of the
50\,000\,acres; this was specified as one acre for each \pounds5
expenditure, with the limitation that no grant was to be for less than
640\,acres.  They were also to provide an agricultural college and
establish fruit packing and canning, and the allotments offered for
sale were to be no larger than 80\,acres for horticulture and
160\,acres for agriculture. The Chaffeys were to take water for
irrigation from the Murray but only as authorised by the
government.\fn{\citet[pp.\,118--19]{alexander1928}.}

News of this agreement provoked lively debate on the Bill in
parliament, continuing long enough to threaten its safe conduct and
thus jeopardise negotiations with the Chaffeys.  Finally, to avoid
defeat, Dea\-kin and the government were obliged to allow tenders to
be called for proposed irrigation at Mildura.  There was promise at
this stage of a rival to the Chaffeys with the announcement that a
Victorian politician was travelling to Britain to seek the necessary
finance.  Prospects for the Chaffeys looked bleak when William Chaffey
reached Melbourne late in December 1886.

\section*{Starting two settlements}

At this point, the South Australian government took the opportunity of
offering the Chaffeys land by the Murray for irrigation development as
proposed for Mildura.  Land held under pastoral lease as the Bookmark
station \index{station!Bookmark} near the Victorian border was to be
made available to the Chaffeys, who in January 1887 spent several days
together with the Conservator of Water (J.\,W.~Jones) \index{Jones,
J.\,W.} inspecting land along the Murray before accepting the offer in
February 1887 for an area later known as Renmark. \index{Renmark, SA}
They had chosen land much less elevated above the river than at
Mildura.  It ranged from clays partly subject to flooding to sandy
ground up to 60\,feet above the river and covered with native pines.
After returning from this part of the colony they soon found that
tenders for the Mildura \index{Mildura, Vic.} project had closed
without any submissions.  The Chaffeys were still interested in the
Mildura scheme, as was the Victorian government, so a new arrangement
was made between the parties to ensure the development.  Meanwhile the
South Australian proposal awaited endorsement in a new parliament to
be elected in the coming winter.\fn{\textsl{Kapunda Herald}, 18 Jan.\
1887;
\citet[pp.\,48--9]{wells1986}.}

Even before the new Victorian agreement known as the Indenture
received vice-regal consent at the end of May, the Chaffeys began work
at Mildura with survey parties.  In the coming winter the work of
clearing and ploughing began with huge steam tractors from England.
They were under the control of a young engineer, Peter McLaren,
\index{McLaren, P.} who
remained in Mildura to become resident engineer for Chaffey Brothers.
It was also time to give serious attention to Renmark as the assent to
the agreement concerning that area had been given.  George Chaffey
took up residence then at Renmark; William Chaffey was occupied mainly
with business affairs in Melbourne leading to establishment of the
firm Chaffey Brothers Ltd in October 1887, its directors being the two
Chaffeys, Stephen Cureton, \index{Cureton, S.} and William Paterson,
\index{Paterson, W.} the Mildura station
manager.  A site six miles downstream from the Mildura homestead had
first been selected for the township but it later became known as `Old
Mildura', and was devoted to irrigation of Lord Ranfurley's
\index{Ranfurley, Lord} orchards
and vineyards.  This change of plans followed Chaffeys' purchase in
August 1887 of the freehold land associated with the homestead.

As clearing and surveys proceeded, the land was divided into blocks
for sale, with access by a gridwork of roads as in their Ontario
settlement in California.  Plans were made for the system of pumps
required to supply water from the river to channels at heights from 50
to 90\,feet above the river in order to supply the higher land south
of the township and the shallow valley or trough to the southwest.  At
first only two streets in the township were distinguished by personal
names: the spacious main avenue was named for Deakin, and another
along the river frontage was named for Cureton, apparently in honour
of his efforts to establish the settlement.\fn{\citet[pp.\,56, 69,
118]{wells1986}; B.\,E.~Butler, Aeolian landscape materials and forms:
examples from the Mildura region, p.\,40, in
\textsl{Aeolian Landscapes in the Semi-arid Zone of South-eastern
Australia}. Procs Mildura Conf.\ 1979,\,1980, Wagga Wagga.}

It might have been expected that the Chaffeys would deliver irrigation
water by piping as they had done at Ontario but the slopes of the land
selected for irrigation at Mildura and Renmark were regarded as
insufficient for gravitation of water through pipes of the style used
by the Chaffeys in California.  However, South Australian patents were
taken out later by the Chaffeys for their composite pipes and pipe
moulds.  During 1887 plans were completed for the location at Mildura
of pumping stations and the positions of the main irrigation
channels.\fn{\citet[p.\,35]{cumming1986}.}

Meanwhile at Renmark progress had been made since August 1887 with
surveying and clearing about 4000\,acres of low-lying land near an
effluent of the river.  George Chaffey's move there put him in a good
position to make adequate arrangements for the river transport vital
to the development of both settlements.  His enterprise soon brought
about the formation of the River Murray Navigation Co Ltd with
possession of steam-boats and the services of Captain Hugh King,
\index{King, H.} an experienced river-man. 
Subsequently Charles Chaffey, \index{Chaffey, C.} the third brother,
took up residence at Renmark as manager of the project and remained
there for several years.\fn{\citet[pp.\,84--6]{wells1986}.}

Irrigation at both sites required centrifugal pumps
\index{pump!centrifugal} to lift water from
the river in stages.  At Mildura the supply of water to channels up to
85\,feet above the river would involve four stages; at Renmark three
stages were needed.  The initial lift at both places involved pumping
from the river into abandoned channels or effluents which could serve
also as reservoirs.

During 1888 contractors at both settlements made earthen channels,
cemented in places to reduce seepage; settlers began planting their
allotments with fruit trees and vines; and some equipment was
installed to pump water from the river.  Temporary arrangements
allowed watering of some 80\,acres at Mildura using a centrifugal pump
and a portable engine and limited irrigation was also undertaken at
Renmark.\fn{\citet[p.\,127]{vincent1889}.}

At Mildura George Chaffey made use of imported triple-expansion steam
engines coupled with centrifugal pumps; \index{pump!steam-driven} some
were delivered in 1888 and the most powerful were made by Tangyes
\index{pump!Tangye} in England to his own design and supplied in 1889.
Less powerful engines and pumps were provided at Renmark to cope with
the smaller area involved and the lower lifts.  Generation of steam by
boilers at the pump sites required substantial supplies of local
firewood.  The Mildura Irrigation Company
\index{irrigation!company!Mildura} and the Renmark Irrigation Company,
\index{irrigation!company!Renmark}  corresponding to the mutual irrigation
companies introduced in California, were formed in 1888 and 1889 to
deal specifically with the supply of irrigation
water.\fn{\citet[pp.\,114--127]{wells1986}.}

Sales of land at the settlements were promoted by imaginative
advertisements and special excursions from Melbourne and Adelaide.
One particularly effective publication was the book entitled `The
Australian Irrigation Colonies', known widely as the Red Book from its
coloured cover.  This appeared in 1888, with a further edition in
1889; many settlers came from Britain in response to its advice on
opportunities to profit from production of oranges, wine, raisins,
olives and fruit from deciduous trees.\fn{\citet{vincent1889}.}

Changes were made during 1888 in the control of the main Chaffey
company.  Stephen Cureton, \index{Cureton, S.} whose involvement in
the irrigation scheme had been important at its inception, retired as
a director of Chaffey Bros Ltd, \index{Chaffey Bros} took out a large
sum of money as his share in the company assets and was replaced by
Jonas Levien (1840--1906), \index{Levien, J.} a Victorian politician
with horticultural experience and a reputation for ability to arrange
financial backing for his enterprises.  Levien soon replaced George
Chaffey as chairman of the board of directors.

With the successful testing of the powerful Chaffey pumping engine
early in 1890, George Chaffey completed his major contribution to
irrigation at Mildura.  Then troubles developed over management of
irrigation there.  W.\,M.~Paterson, \index{Paterson, W.} the Mildura
manager of Chaffey Bros Ltd and a director of the Mildura Irrigation
Co, was replaced as manager by W.\,J.~Waddingham, \index{Waddingham,
W.\,J.} a relative of the Chaffeys.  Paterson then became involved in
local discontent during 1891 and 1892 over charges to be made by the
Irrigation Company: there were claims by some irrigators that no
charges should be made for the supply of water, and in fact no charges
were made for irrigation water before 1890.  The Victorian government
appointed A.\,W.~Howitt \index{Howitt, A.\,W.} in 1892 to make
enquiries, received his report upholding a rating system for Mildura,
and subsequently introduced legislation which became the subject of
another enquiry by a Select Committee in 1892.  The outcome was the
\index{Vic Mildura Rating Act 1893}
Mildura Rating Act.\fn{VicPP 1892--93, V\&P LA vol.\,1 2nd Sess., Rept
Select Comm.\ LA upon the Mildura Settlement;
\citet[pp.\,143--51]{wells1986}.}

Irrigators at Renmark were also concerned about management of the
local irrigation company, then virtually controlled by Chaffey
Brothers.  However, in 1892 after transferring control to the settlers
the Chaffeys contracted to operate the irrigation system.  Next year
the irrigation company was replaced by the Renmark Irrigation
Trust. \index{trust!irrigation!Renmark} This was achieved through
legislation in 1893 which provided for management of the trust by a
board of seven elected irrigators.\fn{\citet[p.\,149]{wells1986}.}

Despite many difficulties in developing the irrigation colonies,
prog\-ress was achieved in the extension of cultivation and the growth
of population.  In 1891 the Chaffeys had sold 17\,000\,acres at
Mildura and 4500\,acres at Renmark; their settlers had planted
6500\,acres at Mildura and about 1500\,acres at Renmark; and
populations were 4000 and 600 respectively.  Mildura in 1894 had more
than 4000\,acres planted with fruit trees including apricots, citrus,
peaches and olives, less than 3500 acres with vines mainly for
raisins, and 700\,acres with annual or green crops.  Renmark in 1895
had 2864\,acres under irrigation with almost half that area planted
with deciduous fruit trees and citrus and the balance made up almost
equally by vines and fodder crops.\fn{CSIR Bulletin 133, A soil survey
of the Mildura irrigation settlement, 1940; \citet[p.\,6]{murray1892};
\citet[p.\,157]{wells1986}.}

Transport became more of a problem when the settlements reached the
stage of producing goods for sale.  Despite encouraging remarks from
Richard Speight, \index{Speight, R.} chairman of the Victorian Railway
Commissioners, on his visit to Mildura in November 1888, nothing had
been done to provide a railway \index{railway} to that settlement and
it remained dependent on river transport.  This was reliable except
during periods of low river levels and these unfortunately occurred in
several years during summer and autumn when fresh fruit awaited
transport to markets.  George Chaffey tried to cope with the problem
of low river levels by importing an American boat,
\index{paddle-steamer} the stern-wheel steamer
\textsl{Pearl}, which had a  low draft, but it was not a success.
The rail connection to Melbourne was not achieved until late in 1903.

Renmark was better placed except in exceptional droughts such as that
of 1914 when the river virtually dried up.  Citrus growers had more
success: in June 1892 the first exports to England were made, and in
1893 hundreds of cases of oranges and lemons reached England and
America.  In order to cope with marketing stone fruit and vine fruits
it was necessary to resort to drying and success with this process
firmly established the dried fruit industry in the two irrigation
settlements.\fn{\citet[p.\,120]{vincent1889};
\citet[pp.\,88--93, 265]{wells1986}; \citet[p.\,125]{hill1937}.}

\section*{Waterlogging and salinity}
\index{waterlogging}\index{salinity}

After irrigation was provided for the orchards and vineyards, many
settlers became aware of unexpected problems associated with watering.
The unlined earthen channels were subject to seepage \index{seepage}
or even collapse after burrowing by freshwater crayfish (`yabbies')
introduced from the river.  This loss of water added to the expense of
pumping from the river and it also involved damage to fruit trees and
vines from waterlogging and salinity.  Seepage was detected at Mildura
in 1892 when it extended 200\,feet from the banks of certain channels.
At Renmark the loss from supply channels was estimated at 20 per cent
before 1895 while at Mildura where the length of main channels
exceeded 100 miles in 1892, the loss was reported to be more than
half. Seepage from channels can be reduced or averted by lining the
earthen structures with impermeable material such as
concrete.\fn{\citet[p.\,198]{alexander1928}; SAPP no.\,113 of
1895. Select Committee Village Settlements etc., Rept; VictPP no.\,19
of 1896, RC~Mildura, Rept, p.\,vi.}

Excessive salinity in the soil was recognised soon after irrigation at
Mildura and Renmark.  This feature might have been anticipated from
soil analyses \index{soil analysis} and it is possible that the
Chaffeys had relevant information when they expressed their interest
in soils during an interview, probably in 1887:
\begin{quote}
	\ldots we have had analyses made of the soil and compared them
	with the results of analyses made of soil from
	successfully-irrigat\-ed colonies in
	America.\fn{\citet[p.\,98, quotation from \textsl{Melbourne
	Argus}]{vincent1889}.}
\end{quote}
George Chaffey told the Victorian Select Committee on Mildura in 1892
that he examined the soils before irrigation channels were made and
found there was an `immense quantity of alkalis' which were `not
injurious to plant life'.\fn{VicPP V\&P LA vol.\,1, 1892--93, 2nd
sess.\ Select Comm.\ Mildura Settlement. Rept \& MoE, Q818--819.}

The first indication that salinity could be a disadvantage to
Australian irrigators came soon after irrigation at Mildura
\index{Mildura, Vic.}  in 1888.
Land had been cleared and cultivated in 1887 and in 1888 irrigation
was begun though only 120\,acres had been planted.  In February 1889
four men from the Western Wimmera Irrigation Trust
\index{trust!irrigation!Western Wimmera} visited Mildura and
reported the existence of a white salty crust on land that had been
watered. A few years later there was more definite evidence that
salinity \index{salinity} was a problem for irrigators at
Mildura. James Henshilwood \index{Henshilwood, J.}  had taken up a
block of 40\,acres on land previously covered by mallee \index{mallee}
vegetation.  He had planted fruit trees of various types in 1890.  In
the next two years some trees failed to make satisfactory growth and
were replaced.  Samples of his soil were sent to Melbourne for
analysis, which showed that one contained `a great excess of salt'.
He was advised to install deep open drains and wash out the salt.  He
had further trouble with citrus though stone fruit did better.
Apparently uncertain of the diagnosis given by the Melbourne analyst,
Henshilwood wrote to Professor Hilgard in California in 1895 for
advice on treatment of his soil.  Hilgard's reply was directed to
finding if the soil also had a problem of alkalinity and asked for
samples of soil.  When these were sent in 1895, the results of
analysis showed no alkalinity, \index{soil alkalinity} and the very low
concentration of salt could be explained as the result of the high
rainfall (30\,inches) in that year.\fn{\citet[p.\,73]{wells1986};
Silenus (A.\,S.~Kenyon), Early Mildura as others saw
it. \textsl{Sunraysia Daily}, 19 Aug.\ 1921; VicPP no.\,19 of 1896,
RC~Mildura, Appendix~A2, pp.\,277--279.}

Before 1900 settlers at Mildura and Renmark were familiar with the
appearance of waterlogging, seepage, and the effects of soil salinity
or alkalinity, all having some ill effects on fruit trees and vines.
The Chaffey brothers apparently had no experience of these problems in
their Californian settlements, but Professor Hilgard's advice on
under-drainage of Californian soils affected by salinity or alkalinity
after irrigation was taken up there by several landholders.  By 1907
individual landholders at Mildura had installed tile drains which
discharged via shafts or bores to underlying beds of sand at depths of
80 to 90\,feet.  This method was less reliable at
Renmark.\fn{A.\,S.~Kenyon, Drainage and irrigation, J.\,Agric.\ Vic.\
1907, \textbf{5}, p.\,206; CSIR Bull.\ no.\,133, 1940, Soil survey of
the Mildura irrigation settlement; CSIR Bull.\ no.\,56, Soil survey of
Renmark irrigation district, South Australia.}

\section*{Chaffey Bros Ltd becomes insolvent}
\index{Chaffey Bros}

While various technical and administrative problems of the two
settlements were being dealt with, a major financial crisis for the
Chaffeys was looming.  Much of their success had occurred during the
prosperity enjoyed in Victoria in the late 1880s following British
investments.  But in the eight months from July 1891 twenty major
financial institutions closed and British investments were curtailed.
The Chaffeys tried to avoid involvement by making personal appearances
in England to raise funds but failed after reports were circulated
there of the difficulties confronting settlers in the irrigation
colonies.  There was worse to come in 1893 when banks in Melbourne
suspended trading and the boom collapsed.  Some settlers left Mildura
and Renmark, seeking work even as far away as the new goldfields in
Western Australia.  Then Chaffey Bros Ltd went into liquidation in
1895 and legislation was passed enabling the establishment of the
First Mildura Irrigation Trust \index{trust!irrigation!First Mildura}
with board members elected by settlers in much the same way as for the
Renmark Irrigation Trust. \index{trust!irrigation!Renmark} A few
months later a Royal Commission was appointed to enquire into
circumstances of the Mildura settlement.  It was followed in South
Australia by a similar enquiry concerning Renmark and other
settlements along the river.\fn{\citet[pp.\,26, 81]{cannon1972}.}

The report of the Victorian royal commission \index{Vic RC Mildura}
found the Chaffeys and the Victorian government had contributed to the
debacle at Mildura.  Many witnesses, including George Chaffey, Alfred
Deakin, and Stephen Cureton had been called to give evidence before
the commissioners.  They found that several provisions of the
indenture of May 1887 had not been met, though these deficiencies did
not include the expenditures of money required to obtain the crown
grants of land.  The enquiry gave attention particularly to the
arrangements for irrigation, which were then in a critical state; as a
result recommendations were made in favour of a loan to ensure the
lining of existing channels, to improve the pumping plant and to make
sure of a water supply during the coming irrigation season.\fn{VicPP
no.\,19 of 1896, RC~Mildura, Rept.}

One outcome of the debacle involving the Chaffey irrigation colonies
on the Murray was the return of George Chaffey \index{Chaffey, G.} to
America after his financial ruin and criticism of his irrigation
development.  An extensive defence of his work was given much later by
his biographer. W.\,B.~Chaffey \index{Chaffey, W.\,B.} remained a
respected figure at Mildura where his fortunes gradually
improved.\fn{\citet{alexander1928}.}

The South Australian \index{SA RC Renmark \& Murray R. Settlements}
enquiry took evidence from many Renmark settlers and some members of
the managerial staff, with general agreement that serious seepage from
unlined earthen channels had occurred.  Recommendations in the
progress report of 1899 were concerned main\-ly with maintenance of
the water supply channels, including provision of funds for the
purpose.\fn{SAPP no.\,37 of 1899, RC~Renmark and Murray River
Settlements, Progress Rept and MoE.}

Improvements in the irrigation systems of the settlements were made
after the two irrigation trusts became responsible, but there were
recurrent difficulties with water supplies when the river flow was
seriously diminished in periods of drought.  Some future improvement
in water supplies from the river was indicated when the Commonwealth
and three States reached agreement in 1915 on management of the river
by provision of weirs and locks and construction of storages.

Despite the reduction in seepage \index{seepage} losses from channels
as their lining with cement extended, waterlogging due to irrigation
forced many settlers to develop individual drainage
systems. \index{drainage} The horticultural industry continued to be
based on production of dried fruit, which gave good financial returns
during the 1914--18 war.  Plantings of wine grapevines began early in
the development of the irrigation colonies and led to establishment of
wineries and distilleries at Mildura and Renmark.\fn{\citet[pp.\,123,
236]{wells1986}.}

\closure
The Mildura and Renmark colonies transferred much of the Californian
methods of irrigation for establishment of horticulture and
viticulture in a region regarded previously by many Australians as
almost unproductive.  The venture by the Chaffeys \index{Chaffey Bros}
captured popular attention at a time of prosperity but almost
collapsed ten years after it began.  The Chaffeys followed the same
procedure as they had used for promotion of their earlier settlements
in California but their Australian ventures were more ambitious and
lacked the access to reliable transport and extensive markets that
enabled their earlier ventures to flourish.  George and W.\,B.~Chaffey
were an impressive pair but they took on responsibilities beyond their
capacity to maintain.  When the major engineering problems of the two
irrigation settlements had been attended to, George was tempted by the
scope for irrigation in other parts of Australia and W.\,B.~Chaffey
had to spend time abroad after personal tragedy.  Management of
affairs at Renmark was handled successfully for a time by the youngest
brother, Charles, while serious problems at Mildura were proving
beyond the capacity of new managers there.

A basic problem of these ventures was that financing development
depend\-ed largely on sales of land.  This method succeeded initially
but could not cope with problems encountered early in the 1890s when
sales declined.

The lack of reliable transport to Mildura was a serious disadvantage
which forced dependence on production of dried or canned fruit but the
Chaffeys failed to develop a canning facility there.  Apart from a few
Americans brought to Mildura and Renmark by the Chaffeys, the settlers
had no prior experience of irrigation.

%\section*{References}
%1. O.O.Winther, The Colony System Of Southern California, Agric.History, 
%     1953, vol.27, pp. 94-102.
%2. D.G.Harris, Irrigation In India, 1923, London.
%3. T.Patterson, A Colony For California, Riverside's First Hundred Years, 
%     1971.
%4. E.Mead, Rise And Future Of Irrigation In The United States, p. 591 in 
%     USDA Yearbook of Agriculture 1899, Washington USA 1900.
%5. J.A.Alexander, The Life Of George Chaffey, 1928, \& R.L.Gentilcore, 
%    Agric.History, 1960, vol.34,p.77-87.
%6. VicPP No. 19A of 1885. R.C.Water Supply, Further Progress Rept. \&
%     J.E.M.Vincent, The Australian Irrigation Colonies On The River Murray In 
%     Victoria And South Australia, 1889, p.101.
%7. VicPP No. 19 of 1885, R.C.Water Supply.
%8. J.A.Alexander, 1928, p.91.
%9. Diary 1885, Deakin papers 1540/2/4, NL.
%10. J.A.Alexander, 1928, p.101, \& Diary 1886, Deakin papers 1540/2/5 NL.
%11. J.A.Alexander, 1928,p.98, R.L.Gentilcore 1960, \&
%      J.E.M.Vincent,1889, p.98, re equivalence of US\$5 to \pounds1.
%12. J.A.Alexander, p.108.
%13. J.A.Alexander, p.109, \& A.S.Kenyon, The Story Of The Mallee, Vict.Hist 
%     Mag. 1914-1915.
%14. S.Wells, Paddle Steamers To Cornucopia, The Renmark-Mildura 
%       Experiment of 1887, 1986, p.40.
%15. J.A.Alexander, 1928, p.108.
%16. Queenslander, 1/11/1890.
%17. Diary 1886, Deakin Papers 1540/2/5, NL.
%18. L.J.Blake, The Land Of The Lowan, 100 Years In Nhill And West 
%     Wimmera, 1976, p.100.
%19. J.A.Alexander, 1928, p.118.
%20. J.A.Alexander, 1928, p.119.
%21. Kapunda Herald, 18/1/1887 \&  S.Wells,1986, pp. 48-49,65.
%22. S.Wells, 1986, p.118.
%23. S.Wells, 1986, p.69.
%24. S.Wells, 1986, p.56.
%25. B.E.Butler, Aeolian Landscape Materials And Forms: Examples From The
%      Mildura Region. , p.40, in Aeolian Landscapes In The Semi-arid Zone Of
%      South-Eastern Australia. Procs Mildura Conf. 1979, 1980, Wagga Wagga.   
%26. D.A.Cumming and G.Moxham, They Built South Australia, p.35,1986, 
%      Adelaide.
%27. S.Wells, 1986,pp84-86.
%28. J.E.M.Vincent, The Australian Irrigation Colonies, 1889, p.127.
%29. S.Wells, 1986, pp114-115.
%30. S.Wells, 1986, p.127.
%31. J.E.M.Vincent, The Australian Irrigation Colonies 1889. 
%32. S.Wells, 1986, p.150-51.
%33. VicPP 1892-93. V.\&P LA Vol.1 2nd Sess., Rept Select Comm.
%      LA upon the Mildura Settlement \& S.Wells, 1986, pp 143-44.
%34. S.Wells, 1986, p.149
%35. CSIR Bulletin 133, A Soil Survey Of The Mildura Irrigation Settlement,
%      1940, \& S. Murray, Irrigation In Victoria, 1892 p.6.
%36. S.Wells, 1986, p.157.
%37. J.E.M.Vincent, 1889, p.120.
%38. S.Wells, 1986, p.265.
%39. S.Wells, 1986, p.88-89.
%40. Ernestine Hill, Water Into Gold, 1937, p.125.
%41. J.A.Alexander, 1928, p.198.
%42. SAPP No. 113 of 1895. Select Committee Village Settlements etc. Rept.
%      \&  VictPP No.19 of 1896, R.C.Mildura, Rept. p.vi.
%43.  J.E.M.Vincent, 1889, p.98 quotation from Melbourne Argus.
%44. VicPP V \& P LA Vol.1, 1892-93, 2nd sess. Select Comm. 
%       Mildura Settlement. Rept. \& MoE, Q818-819.
%45. S.Wells, 1986, p.73.
%46. Silenus(A.S.Kenyon),Early Mildura As Others Saw It. Sunraysia Daily,
%      19/8/1921.
%47. VicPP No.19 of 1896, R.C.Mildura, Appendix A2,pp. 277-279.
%48. A.S.Kenyon, Drainage And Irrigation, J.Agric.Vict. 1907, vol.5,
%      206, \& .  CSIR Bull. No.133, 1940, Soil Survey Of The Mildura
%      Irrigation Settlement.
%49. CSIR Bull. No.56, Soil Survey of . . . . Renmark Irrigation
%       District, South Australia.
%50. M.Cannon, The Land Boomers, 1986, p.26.
%51. M.Cannon, 1986, p.81.
%52. VicPP No.19 of 1896, R.C.Mildura, Rept.
%53. J.A.Alexander, 1928.
%54. SAPP No. 37 of 1899, R.C.Renmark and Murray River Settlements,
%      Progress Rept and MoE.
%55. S.Wells, 1986, pp 123 \& 236.

