% $Id$
% CHAPTER TEN
% 6160 words at 2/5/99

\setcounter{endnote}{0}

\chapter{Proposals for Other Irrigation Colonies}
\index{irrigation!colonies!proposed}
\label{ch:proposals}\addtoendnotes{\protect\subsection*{Chapter \thechapter}}
%\markboth{Chapter \thechapter. Proposals for Other Colonies}%
%{Pioneering Irrigation in Australia}

\fancyhead[RE]{\sffamily \small Chapter \thechapter.\ %
               Proposals for Other Colonies}

More than twenty irrigation colonies were proposed in Victoria, New
South Wales and South Australia, generally in response to progress
with those at Mildura and Rermark.  They were all initiated by private
companies which had to rely on governments or semi-governmental bodies
for access to a water supply. With one exception, the following
schemes were proposed in a period of about ten years from 1884.  Eight
colonies, all in Victoria, were the only ones to come into existence
and survive after 1920.

\section*{Murray Valley}

\subsection*{Morgan, SA} \index{Morgan, SA}

Frederick Ludwig Bevilaqua (1832--1913) \index{Bevilaqua, F.\,L.} came
to South Australia from Germany in 1854 and for many years was a
trader in the South-East and the Barossa Valley. \index{Barossa
valley, SA} His interest in irrigation at first in the 1860s went as
far as an unsuccessful attempt to use water from the Murray River
\index{river!Murray} near Mannum. \index{Mannum, SA} In 1884 he
launched his proposal for an irrigation settlement near Morgan to be
developed by the South Australian Irrigation Company
\index{irrigation!company!SA} on 25\,000\,acres, part of which was to
be subdivided into holdings of 50 to 100\,acres irrigated from the
Murray by means of pumps.  The chosen area was approved by George
Goyder, Surveyor-General, \index{Goyder, G.\,W.}  and endorsed by John
Custance, \index{Custace, J.} Professor of Agriculture at the recently
established Roseworthy Agricultural College. 
\index{Roseworthy Agricult.\ College, SA} Bevilaqua was aware of
contemporary interest in irrigation as expressed in Victoria by the
Australian Irrigationist and the newspaper reports of American
irrigation provided by T.\,K.~Dow.\index{Dow, T.\,K.}  However,
despite further enquiries, nothing came of Bevilaqua's
proposal.\fn{F.~Bevilaqua,
\textit{General Remarks and Detail Accounts of the South Australian
Irrigation Company}, 1884; SAPP no.\,158 of 1884; SAPP no.\,73 of
1885.}

\subsection*{Lake Bonney, SA} \index{lake!Bonney}

Lake Bonney, maintained by inflow from the Murray River, lies about 20
miles west of Renmark.  It was mentioned in a scheme for irrigation
proposed in 1888 by business men in Melbourne and Adelaide.  They
sought the use of 150\,000\,acres of land, then held under pastoral
lease on the Cobdogla pastoral station. \index{station!Cobdogla} The
group, soon identified as the Lake Bonney Irrigation Company,
\index{irrigation!company!Lake Bonney} began negotiations with South
Australian government officials in May 1888 and arranged surveys to
determine levels and the possible location of irrigation
channels. Their application went to the South Australian government
and was publicised in the Adelaide press, with some expressions of
opposition from those preferring development by the government rather
than a company.  It was intended that the company would spend
\pounds400\,000 on the scheme over 20 years.  The application was
considered by the South Australian parliament during September and
October but was not agreed to.  One stumbling block was the
government's refusal to allow subdivision of the irrigable land into
lots of 1000\,acres or more.\fn{\textit{Adelaide Observer} 22 Sep.\ 1888
\& 29 Sep.\ 1888; SAPD 1888, p.\,1267, 9 Oct.\ 1888.}

One of the promoters was Jonas Levien \index{Levien, J.} who had
horticultural interests at Geelong and had been Minister for Mines and
Agriculture before opposing the Mildura agreement submitted by Deakin
to the Victorian Parliament in 1886.  Two others, Benjamin Fink
\index{Fink, B.} and
R.\,W.~Best, \index{Best, R.\,W.} were prominent in the Victorian land
boom.  It is surprising that Jonas Levien was so involved with this
scheme yet in August 1888, during the period of negotiations with the
SA government by the Lake Bonney Irrigation Company, he became
chairman of the board of directors of Chaffey Bros Ltd, \index{Chaffey
Bros} whose interests he had contested
previously.\fn{\cite{cannon1972}}

Irrigation of at least part of the land near Lake Bonney was
sufficiently practicable and attractive to warrant further interest,
leading to important development in the district thirty years later.

\subsection*{Milang, SA}\index{Milang, SA}

Albert Henry Landseer \index{Landseer, A.\,H.} was engaged in trade
along the Murray River from his headquarters at Milang on Lake
Alexandrina.  \index{lake!Alexandrina} He promoted an irrigation
scheme close to Milang, to be supplied with water pumped from the lake
to an area of more than 15\,000\,acres.  The area was inspected in May
1892 by Professor Lowrie \index{Lowrie, Prof} of Roseworthy College
\index{Roseworthy Agricult.\ College, SA} 
who regarded the land as third-rate in quality and advised that only a
small scheme should be tried initially.  In July a parliamentary
delegation including two government ministers inspected the land under
consideration and the short-lived Holder government was apparently in
favour of a trial for the scheme.  Nothing came of this scheme but
interest in local irrigation was revived in 1893 when an English
syndicate was reported to be negotiating for irrigation of 2000\,acres
of land close to Milang.  This project also came to
nothing.\fn{\cite{stimson1983};
\textit{Southern Argus}, 5 May 1892, 23 June 1892, 20 Oct.\ 1892 \&
2~Feb.\ 1893;
\cite{faull1981}.}

\subsection*{Reedy Creek, SA} \index{creek!Reedy}

An irrigation project proposed in 1888 by J.\,D.~Derry, C.\,N.~Long,
and W.\,R. San\-do \index{Derry, J.\,D.} \index{Long, C.\,N.}
\index{Sando, W.\,R.} involved a storage reservoir with a capacity of
17\,000\,acre-feet to impound the waters of Reedy Creek, which
discharges into the Murray \index{river!Murray} near Mannum,
\index{Mannum, SA} and
construction of distribution channels for irrigation of 20\,000\,acres in
the Mannum Irrigation Colony. \index{irrigation!colony!Mannum} The
promoters expected to spend
\pounds80\,000 on construction.  Their application was mentioned in
the 1890 report on South Australian public works; some details of the
scheme were provided later by Alfred Deakin.  Two years later two of
the promoters had withdrawn leaving the project entirely to
J.\,D.~Derry, the Victorian engineer involved with irrigation near
Horsham, but nothing eventuated before Derry's return to England in
1894.\fn{SAPP no.\,29 of 1890, no.\,29 of 1892,
\cite[p.\,92]{deakin1892};
\cite{blake1972}.}

\subsection*{Lake Boga Irrigation Company, Vic.} \index{lake!Boga}

Lake Boga is one of the larger freshwater lakes fed by overflow of the
Murray River along the Little Murray River \index{river!Little Murray}
in northwestern Victoria.  The Lake Boga Irrigation Co.\
\index{irrigation!company!Lake Boga} was concerned with development of
irrigated horticulture on land close to Lake Boga and the nearby
smaller Long Lake, \index{lake!Long} between which the railway had
been extended from Kerang \index{Kerang, Vic.} to Swan Hill
\index{Swan Hill, Vic.}  in 1889.  Water for irrigation was to be
pumped from the lakes.\fn{\cite[p.\,77]{wells1986}.}

The scheme for irrigation was advertised in at least two district
newspapers and a Melbourne weekly in 1890 after formation of the
company.  Its prospectus showed an intention to sell township
allotments (20), horticultural allotments (135) ranging in size from 5
to 12\,acres and totalling 600\,acres, as well as an area of
106\,square miles to the south and west held under mallee pastoral
lease. Six men responsible for the project included three from St
Arnaud, and one each from Ballarat, Melbourne, and Torrumbarry.  Three
were engineers, including George Gordon, formerly of the Victorian
Water Supply Department.  Excursions to the area from Donald were
advertised in November 1890, after earlier efforts to sell allotments
by auction.  One advertisement stressed the accessibility of the
project area, with disparaging reference to the Mildura irrigation
settlement as `remote and outlandish'.\fn{G.~Gordon papers, File
H17329, LaTL; \textit{Swan Hill Guardian} 19 Nov.\ 1890.}

By 1892 the company had sold 7590\,acres, mainly for dry-farm\-ing and
without much progress in selling the irrigable blocks.  The company
continued in existence for at least 10 more years but it failed to
make any significant development of irrigation.  Its difficulty arose
partly from the refusal of the Water Supply Department to allow
pumping from Lake Boga, a decision apparently prompted by government
support for the Mildura scheme.  This attempt at irrigation in the
vicinity of Lake Boga may have influenced a later effort there by
another company.\fn{G.~Gordon papers, File H17329, LaTL;
\cite[p.\,262]{barr1992}.}

\subsection*{Tresco, Vic.}
\index{Tresco, Vic.}

Irrigation at Tresco was promoted in 1913 by Australian Farms Ltd, a
company recently set up to establish farms in different parts of
Australia and New Zealand.  Its foundation directors were mainly
Victorian graziers, and it had the services of R.\,V.~Billis,
\index{Billis, R.\,V.} formerly
a public servant dealing with projects for closer settlement and
immigration.

The company first developed irrigation at the margin of the riverine
tract between Kerang and Swan Hill.  A large area of wheat-farming
land adjacent to the Swan Hill railway line was purchased from a local
resident (Mr~Cornish) and a scheme was promoted for irrigation of
fruit trees and vines on about 4000\,acres divided into small
allotments, with water to be pumped from Lake Boga, \index{lake!Boga}
filled by overflow from the Murray River.  Sales of land in this area
known as Tresco were enhanced by the scope for speculators to purchase
land without being obliged to reside there.  Encouraged by the
response, the company acquired more land, firstly an estate adjoining
the southern part of Tresco and known as Mystic Park, and in 1919 land
to the west of Tresco, known as Tresco West.  Tresco Irrigation Ltd
\index{Tresco Irrigation Ltd}
was the company which took over the plant belonging to Australian
Farms Ltd at Lake Boga and Tresco; it was the body which had to deal
with the problem of salinity which emerged by
1920.\fn{\cite[p.\,328]{barr1992}.}

The settlement suffered its first setback from salinity about five
years after its establishment in 1913.  Citrus and vines had been
planted over about 2500\,acres, including two ridges with sandy
topsoil and extensive flats with loamy soils.  Salt damage
\index{salinity} first became evident on the flats after a water-table
developed close to the surface and salt incrustations were noticed on
the surface of the ground.  Drains were installed but their outfall to
Lake Boga probably added to the salt content of its water, which was
apparently excessive before 1920. The damage extended to all the flat
land; satisfactory citrus orchards remained only on the high sandy
ridges.\fn{F.\,M.~Read, J.\,Agric.\ Vic.\ 1930, \textbf{28}, p.\,65 \&
1931 \textbf{29}, pp.\,551--563.}

The Tresco settlement had attracted investment by several men
pro\-minent in public affairs.  They found results at Tresco so
disappointing that in 1920 they looked for an alternative site for
irrigation development and chose land to the west of Kangaroo Lake,
south of the Mystic Park area.\fn{\cite{trengrove1969}.}

\subsection*{Torrumbarry North, Vic.}
\index{Torrumbarry North, Vic.}

An area of almost 20\,000\,acres near the Murray River between Echuca
and Kerang was acquired by a company in 1888 with the intention of
subdivision into small holdings to be irrigated for fruit-growing.
Negotiation with the Government led to the constitution in March 1889
of the Torrumbarry North Irrigation Trust
\index{irrigation!trust!Torrumbarry North} which was granted by the
government a loan of
\pounds12\,300, spent on machinery and an engine house.  The company
first involved with the scheme went into liquidation, apparently after
the collapse of the Victorian land boom, and the trust recorded
irrigation only in 1891, of 17\,acres under fruit trees and vines.  In
1896 the irrigation trust was practically non-existent and the land
was owned by the Murray River Irrigation Estate Co Ltd.\fn{VicPP
nos.\,20 \& 21 of 1896, RC~Water Supply, Rept pp.\,176--180, MoE
p.\,438 \& Appendix~C.}

\section*{Goulburn River Valley}
\index{river!Goulburn}

\subsection*{Ardmona, Vic.}
\index{Ardmona, Vic.}

The Ardmona irrigation settlement was the first established in the
valley of the Goulburn River, the major Victorian tributary of the
Murray.  Its development took place on small holdings formed over a
period of years by successive subdivision of larger holdings.  Ardmona
was the name given to one of these holdings by its owners, G.~and
L.~McDonald, \index{McDonald, G. \& L.}  who came to the district
along with many other settlers.  Cereal growing was undertaken with
reasonable harvests in the 1870s but droughts in the early 80s caused
hardship.  Adequate supplies of water were needed but settlers had to
depend on wells or cartage from the Goulburn River.

Legislation in 1881 providing for establishment of waterworks trusts
facilitated a scheme for distribution of river water by open channels
to the many farms in the region lying between the Goulburn, the
Campaspe, and the Murray Rivers.  The principal aim of the scheme was
to provide water for stock and domestic purposes; irrigation was a
minor consideration.  A weir on the Goulburn would be needed to divert
water into the channel system and its construction, under
consideration from 1884, was started in 1887 and did not provide water
for irrigation until the spring of 1891. Meanwhile a pumping plant was
installed and began operation in September 1885 to supply water to the
distribution channels then
completed.\fn{\cite[pp.\,105--106]{bossence1965}; \textit{Aqua},
Nov.\ 1968, p.\,53.}

A few landholders near Mooroopna \index{Mooroopna, Vic.} on the west
bank of the Goulburn were interested in horticulture and viticulture
before water supplies became available.  They together took steps to
create small holdings intended for fruit and vines.  A portion of
W.~Davis's farm was acquired by Fred J.~Young, \index{Young, F.\,J.} a
local farmer since 1870, who subdivided it in 1886 to provide five
allotments varying in size from 8 to 30\,acres. One was bought by John
West, editor since 1883 of the local paper, \textit{Goulburn Valley
Yeoman}.  An adjoining farm of 733\,acres, known as Ardmona, was
purchased in 1886 by five members of a syndicate: West, Michael
Kavanagh, Fred Young, M.~Cussen, and A.\,B.~Patterson, with the
intention of subdivision.  Then George Pagan bought an adjoining
property of 329\,acres for subdivision and began selling allotments a
few years later.\fn{\cite{steven1990};
\cite[pp.\,57--58]{martin1955}.}

Vines and fruit trees planted earlier by Michael Kavanagh,
\index{Kavanagh, M.} a Canadian,
on his `Lake Erie' farm near Ardmona, and by Fred Young, produced
grapes \index{grapes} for wine or raisins and currants as well as
fresh fruit for sale.  John West \index{West, J.} began a plant
nursery and orchard on his block.  All three were able to continue
production without irrigation but they were ready to make use of it
when water became available in 1886.  In that year the rainfall in the
district was low in autumn but otherwise normal (18.6\,inches at
Shepparton) and in 1887 rains were more than adequate (26.3\,inches).
The next year brought a drought and there was insufficient water to
allow irrigation.  Despite that handicap, John West gained first prize
in the government competition for the best variety of irrigated
crops,. His farm, named Milvina, was reported as being a portion of
`Fresno' colony.\fn{\textit{Australasian}, 18 Feb.\ 1888, p.\,354.}

In 1890 the Minister for Water Supply, Alfred Deakin \index{Deakin,
A.} announced the Cabinet decision to appoint John West as an expert
with practical knowledge of irrigation to teach farmers; he was sent
to California and Colorado for several months to study irrigation
methods. On return to Victoria he gave a number of lectures on
irrigation at towns in the Goulburn Valley.\fn{\textit{Goulburn Valley
Yeoman}, 2 May 1890.}

The extent of irrigation at Ardmona was reported as about 300\,acres in
1892; it increased gradually later and involved 4000\,acres by
1916.\fn{\cite{murray1892}; Dept.\ Agric.\ Vic.\ Tech.\ Bull.\
no.\,3, 1944.}

\subsection*{Eshcol, Vic.}
\index{Eschol, Vic.}

The Eshcol irrigation settlement was promoted by J.\,B.~Campbell
\index{Campbell, J.\,B.} who
in 1894 issued a colourful prospectus indicating that the village of
Esh\-col would include a cannery, creamery, and drying houses, as well
as a library, school, and post office.  The proposed location was to
be west of Ardmona and three miles north of Tatura but there is no
evidence that such an irrigation settlement was ever
achieved.\fn{\cite{bossence1969}.}

\subsection*{Mount Scobie, Vic.}
\index{Mount Scobie, Vic.}

Mount Scobie, a solitary hill south of Kyabram in the Goulburn Valley
was the name given to an estate of 9000\,acres bought for
\pounds90\,000 by James Mirams, \index{Mirams, J.} former politician
turned land agent.  Mirams advertised allotments for sale in 1890 as
the Mt Scobie irrigation colony, with minimum holdings of 10\,acres.
Later in the year, Mirams, a prominent figure in the Victorian land
boom, was convicted of issuing a false balance sheet for a business
and was sentenced to a year's imprisonment.  Though some blocks of
land were apparently sold at Mount Scobie, there is no evidence that
an irrigation colony was realised
there.\fn{\cite[p.\,150]{cannon1972}; \textit{Australasian}, 1 Mar.\
1890; \cite{ingham1974b}.}

\subsection*{Mooroopna Irrigation Company, Vic.}

An article published in 1890 refers to the Mooroopna irrigation colony
three miles north of Ardmona and mentioned the involvement of John
West, Mich\-ael Kavanagh, M.~Cussen, W.~Burnett, and Strachan~\& Co.
This followed an advertisement in the same paper for the Mooroopna
Irrigable Lands Investment Co Ltd. The outcome of this venture is not
clear; it may have become part of the Ardmona
settlement.\fn{\textit{Australasian}, 25 Oct.\ 1890 \& 8 Nov.\ 1890,
p.\,878.}

\section*{Campaspe River Valley}
\index{river!Campaspe}

\subsection*{Echuca Irrigation and Freehold Land Company, Vic.}

Efforts to develop irrigation near the downstream course of the
Campaspe River, west of Echuca, began in 1888--89 when a group of
investors arranged to buy three pastoral properties\,---\,Restdown,
Mara\-thon, and Wharparilla\,---\,with a combined area of almost
40\,000\,acres.\fn{VicPP no.\,21 of 1896, RC~Water Supply, MoE,
p.\,436.}

James Shackell, \index{Shackell, J.} prominent in Echuca as an
auctioneer and politician, took a leading part in the scheme to
irrigate that land.  He had been a member of the 1884 royal commission
on water supply and knew of the possibility of bringing irrigation
water to the area as indicated by Gordon and Black in their second
report on irrigation.  The Millewa Irrigation Trust
\index{irrigation!trust!Millewa} was formed in 1890 to undertake
irrigation on approximately 31\,000\,acres of the land purchased by
the investment company.  When the affairs of the trust were examined
by the 1896 royal commission on water supply, Shackell stated that the
scheme, involving the formation of an irrigation colony, relied on the
expectation of getting gravity water `from the Goulburn channel that
was to be constructed at the expense of the government and deemed to
be national works'.\fn{G.~Gordon \& A.~Black, Second Report on
Irrigation, 1884; VicPP no.\,21 RC~Water Supply, MoE p.\,436.}

The scheme was in jeopardy, however, when the purchase of land by the
investment company could not be completed and Shackell went to London
to arrange alternative finance.  He then promoted the Echuca
Irrigation and Freehold Land Company Ltd to complete purchase of the
land for irrigation with subdivision of the three properties into
blocks varying in size from five to 25\,acres.  Shackell's rescue
operation failed and in 1892 the land reverted to the previous owners
and the Millewa trust ceased operations without ever having undertaken
irrigation.\fn{\cite[pp.\,209--210]{coulson1995}.}

The Millewa trust was one of a few Victorian irrigation trusts formed
not by a group of landholders but with only one owner\,---\,a syndicate
or private company.  Those trusts were known as private irrigation
trusts and were later considered to be promoted for speculative
purposes.

\section*{Wimmera River Valley}
\index{river!Wimmera}

\subsection*{Young Brothers Irrigation Colony, Vic.}
\index{irrigation!colony!Young Bros}

The first irrigation colony in the Wimmera district was at Horsham,
where the mean annual rainfall is 17.5\,inches and summers are dry and
hot.  Irrigation was promoted mainly by Thomas and George Young,
\index{Young, T. \& G.} stock
and station agents, who were involved with irrigation at two places
near Horsham about 1890, firstly at Dooen, \index{Dooen, Vic.}  five
miles NE of Horsham, and later within two miles of Horsham.  Their
enterprise at Dooen has been referred to as an irrigation colony
although in 1892 Stuart Murray reported only two holdings
there\,---\,one for Young Bros and another for Mr~Stephens.  The Young
brothers were irrigating 10\,acres at Dooen in 1885 with water drawn
from a supply reservoir fed from the Wimmera River.  The Young
Brothers Irrigation Colony Ltd was in existence on the north side of
the Wimmera River in 1891 and in 1892 the colony involved 260\,acres
all with irrigation channels.  At that time 160\,acres had been sold
and all but 60\,acres of the area had been planted with fruit trees
and vines.\fn{\cite{murray1892}; VicPP no.\,53 of 1885, RC~Water
Supply, Further Progress Rept MoE, p.\,124.}

\subsection*{Horsham Irrigation Colonising Company, Burnlea, Vic.}
\index{irrigation!colony!Horsham, Co.}

The Horsham Irrigation Colonising Company began the Burnlea colony in
1890 on the south side of the Wimmera River at about one mile from
Horsham.  The company was formed by Thomas Young, James Brake,
Frederick Hagelthorn, and two engineers John D.~Derry and Percy
Learmonth.  By 1892 its extent was 550\,acres, all with access to
irrigation channels, and about 100\,acres had been sold with 50\,acres
planted to fruit trees.\fn{\cite[p.\,35]{blake1973};
\cite{murray1892}}

\subsection*{American Colonisation Company, Riverside, Vic.}
\index{irrigation!colony!American, Co.}

In 1892 the American Colonisation Company promoted Riverside, another
colony on the south side of the river, about two miles from Horsham.
Its 430\,acres were provided with irrigation channels, all allotments
were sold, and 150\,acres had been planted with vines and fruit trees
before the end of 1892.\fn{\cite{murray1892}}

\subsection*{Quantong Irrigation Colony, Vic.}
\index{irrigation!colony!Quantong}

An irrigation colony west of Horsham near the Wimmera River was first
proposed in 1890 by Brake and Derry when they bought 2253\,acres.  In
1891 the colony was being promoted by the Cooperative Irrigation and
Mercantile Society of Australia Ltd \index{irrigation!company!Coop.\
Irrig.\ \& Merc.\ Soc.\ Aust.\ Ltd} which advertised more than 80
horticultural allotments varying in size from 5 to 20\,acres and a
large number of township allotments.  The promoting company had
Alexander Baird \index{Baird, A.} as its secretary in Melbourne and
many Quantong settlers had dealings with this successful business man
whose interests later included mining, brewing, and cement.  By about
1895 there were about 60 settlers in the area.  Although the scheme
was promoted as an irrigation colony, there was no supply of water
until 1905.  The earliest products of the colony were dried fruit,
including raisins, currants, and sultanas; fruit trees came into
bearing before 1920.\fn{\cite[pp.\,35, 98]{blake1973};
\cite{horsham1991};
\cite[p.\,28]{gibbney1987}.}

\subsection*{Mount Arapiles, Vic.}
\index{Mount Arapiles, Vic.}

Near Mt Arapiles another colony was promoted in the early 90s; it was
reported in 1892 as involving 640\,acres, with 200\,acres sold in
blocks of 10\,acres or more.  It was expected to be irrigated by water
from the Grampians supplied by the channel
system.\fn{\cite{murray1892}}

\subsection*{Haven, Vic.}
\index{Haven, Vic.}

Haven was the name given to the last colony established near the
Wimmera River.  It was not begun until 1894 apparently as an outcome
of the village settlement south of Horsham associated with Rev.~Horace
Finn Tucker.

Some of these colonies in the Wimmera district were apparently
receiving irrigation in 1892.  By 1893 a few hundred acres were
recorded as receiving irrigation in the Western Wimmera Irrigation
Trust area. \index{irrigation!trust!West Wimmera} By 1907 the
irrigated area in the district did not exceed a thousand acres and
mainly concerned vineyards and orchards. Quantong became the largest
irrigation enterprise in the district.  Irrigation of these colonies
generally depended on supplies from storages dependent on runoff from
the nearby Grampians mountains, with gravity supply of water by
streams and artificial channels developed by the local trust
responsible for water supply. John Derry, \index{Derry, J.} an
engineer with experience of irrigation in India, came to the Wimmera
in the early 80s, visited USA with Alfred Deakin for the Royal
Commission on Water Supply, and was closely involved with the Burnlea
and Quantong colonies.\fn{\cite{murray1892}; VicPP no.\,20 of 1896,
RC~Water Supply, Appendix~C; VicPP no.\,38 of 1907, Annual Rept SRWSC,
p.\,8.}

\section*{Werribee River Valley}
\index{river!Werribee}

\subsection*{Werribee Irrigation Colony, Vic.}
\index{irrigation!colony!Werribee} \index{Werribee, Vic.}

The plain traversed by the Werribee River on its way to Port Phillip
Bay remained for many years under the control of pastoralists,
especially the Chirnside family which in 1875 held 85\,000\,acres
freehold and grazed 80\,000 sheep.  Agriculture involved less than 5
per cent of the area although rainfall, on the average 20\,inches in
the year, was sufficient to produce good crops of cereal hay and
grass.  In 1880 less than 100\,acres were devoted to vegetables,
orchards, and gardens.\fn{\cite[pp.\,162--63]{peel1974}.}

The Werribee River, a relatively small stream by Victorian standards,
occupies a deep valley and changed seasonally from a torrent to a
series of pools until dams were constructed in its upper reaches.  Its
tidal reach was used for some years by paddle-steamers
\index{paddle-steamer} travelling five
miles from the Bay.

When public interest in irrigation was stimulated by Deakin's visit to
America, enactment of his irrigation legislation, and the arrival of
the Chaffey brothers to promote irrigation, the Werribee people were
not backward in responding: they held a public meeting in May 1886 to
register interest in local irrigation.  The provisions of legislation
that year for the development of irrigation by irrigation trusts
supported by loans from the government led to the formation of the
Werribee Irrigation Trust in November 1888, a time of low rainfall at
Werribee where only 14\,inches were recorded that year.  The trust
concerned an area of about 1450\,acres acquired from a local resident
by the Werribee Irrigation and Investment Co
\index{irrigation!company!Werribee Irrig.\
\& Invest.} which initiated formation
of the trust.  This company was launched in October 1888 and a few
months later issued 100\,000 shares of \pounds1 each.  George Chaffey
\index{Chaffey, G.} 
became its chairman of directors and E.\,C.~De Garis \index{De Garis,
E.\,C.} was the managing director.  These men were also respectively
the chairman and secretary of the irrigation
trust.\fn{\textit{Leader}, 29 May 1886; VicPP no.\,20 of 1896,
RC~Water Supply, Rept pp.\,179--180, Evidence pp.\,425--433;
\cite{ronald1978}; \cite[p.\,192]{alexander1928}.}

The company intended that the property would be subdivided into small
holdings provided with water stored on the river by a weir to be
erected a few miles upstream from the irrigation colony.  A temporary
pumping plant was being installed on the river to allow a considerable
portion of the land to be irrigated before completion of the weir.
The site involved was on the east side of the river, within a few
miles of the Werribee township.\fn{\textit{Mildura Cultivator}, 14
Feb.\ 1889.}

Subdivision of the area went ahead, the pumping plant was installed on
the river, and an irrigation channel was made.  Boundaries of
allotments were planted with trees and advertisements enticing
settlers with the slogan `Fortunes in Vegetables' were made late in
1890 with advice that payment for allotments could be spread over 15
years.\fn{\textit{Australasian}, 25 Oct.\ 1890, p.\,804.}

The irrigation company wished to install a pipeline in order to
irrigate more land; this depended on agreement with Percy Chirnside,
\index{Chirnside, P.}
the owner of land to be crossed by the pipeline.  In order to
propitiate him, the Company proposed calling their irrigation colony
`Chirnside', but the stumbling block of compensation persisted and
delayed further action.  Whether this pipeline was to bring water from
the proposed weir or for delivery of supplies from the temporary
pumping plant is not clear, but there is no record of construction of
the weir or installation of the pipeline.\fn{\cite{ronald1978}}

Within two years of formation, some progress had been made with the
irrigation colony.  The Melbourne Nursery and Poultry Farm Company
\index{Melbourne Nursery \& Poultry Farm Co.} had
purchased 96\,acres of land for \pounds7500, the company balance sheet
showed a profit, and buildings were going up on the land.  One was the
expensive residence for George Chaffey, reputed to have cost
\pounds3000 and known as Manchester Park.  It was later let to 
Percy Chirnside while his mansion, Manor Park, was being built just
across the river.  Manchester Park burnt to the ground in
1924.\fn{\cite[p.\,172]{ronald1978}.}

In 1891 the Trust was responsible for irrigating no more than
200\,acres and when the company went into liquidation that year, its
responsibility was assumed by George Chaffey who also failed to carry
on the undertakings and the land reverted to its previous owner,
Thomas Agar, \index{Agar, T.} who was elected by the remaining
settlers as chairman of the trust.  De Garis, who had lived in the
district for a few years while active in the irrigation development,
moved to Mildura late in 1891, while George Chaffey and his wife made
the same change in 1892.  A few settlers continued with irrigation of
vegetables and fruit trees.\fn{VicPP no.\,20 of 1896, RC~Water Supply,
Rept pp.\,179--80, Evidence pp.\,425--433.}

The Werribee trust was one of three private irrigation trusts which
had been assisted by the government with loans not repaid by 1896 when
the report by the royal commission found that:
\begin{Quote}
	the promoters of these trusts had in each case conditionally
	purchased an area of lands which they desired to form into an
	irrigation trust, not for the purpose of working the land
	themselves, but with the object of cutting it up and selling
	it to others.\fn{VicPP no.\,20 of 1896, RC~Water Supply,
	Rept pp.\,176--80.}
\end{Quote}
After this finding, the commissioners expressed the opinion that `the
State money had been advanced to irrigation trusts in a most lavish
and prodigal manner, and without careful investigation'. The
commission also found that in December 1889 George Chaffey had
obtained an advance of \pounds3000 from the government without
security and that the Minister, Alfred Deakin, was at fault in making
the loan.\fn{VicPP no.\,20 of 1896, Rept pp.\,185, 190.}

Deakin's speech in 1896 in response to the report of the Commission,
which touched on so much of his responsibility for water supply in the
period 1886--90, was notable for its length and the fact that he was
obliged to acknowledge his personal responsibility for the loan to
George Chaffey.\fn{VicPD, 18 Dec.\ 1896 vol.\,84, p.\,4738.}

\section*{Nepean River Valley}
\index{river!Nepean}

\subsection*{Mulgoa Irrigation Company, NSW}
\index{irrigation!scheme!Mulgoa}

Irrigation of fruit trees and vines on an area of more than
10\,000\,acres lying 30 miles west of Sydney, extending southwest from
St Marys to the Nepean River was promoted by George Chaffey
\index{Chaffey, G.} and Henry
Gorman \index{Groman, H.} in the late 1880s.  At that time George
Chaffey had experience of developing irrigation settlements in areas
with very low rainfall at Mildura and Renmark, and at Werribee with
moderate rainfall.  The Mulgoa scheme was unusual because its area had
an average rainfall of 27\,inches.

The scheme was probably initiated in 1888, judging by a published
reference then to proposed irrigation near Penrith involving the
Chaffeys.  Promotion in 1891 apparently involved several men including
E.\,C.~De Garis in addition to the Chaffey brothers.  The Mulgoa
Irrigation Bill came before the New South Wales parliament in 1890 and
was considered by a select committee.  Several witnesses appeared
before that committee including the principal promoters, G.~Chaffey
and Henry Gorman, and H.\,G.~McKinney, \index{McKinney, H.\,G.} chief
engineer for water conservation.  The Bill was passed late in
1890.\fn{\textit{Mildura Cultivator}, 20 Dec.\ 1888;
\cite{alexander1928}; NSW LC Journal vol.\,47, pt.\,2 1890, Rept
Select Comm.\ on the Mulgoa Irrigation Bill.}

The Mulgoa scheme involved (i) purchase of 16\,000\,acres from several
more or less contiguous properties in the Penrith district and
division into allotments of three types: town, suburban villas
(two--five acres), and horticultural (10--20\,acres); (ii) supply of
water for irrigation by lifting water 130\,feet by pumping from the
Nepean River to a reservoir, reticulation by gravity via earthen
channels, and a subsidiary lift of a further 50\,feet to water the
higher parts of the area; and (iii) provision of weirs on the Nepean
and/or its tributary the Warragamba, \index{river!Warragamba} to
ensure adequate supplies.

Two men responsible for the engineering design were George Chaffey and
his associate J.\,G.~Starr.\index{Starr, J.\,G.}  The scheme was
publicised early in 1891 by articles in several Sydney newspapers and
the Mulgoa Irrigation Co Ltd was subsequently formed, with
A.\,W.~Stephen \index{Stephen, A.\,W.} as managing director.  The
company then invited applications for purchases of the different types
of holdings and stated the arrangements for water supply, giving each
settler `a perpetual claim upon the company for water on the specified
terms' without the establishment of a water company.\fn{\textit{Sydney
Morning Herald}, 3 Jan.\ 1891;
\textit{Mulgoa Irrigation Settlement}, brochure 1892.}

The company apparently made little progress with the irrigation
sch\-eme, and the government was invited to take over the scheme in
1896 when legislation on water rights gave the opportunity for greater
state involvement in irrigation.  This proposition was considered by
Colonel Home, \index{Home, Col.} the irrigation engineer from India
invited to report to the government on irrigation prospects in the
colony.  His report showed that the Mulgoa scheme would be `far from
remunerative'.  There was no subsequent development of the Mulgoa
irrigation scheme.\fn{NSW V\&P 1897, vol.\,5, Rept on Prospects of
Irrigation and Water Conservation in New South Wales, by Colonel
F.\,J.~Home.}

\section*{Hunter River Valley}
\index{river!Hunter}

\subsection*{Segenhoe Estate, NSW}
\index{Segenhoe estate, NSW}

Segenhoe was the name of an agricultural settlement which flourished
in the Hunter valley before 1830.  Prospective irrigation of the
Segenhoe estate, which involved 25\,000\,acres, was reported on by a
select committee of the New South Wales Legislative Assembly in 1891
when it considered the Segenhoe Estate Irrigation Bill then before
parliament.  William Harris told the committee that he was managing
director of the Land Company of Australasia Ltd, which owned the
property and proposed raising funds from abroad to impound at least
800 million gallons of flood waters of the Hunter River system
upstream from Aberdeen and Muswellbrook.  F.\.B.~Gipps \index{Gipps,
F.\,B.} was the civil engineer responsible for the scheme which
involved gravitation of supplies to pastoralists acquiring portions of
the property.  Other witnesses included local pastoralists and
townspeople from Muswellbrook which might incur some disadvantage to
water supply.

After the select committee reported to the legislative assembly the
proposed legislation apparently lapsed.\fn{NSW V\&P LA 1891--92
vol.\,7, pp.\,713--742, Rept Select Comm.\ on the Segenhoe Estate
Irrigation Bill.}

\section*{Darling River}
\index{river!Darling}

\subsection*{Menindie Irrigation Scheme, NSW}
\index{irrigation!scheme!Medindie}

An area of 25\,000\,acres extending northwards from Lake Menindie on
the Darling River was proposed for irrigation in 1892 by the Menindie
Irrigation Settlement Co Ltd with capital of \pounds50\,000.  The
scheme was considered by a select committee of New South Walews
parliament in 1893 when it was called on to report on the Menindie
Irrigation Bill.  Witnesses who gave evidence to the committee
included H.\,G.~McKinney, chief engineer for water conservation,
C.\,W.~Darley, engineer in chief for harbours and rivers, C.\,E.~Hogg,
engineer to the company involved, and J.\,M.~Purves, a land agent from
Sydney who represented the company.  This scheme was associated with
the projected Menindie and Broken Hill tramway, also the subject of a
report by a parliamentary select committee.  Neither of these projects
was realised.\fn{NSW V\&P LA 1892--93, vol.\,8, Rept Select Comm.\ on
the Menindie Irrigation Bill.}

\section*{Gawler River}
\index{river!Gawler}

\subsection*{Gawler Plains Irrigation Colony, SA}
\index{irrigation!scheme!Gawler Plains}

A scheme to provide irrigation on the Gawler plains north of Adelaide
was considered in 1889 by the Barossa Water Commission as one option
associated with construction of a reservoir on a tributary of the
Gawler River intended to augment supplies to the Adelaide metropolis.
Apparently inspired by the development of the Chaffey irrigation
colonies on the Murray River, the establishment of an irrigation
colony involving 12\,000\,acres near Smithfield was supported by several
prominent agriculturists but was viewed unfavourably by W.~Culcheth, a
Victorian irrigation engineer, and by the Commission.\fn{SAPP no.\,25
of 1889, Rept of the Barossa Water Commission.}

The scheme was revived later by two men: Packard, a surveyor, and
Lutz, an engineer.  In 1894 they were authorised by the South
Australian government to proceed with their plan for irrigation on the
Gawler plains using water from a reservoir to be constructed on the
South Para River. \index{river!South Para} Their project, referred to
in an official report as the Gawler Plains Irrigation Colony was
expected to involve subdivision of 10\,000\,acres between Smithfield
\index{Smithfield, SA}
and Salisbury \index{Salisbury, SA} into blocks of five to 10\,acres
for `intense irrigated culture of vineyards, orchards, orangeries,
etc', and into larger blocks for growth of fodder for dairying and
grazing purposes.  Water was to be provided from a storage reservoir
on the South Para River, with an embankment near Williamstown.\fn{SAPP
no.\,111 of 1896, The Proposed Gawler Plains Irrigation Scheme.}

As with other promoters of irrigation colonies in Australia during the
1890s, Packard and Lutz failed to secure adequate financial backing
from their associates in England.  However, they prevailed on the
South Australian government to give limited financial help, by
guaranteeing a small amount of money, so that the work could proceed.
This help was authorized by the Gawler Plains Irrigation Act of 1896
but it failed to bring the project any nearer to realisation.
Eventually the Barossa Reservoir \index{reservoir!Barossa} was started
in 1898 by the government to provide water supply to Gawler and
settlements to the north.

\closure
Many of the schemes were never tested in practice after failing to
attract sufficient support at the outset. This lack of support
indicates their difficulty in competing with many other speculative
ventures proposed during the land boom which affected several
Australian colonies. Except for the Tresco settlement which started
much later, the irrigation colonies which came into existence were
relatively small ventures in the Wimmera and Goulburn Valley of
Victoria.

%\section*{References}
%1. F.Bevilaqua, General Remarks And Detail Accounts Of The South Australian
%    Irrigation Company, 1884.
%2. SAPP No.158 of 1884 \& SAPP No.73 of 1885.
%3. Adelaide Observer 22/9/1888 \& 29/9/1888.
%4. SAPD 1888, P.1267, 9/10/1888.
%5. M.Cannon, The Land Boomers, 1986.
%6. A.J.Stimson, ADB Vol.9, p.655.
%7. Southern Argus, 5/5/1892.
%8. Southern Argus, 23/6/1892.
%9. Southern Argus, 20/10/1892.
%10. Southern Argus, 2/2/1893, \& J.Faull, Alexandrina's Shore, A
%      History Of The Milang district, 1981.
%11. SAPP No.29 of 1890, A.Deakin, Irrigation In Australia, 1892, p.92.
%12. SAPP No.29 of 1892, L.J.Blake, ADB vol.4,pp.58-59.
%13. S.Wells, Paddle Steamers to Cornucopia, 1986, p.77.
%14. G.Gordon papers, File H17329, La TL.
%15. Swan Hill Guardian 19/11/1890.
%16. G.Gordon papers, File H17329, La TL.
%17. N.Barr \& J.Cary, Greening A Brown Land, 1992, p.262.
%18. N.Barr \& J.Cary, 1992. p.328.
%19. F.M.Read, J.Agric.Vict. 1931 vol.29, p.551.
%20. F.M.Read, J.Agric.Vict. 1930, vol.28,p.65.
%21. F.M.Read, 1931, p.563.
%22. A.Trengrove, John Grey Gorton, 1969.
%23. VicPP No.20 \& 21 of 1896, R.C.Water Supply, Rept.  pp.176-180 \& 
%      MoE p.438.
%24. VicPP No 20 of 1896, Rept.  pp. 176-180 \& Appendix C.
%25. VicPP No.21 of 1896, MoE p.438
%26. W.H.Bossence, Murchison, 1965, pp.105-106.
%27. Aqua, Nov. 1968, p.53.
%28. Margaret Steven, ADB vol.12, 1990, p.445.
%29. C.S.Martin, Irrigation And Closer Settlement In The Shepparton
%    District 1836--1905, 1955, pp. 57-58.
%30. Australasian, 18/2/1888, p.354.
%31. Goulburn Valley Yeoman, 2/5/1890.
%32. S.Murray, Irrigation In Victoria, 1892, \&
%    Dept.Agric.Vict. Tech. Bull. No.3,1944.
%33. W.H.Bossence, Tatura And The Shire Of Rodney, 1969.
%34. M.Cannon, The Land Boomers, 1986, p.150.
%35. Australasian, 1/3/1890.
%36. S.M.Ingham, ADB, vol.5, 1974, p.258.
%37. Australasian, 8/11/1890, p.878.
%38. Australasian, 25/10/1890.
%39. VicPP No.21 of 1896, R.C.Water Supply, MoE, p.436.
%40. G.Gordon \& A.Black, Second report on irrigation, 1884.
%41. VicPP No.21 R.C.Water Supply, MoE p.436.
%42. Helen Coulson, Echuca -Moama On The Murray, 1995, pp.209-210.
%43. S.Murray, Irrigation In Victoria, 1892.
%44. VicPP No.53 of 1885, R.C.Water Supply, Further Progress Rept.
%      MoE, p.124.
%45. S.Murray, Irrigation In Victoria, 1892.
%46. L.J.Blake, Wimmera, 1973, p.35.
%47. S.Murray, Irrigation In Victoria, 1892.
%48. S.Murray, Irrigation In Victoria, 1892.
%49. L.J.Blake, Wimmera, 1973, p.35.
%50. Horsham \& District Historical Soc., Memories of Quantong, 1991.
%51. H.J.Gibbney \& Ann G.Smith, 1987, A Biographical Register
%      1788-1939, vol. 1 p.28.
%52. L.J.Blake \& K.H.Lovett, Wimmera Shire Centenary, 1962, p.98.
%53. Horsham \& District Historical Soc., Memories of Quantong 1991.
%54. S.Murray, Irrigation In Victoria, 1892.
%55. S.Murray, 1892.
%56. VicPP No.20 of 1896, R.C. Water Supply,Appendix C.
%57. VicPP No 38 of 1907, Annual Rept SRWSC,p.8.
%58. Lynette J.Peel, Rural Industry In The Port Phillip Region
%      1835-1880, 1974, pp. 162-163.
%59. Leader, 29/5/1886.
%60. VicPP No.20 of 1896, R.C.Water Supply, Rept.pp.179-180.
%61. Heather B.Ronald, Wool Past The Winning Post, 1978.
%62. J.A.Alexander, The Life Of George Chaffey, 1928, p.192.
%63. VicPP No.20 of 1896, R.C.Water Supply, Evidence pp. 425-433.
%64. Mildura Cultivator, 14/2/1889.
%65. Australasian, 25/10/1890, p.804.
%66. Heather B. Ronald, 1978.
%67. Heather B.Ronald, 1978, p.172.
%68. VicPP No.20 of 1896, R.C.Water Supply, Rept., pp.179-80, \& 
%      Evidence pp. 425-433.
%69. VicPP No.20 of 1896, R.C.Water Supply, Rept. pp.176-80.
%70. VicPP No. 20 of 1896, Rept. p.185.
%71. VicPP No 20 of 1896, Rept. p.190.
%72. VicPD, 18/12/1896 vol.84, p.4738.
%73. Mildura Cultivator, 20/12/1888.
%74. J.A.Alexander, The Life Of George Chaffey, 1928.
%75. NSW LC Journal Vol.47, Pt.2 1890,Rept Select Comm.on the Mulgoa
%      Irrigation Bill.
%76. Sydney Morning Herald, 3/1/1891.
%77. Mulgoa Irrigation Settlement, brochure 1892.
%78. NSW V\&P 1897, vol.5, Rept on Prospects of Irrigation and Water
%      Conservation in New South Wales, by Colonel F.J.Home.
%79. NSW V\&P LA 1891-92 vol. 7,pp.713-742, Rept. Select Comm.on the
%      Segenhoe Estate Irrigation Bill.
%80. NSW V\&P LA 1892-93, vol.8, Rept Select Comm. on the Menindie
%      Irrigation Bill.
%81. SAPP No.25 of 1889, Rept of the Barossa Water Commission.
%82. SAPP No.111 of 1896, The Proposed Gawler Plains Irrigation Scheme.

