% $Id$
% CHAPTER FOURTEEN
% 9056 words at 3/5/99

\setcounter{endnote}{0}

\chapter{State Intervention in Irrigation 1900--1920}
\label{ch:stateint}\addtoendnotes{\protect\subsection*{Chapter \thechapter}}
%\markboth{Chapter \thechapter. State Intervention}%
%{Pioneering Irrigation in Australia}
\index{state intervention}

\fancyhead[RE]{\sffamily \small Chapter \thechapter.\ %
               State Intervention}

Despite the success obtained by many who independently took up
irrigation and the enthusiasm displayed by those who regarded its
development as the much-needed high technology for the countryside,
little progress had been made by the end of the century.  On the
mainland Victoria and New South Wales had almost extinguished riparian
rights, an important prelude to government control of irrigation, but
efforts to foster the practice by establishment of irrigation trusts
or help for speculative promoters of irrigation colonies had little
success.

Direct government involvement with irrigation in Victoria and New
South Wales followed the widespread drought of 1902 and was followed
later in South Australia and Western Australia.  Legislation was
passed in Queensland to provide for state control of water resources
but no government development of irrigation occurred before 1920.  In
Tasmania riparian rights remained in force and governments refused to
become actively involved in irrigation.

\section*{Victoria}
\index{Victoria}

As related in a previous chapter, a succession of governments failed
to solve problems of the numerous irrigation trusts which had nominal
control of more than 700\,000\,acres, mostly suitable for irrigation, but
never managed to water more than 120\,000\,acres in a year.  Irrigation
trusts generally proved unable to meet their debts by collection of
rates from landholders, many being disinclined to adopt
irrigation.\fn{\citet[p.\,121]{powell1989}.}

The Water Act of 1905 provided for establishment of a new
authority\,---\,the State Rivers and Water Supply Commission
(SRWSC)\,---\,to control \index{SRWSC} surface water resources of
Victoria other than those of the Melbourne metropolis and use them to
provide water for domestic use and livestock and for irrigation.  The
Commission was controlled by three men appointed by the government;
they had power to construct headworks, levy rates, and make charges
for delivery of water.  Though expected to operate like a commercial
business and to be free from political influence, the Commission was
not expected to make a profit; it depended on allocation of funds by
government for capital works and on the rates paid by irrigators.  As
far as irrigation was concerned, the 1905 legislation had some
important features.

State control of water was extended to embrace all streams.  This was
an advance on the Irrigation Act of 1886 which left intact those
riparian rights established previously and thus limited the power of
the government to make reservoirs in the head waters of any Victorian
stream, a matter referred to earlier by Stuart Murray. The extent of
riparian rights had been limited by reservations made in 1881 of all
frontages to watercourses adjoining crown land, but those rights
apparently remained in substantial parts of Victoria previously
alienated from the crown.  When the 1905 Water Bill was debated strong
opposition to its blanket provisions regarding control of stream
waters came from one parliamentarian from the Western District, where
many pastoralists had long-established riparian
rights. \index{riparian rights} One river apparently then subject to
those rights along parts of its course was the Glenelg
\index{river!Glenelg} flowing from
the Grampians mountains in western Victoria to the coast.  The
diversion of its headwaters northwards to improve supply for
irrigation and water supply in the Wimmera was considered before 1892
but not undertaken then, possibly because of riparian rights and
expense.  Existing riparian rights were reduced by the 1905 Act to
allow landholders involved to make only limited use of stream water,
generally for stock and domestic use.\fn{S.~Murray (1889), quoted by
L.\,R.~East in Aqua
\textbf{9}, 1958, p.\,248; \citet[fig.\,35, p.\,148, p.\,150
re W.\,S.~Manifold]{powell1989}; VicPD vol.\,110 30 Aug.\ 1905,
pp.\,1247--8; L.\,R.~East (1967), Vic.\ Hist.\ Mag.\ p.\,182;
A.~Deakin (1892), Yearbook of Australia p.\,9.}

After establishment of the SRWSC the maintenance of irrigation was
continued temporarily by the irrigation trusts.  In 1906 the area
irrigated in Victoria was reported by the Commission as approximately
104\,000\,acres including the Mildura settlement. \index{Mildura,
Vic.} Storage capacity, required for irrigation and for stock and
domestic supply, was then about 370\,000\,acre-feet including
200\,000\,acre-feet recently added on completion of the initial design
of Waranga Basin.  Additional supplies were obtained from the Murray
River by pumping and many permits and licences were issued by the
Commission for private diversion from different streams.  Between 1910
and 1915, the irrigation annually dependent on these diversions ranged
from 14\,000\,acres to 24\,000\,acres.\fn{\citet[p.\,54]{mccoy1988};
VicPP, Ann.\ Repts SRWSC 1906--7, 1914--15, also E.~Mead (1914), BAAS
Hdbk Vic., p.\,266.}

The provision of stock and domestic supplies by the SRWSC became a
major undertaking as an extensive system of earthen channels was
developed by 1914 to serve wheat-growing farms in northwestern
Victoria, an area largely devoid of useful streams. That system was
for many years largely dependent on supplies from reservoirs filled
from the Grampians, with the consequence that irrigation could not
then be extended further in the nearby Wimmera
district.\fn{L.\,R.~East (1967), Vic.\ Hist.\ Mag.\ \textbf{3}8,
p.\,205; \citet{mcnab1944}.}

\subsection*{Elwood Mead's Innovations}
\index{Mead, E.}

Continuity with the previous management of irrigation was ensured by
the appointment of Stuart Murray \index{Murray, S.} as first chairman
of SRWSC.  On his retirement in 1907, the government appointed Elwood
Mead to the position.  Mead, an irrigation engineer and professor of
irrigation in California, had spent ten years with the US Department
of Agriculture in charge of irrigation investigations.  In Victoria he
sought increased use of irrigation by a new rating system and
establishing new irrigation settlements.

The SRWSC had initially fixed rates for water supply on the basis of
land values but in 1909 the Water Act was amended to provide a scale
of water-rights according to use of the irrigable land, and a
proportionate charge for water to be met by landholders whether or not
it was used for irrigation.  For each irrigation district the rate of
payment would be fixed to cover the maintenance and management of the
works supplying irrigation water, the interest on the cost of such
works and a redemption fund on the cost of such works.  No charges
were to be made in respect to what were designated as `free
headworks'\,---\,their identification in the relevant statute shows that
they included the national works provided by the State before 1905.
This system of charging irrigators remained in force for many
years.\fn{Vic.\ Water Act 1905, 1909, and 1915.}

The new settlements sought by Mead were established on properties
hitherto unirrigated. They were to be subdivided into holdings ranging
in area from 20 to 200\,acres depending on soils, access to markets, and
nature of intended crops.  They would be provided with channels and
works necessary for irrigation of the new farms.  The settlers would
preferably be residents of Victoria but if their response was not
adequate steps should be taken to attract settlers from Europe and
America.\fn{Ann.\ Rept SRWSC 1908; \citet[J.~Rutherford,
p.\,126]{powell1974}.}

Mead's proposals came at a time when Victorian governments were trying
to redress the loss of almost 10 per cent of the population during the
depression of the 1890s, mainly to Western Australia, by attracting
migrants to country districts.  A Closer Settlement Board had been set
up a few years earlier to establish new farms, generally by
subdivision of sizable holdings purchased by the government.  During a
period of about ten years the Board acquired more than 500\,000\,acres
of land it regarded as suitable for farming; the SRWSC became
responsible for developing irrigated agriculture on no more than
100\,000\,acres.  The first involvement of the SRWSC in this activity
related to the establishment of new irrigation settlements at Merbein
and Nyah, \index{irrigation!settlement!Merebin} both to be supplied
with water by pumping from the nearby River Murray.\fn{S.~Whitehead
(1914), BAAS Hdbk Vic.\
p.\,95.}
\index{irrigation!settlement!Merebin}\index{irrigation!settlement!Nyah}

The Merbein settlement, originally known as White Cliffs, was
established on crown land covered by mallee scrub and woodland
alongside the northwest boundary of the concession obtained by Chaffey
Brothers for irrigation development at Mildura.  Plans for subdivision
were made in 1908 and next year the SRWSC offered 150 blocks for
selection in an area of 5000\,acres.  The Commission intended dairying
to be the main industry, based on lucerne irrigated with water lifted
about 90\,feet from the River Murray.  By 1914 there were 182 blocks
occupied by settlers; the average block size was 32\,acres.
Horticultural crops were soon established and with the increased price
for dried fruit during the 1914--18 war this commodity became the main
product of the settlement.  Extensions to the irrigated area began in
1917 for settlement of war veterans, including one portion known as
Birdwoodton which occupied part of the concession area granted to the
Chaffey Brothers in the 1880s at Mildura.\fn{E.\,Mead (1914), BAAS
Hdbk Vic.\ p.\,265; \citet[p.\,265]{wells1986}; \citet{mcintyre1948};
CSIR Bull.\ np.\,123, 1939, Merbein Rept.}

The Nyah settlement compared with Merbein in respect to topography,
vegetation and supply of irrigation water by pumping from the Murray.
It was developed by the Commission in 1909 on an area of 3000\,acres,
which was subdivided into 100 horticultural holdings with an average
size of 29\,acres but many blocks were unoccupied in 1914. An
additional 45 blocks became available to settlers in 1919 in the areas
known as Vinifera and North Nyah Extension.  The main produce of the
settlement was dried fruit.  Irrigation depended on pumping from the
river to a height of 80\,feet.\fn{E.~Mead (1914); CSIR Bull.\ no.\,73,
1933.}

Experiments significant for Australian irrigation of rice though
apparently started without provision of irrigation by the SRWSC were
conducted for several years by the Japanese immigrant
J{\={o}}~Takasuka \index{Takasuka, J.} near the Murray river between
Swan Hill and Nyah.  In 1908, he obtained occupation of 200\,acres of
land subject to flooding and ultimately succeeded in 1911 in raising
rice from seed sown directly in the ground.  Seed that he had raised
was apparently used by another grower at Koyuga, where the crop sown
in November 1915 and watered weekly from a special dam filled from
irrigation channels gave a marketable yield of grain impressive enough
for others in the district to start sowing
rice.\fn{\citet{sissons1990};
\citet[p.\,117]{scholes1989}; The Australian Encyclopedia, 1927,
article on rice; T.~Smith (1916), J.\,Agric.\ Vic.\ \textbf{14},
p.\,493.}

The major implementation of Mead's plans took place in northern
Victoria, where several holdings with little or no previous use for
irrigation were purchased for closer settlement and subdivided to
provide for intensive production on irrigated holdings.  Most were to
be watered by gravitation from the Goulburn River via the Waranga
Basin and the Western Waranga Main Channel, then being extended
towards the Loddon Valley which it reached at Serpentine Creek
\index{creek!Serpentine} in 1912.  The new irrigation settlements stood
like islands in a region largely used for cereal growing and pastures
with supplementary irrigation sought only in dry seasons for a
proportion of a holding.  By 1914 more than 800 holdings varying in
size from about 30\,acres to more than 80\,acres had become available
in northern Victoria by subdivision of estates for irrigation of
orchards and lucerne near Shepparton, Kyabram, Tongala, and
Rochester.\fn{\citet[p.\,16]{mccoy1988}; E.~Mead (1914).}

Other additions to the area under irrigation were achieved at Cohuna,
also in northern Victoria, relying on water drawn from the Murray
along Gunbower Creek; near Swan Hill using water pumped from the
Murray; and at Werribee in central Victoria where an irrigation
settlement occupying 6000\,acres and supplied with water from the
Werribee River was opened in 1912, following construction of a
reservoir at Pykes Creek in 1911.\fn{Aqua Jan.\ 1969, \citet[pp.\,86,
90, 97]{james1985}.}

Some of the new irrigation settlements failed to attract a sufficient
number of applicants by 1910, a situation prompting efforts to attract
new settlers and agricultural labourers from the northern hemisphere.
Mead and Hugh McKenzie, Minister of Lands, visited Europe and Nor\-th
America for six months intending to recruit six thousand immigrants
for the new irrigation farms. Their mission visited Italy, Sweden,
Denmark, the United Kingdom and the United States, with the result
that almost 200 migrant settlers took up blocks before 1912.  Further
efforts were made in 1912 by William Cattanach of the SRWSC to recruit
settlers from the United States and by R.\,B.~Rees, a parliamentarian,
to persuade Welsh settlers in Patagonia to migrate to Victoria.
Irrigation was also taken up by Spaniards from Catalonia at Echuca and
Cohuna and by the colony of Russian Jews near Shepparton, but in 1914
there were still many closer settlement blocks without occupants.

While Mead was prominent in the SRWSC the irrigated area in Victoria
increased, following the creation of new irrigation settlements and
improvements in the areas taken over from the irrigation trusts.
Nevertheless there was criticism of closer settlement by many settlers
and questions were raised about purchases of land in electorates of
government ministers.  Eventually in 1914 the government ordered a
royal commission on closer settlement, leading to separate reports on
dryland and irrigated settlement.  The association of the Closer
Settlement Board and SRWSC ended after the Commission in 1916 found
evidence of some earlier purchases of land unsuitable for irrigation.
One instance related to an area of 11\,000\,acres in the Cohuna
district, of which 1200\,acres in the locality known as Mead had been
so affected by salinity \index{salinity} as to render the land
temporarily unfit for production of any kind.  The problem was
investigated by the Victorian Department of Agriculture in 1913, with
the conclusion that hundreds of acres had been rendered infertile by
the accumulation of chlorides and sulphates mainly in the past two
years following the extension of irrigation.

Drainage \index{drainage} was held to offer the only effectual cure
for the salt problem.  The widespread and persistent drought of 1914
showed that the success of irrigation depended on the reliability of
water supplies, wh\-ich were maintained from the Waranga Reservoir
thoughout much of 1914 but failed early in 1915.  One outcome was the
decision in 1915 to increase its capacity and to build a new dam on
the upper Goulburn River. \index{river!Goulburn} In addition, the
supply of irrigation water from the Murray was to become more assured
following the River Murray waters agreement of 1915 and the ensuing
work of the River Murray Commission which from 1919 included
construction on the Murray of the Hume Dam \index{dam!Hume}
\index{weir!Torrumbarry} and the
Torrumbarry Weir.\fn{\citet[pp.\,131--33]{broome1984};
\citet[pp.\,116--24]{scholes1989}; \citet{coulson1995};
\citet{trevaks1937}; Mead (1914), p.\,265; \textsl{Melbourne Argus}, 20
Mar.\ 1913, p.\,15; VicPP RC~Closer Settlement 1916, Final
Rept; \citet[p.\,99]{keating1992}; \citet[p.97]{forster1965}.}

The first world war and the departure of Elwood Mead \index{Mead, E.}
in 1915 led to some reduction of efforts by the SRWSC until in 1918,
following the Closer Settlement Act, it became responsible for land
purchase and settlement in irrigation areas.  Additions to some
irrigated settlements were then begun to provide for returned
servicemen.  Work also began on a new settlement known as Red Cliffs,
\index{Red Cliffs, Vic.} for which A.\,S.~Kenyon \index{Kenyon, A.\,S.}
 at the end of 1918 proposed the development by the SRWSC of
10\,000\,acres under irrigation near Mildura.  The scheme depended on
purchasing portion of the Chaffey concession land then held by
creditors of the Chaffeys.  This was finalised to allow survey work to
begin there early in 1920 and deliveries of water pumped from the
river began in 1921.  Red Cliffs later included more than 700
allotments held mainly by discharged soldiers for production of dried
fruit.\fn{\citet[p.\,14]{mccoy1988}; A.\,J.~Tisdall (1964), Irrigation
in Victoria, in Australian Academy of Science, Water Resources Use and
Management; J.\,N.~Churchyard (1960), Aqua , \textbf{11}(12); CSIR
Bull. no.\,137.}

Ex-servicemen were also settled on many of the vacant allotments in
closer settlement areas developed by the SRWSC after Mead's
appointment.  In 1914 there were almost 400 such allotments with a
total area of 19\,000\,acres in 10 settlements. The outcome of this
post-war settlement is reported by the Victorian Royal Commission on
Soldier Settlement in 1925.\fn{E.~Mead (1914); VicPP no.\,32 of 1925;
\citet{scholes1989}.}

The SRWSC relied entirely on earthen channels for distribution of
water to irrigators, thus efficiency was lost as a result of seepage
\index{seepage} 
which in some places contributed later to waterlogging
\index{waterlogging} with harmful
effects for irrigated crops.  The system of varying water rights for
different classes of irrigable land required some means of measuring
the use of water by each irrigator; after testing different metering
devices the Commission in 1911 adopted one invented by John Dethridge,
a member of the Commission, but it was not in general use before
1920.\fn{I.~Meacham (1961) Aqua, p.\,184.}

By 1920 the Victorian area under irrigation, including Mildura,
\index{Mildura, Vic.} 
increased to about 370\,000 ac from 104\,000 in 1906--07.  Pastures
and luc\-er\-ne accounted for most of the irrigation by 1920, with
vines, orchards and vegetables covering less than a quarter of the
irrigated area. Considering that the area irrigated by private
licensed diversionsof water accounted for about 20\,000\,acres in 1914
and 17\,000\,acres in 1917--18 while Mildura accounted for a smaller
area, the Victorian achievement of irrigation was due mainly to
government intervention.\fn{\citet[pp.\,54--55]{mccoy1988}; Ann.\ Rept
SRWSC 1918.}

\section*{New South Wales}

The entry of government into the field of irrigation came after its
decision to establish a major water storage in headwaters of the
Murrumbidgee River \index{river!Murrumbidgee} as a prelude to delivery
of supplies to plains traversed by the river downstream.

The first step taken towards government intervention was in 1896 when
the Water Rights Act vested water rights in the Crown and instituted a
system of licences for dams and diversions from streams for irrigation
by individual landholders.  The immediate effect of the legislation
was to recognise a practice which had been in existence for some time,
often with considerable friction between water-users.  Two landholders
who made early use of the provisions of the Act to divert water for
irrigation were Samuel McCaughey \index{McCaughey, S.} in the
Riverina, and Thomas Ellis on the Hunter River \index{river!Hunter} in
north of the State.  In the longer term, however, the Act provided the
basis for further intervention by the state.\fn{\citet{evatt1938};
\citet[p.\,183]{lloyd1988}.}

Another important step by the government in 1896 was to call on
Colonel Home, \index{Home, Col.} an irrigation engineer recently
retired from service in India, to report on prospects for irrigation
in New South Wales.  Home spent 12 months in the colony, concentrating
attention on the inland rivers and finding only the Murray and
Murrumbidgee offered potential for major irrigation development.  He
reviewed the earlier proposals made by McKinney for development from
these rivers and favoured use of the Murrumbidgee because it was
entirely within the colony. Neither river could sustain major
extension of irrigation without a storage reservoir; he favoured the
Barren Jack site, southwest of Yass, as most appropriate for storage
on the Murrumbidgee.\fn{NSW Rept RC Water Conserv.\ 1885--87; NSW LA
V\&P 1897, vol.\,5 Rept On the Prospects of Irrigation in NSW by Col.\
F.\,J.~Home; \citet[pp.\,178--80, 182--83]{lloyd1988}.}

The serious drought which culminated in 1902 led to a renewal of
interest in irrigation, expressed in several ways.  Passage of the
Water and Drainage Act 1902 provided funds for waterworks, established
the State as a constructing body and authorised establishment of
trusts to administer waterworks.  Irrigation leagues in the Riverina
sought diversion of Murray water by a canal.  Their activity led to
the notable Corowa Conference in 1902, which attracted support from
beyond New South Wales and led to the establishment by three States of
the Inter-State Royal Commission on the Murray Waters.  That inquiry
collected much evidence on development of irrigation as well as use of
the river system for navigation. Samuel McCaughey's success with
irrigation near the Murrumbidgee River was reported by him to the
Commission and gained wide publicity. After becoming owner of the
North Yanco \index{North Yanco estate} estate of 100\,000\,acres in
1899, he had begun extensive irrigation, including 750\,acres of
lucerne and 250\,acres of sorghum.\fn{\citet[pp.\,124--25]{lloyd1988};
\citet[p.\,16]{evatt1938}; VicPP no.\,35 of 1902--03;
\citet{mccaughey1955}.}


Another important proposal at this time was made by Robert Gibson
\index{Gibson, R.} of Hay and Hugh McKinney, \index{KcKinney, H.} the
former chief irrigation engineer, for private irrigation of a large
tract of land north of the Murrumbidgee
River. \index{river!Murrumbidgee} This was identifed at one time as
the Gunbar scheme, referring to the destination of a proposed major
canal to end 60\,miles northeast of Hay.  Gibson's scheme was to
involve irrigation of 150\,000 acres of first-class land and limited
(winter) irrigation of up to 1.25 million acres of natural pasture.
Gibson, a pastoralist, irrigator and land agent, had options on
170\,000\,acres of land in the region.  He acknowledged financial
backing from Victoria and England and probably also had the promise of
support from Samuel McCaughey.  His special interest in the scheme was
that of a land agent with a speculative interest in the purchases to
be made by would-be irrigators.  Hugh McKinney was to be the
irrigation engineer for the project.

Arrangements were made to have Gibson's proposal submitted to
parliament for its approval.  Accordingly the Murrumbidgee Northern
Water Supply and Irrigation Bill was submitted to parliament in
September 1903 by Mr~G.\,S.~Briner~MP. \index{Briner, G.S.} Progress
with the Bill was impeded by changes to the scheme and a change of
government. During this delay it emerged that since 1901 the Public
Works Department had been planning irrigation in the same region;
L.\,A.\,B.~Wade,
\index{Wade, L.\,A.\,B.} its
engineer, explained that a major distinction from Gibson's scheme was
the consideration of irrigation both north and south of the
river.\fn{J.\,A.~Gibson (1962), Proc.\ Hay Historical Soc.; NSW LA
V\&P 1903 vol.\,1, NSW V\&P 1906 vol.\,5. Standing Comm.\ Public
Works, Murrumbidgee Rept \& MoE, MoE p.\,96;
\citet{mccaughey1955};
\citet{lloyd1988}; \citet[p.\,135
re McKinney as the real planner]{munro1974}.}

While Gibson's proposal was still before parliament, the Minister for
Public Works, C.\,A. Lee, \index{Lee, C.A.} convened a conference
early in 1905 to consider irrigation.  It was widely attended; among
the delegates was the engineer John Monash representing the Lower
Murrumbidgee district.  The conference agreed that New South Wales
should begin a comprehensive scheme of water conservation and
irrigation and left it open as to whether a private or State scheme
should be adopted.  By this time L.\,A.\,B.~Wade, the principal
engineer of water supply and sewerage in the public works department,
had visited USA to study irrigation and was preparing plans for the
Burrinjuck reservoir \index{reservoir!Burrinjuck} on the Murrumbidgee
and the use of its water for irrigation beyond the highlands.  Both
Gibson's and Wade's schemes for irrigation to the north of the
Murrumbidgee came before a parliamentary committee in 1906, which
collected evidence in the district under consideration and in Victoria
at Mildura, Kerang, Tatura and Melbourne.

One point already decided was that the government had taken
responsibility for building the storage dam.  The many witnesses
included Samuel McCaughey of North Yanco; Robert Gibson of Hay;
several government officers including L.\,A.\,B.~Wade and Joseph
Davis; and Victorians including Stuart Murray, first chairman of
SRWSC; George Swinburne, MP and Minister for Water Supply; John
Monash, engineer; and W.\,B.~Chaffey of Mildura. This committee
recommended the departmental scheme with some modifications due to
Stuart Murray and declined that proposed by Robert Gibson.  John
Monash appeared for pastoralists whose livelihood depended on periodic
flooding of the Lower Murrumbidgee below Hay; he voiced their claim
that Gibson's scheme would reduce that inundation.  Wade's plans shown
to the committee indicated an area of first-class land north of
Mirrool and Wah Wah Creeks with 196\,000\,acres, and 160\,870\,acres
of second-class land stretching from Yanco westerly towards Hay. These
could be served by two canals diverging from Yanco: the Gunbar canal
\index{canal!Gunbar} to the north and the Hay canal \index{canal!Hay} to the
west.\fn{\citet[p.\,187]{lloyd1988}; R.\,T.~McKay (1907), The
Utilisation of the Murrumbidgee Waters, Agr.\ Gaz.\ NSW,
\textbf{18}; p.,103; NSW PP V\&P Session 1906, Parl.\ Standing Comm.\
Public Works, Rept \& MoE.}

The government acted promptly on the report of the Standing Committee
and successfully introduced legislation authorising construction of
the Burrinjuck dam and the necessary works for irrigation: the Barren
Jack Dam and Murrumbidgee Canals Construction Act 1906.  The Public
Works Department was to be the constructing authority.  Work on the
dam commenced in 1908 and the first supply of water for irrigation was
expected after the dam reached a height of
70\,feet.\fn{W.\,H.~Williamson (1968), A Century of Scientific
Progress, Royal Soc.\ NSW, ch.\,2, p.\,63.}

In 1910 the Murrumbidgee Resumption Act and the Murrumbidgee
Irrigation Act authorised the resumption of 1\,668\,000\,acres of land
for the scheme and its management by a Trust consisting of the
Secretaries (Ministers) for Public Works and for Lands together with
the Minister for Agriculture. This arrangement lasted until the
Irrigation Act of 1912 which led to the appointment of a single
Commissioner for Water Conservation and Irrigation, a position awarded
to L.\,A.\,B.~Wade.\fn{\citet[p.\,127]{frith1974}.}

Resumption of land started in 1911, when 60\,000\,acres of
McCaugh\-ey's North Yanco estate was purchased.  By 1913 more than
300\,000\,acres had been acquired to provide two irrigation areas to
be named Yanco and Mirrool, \index{Yanco, NSW} \index{Mirrool, NSW}
close to the western edge of the highlands and served by a main canal
starting at the Berembed weir \index{weir!Berembed} on the
Murrumbidgee and stretching more than 100 miles to its terminus in the
Mirrool Irrigation Area. \index{irrigation!area!Mirrool} The first
water for irrigation in the Yanco area was delivered in July 1912, and
to the Mirrool area in October 1913.  Towns designed by the American
architect Walter Burley Griffin
\index{Burley Griffin, W.} were established in 1914: Leeton
\index{Leeton, NSW} in the Yanco area named for C.\,A.~Lee, \index{Lee,
C.A.} the Minister for Public Works first associated with the
Murrumbidgee Irrigation Area and Griffith \index{Griffith, NSW} in the
Mirrool area for A.~Griffith, the second
Minister.\fn{\citet[p.\,114]{jeffcoat1988}.}

The Murrumbidgee Irrigation Area (MIA)
\index{irrigation!area!Murrumbidgee} was divided into holdings varying
in size from two acres for workingmen's blocks, to 10\,acres for
horticulture and 50\,acres for mixed farming.  The largest blocks were
expected to be used for production of lucerne and sorghum.  Variation
in natural vegetation on the project area provided a guide to the
soils which ranged from deep sands to heavy clays.  The better-drained
and more porous soils of land regarded as of first-class quality were
intended for horticulture; they were generally associated with stands
of native pine.

Prospective settlers began selecting blocks when these became
available in 1912.  Many of the first settlers were miners from Broken
Hill, some with occupational pulmonary disorders.  Another group
included the distinctive Welsh settlers from Patagonia where they had
gained experience with irrigation; they had been sought initially by a
Victorian agent but were lured to the MIA.  One Italian, though
originally from Switzerland, was allotted a farm in 1914 and two
Spaniards obtained farms in 1916.  A few Italians came to work as
labourers before 1920 and several irrigation blocks were allotted to
American migrants.  The well-known Australian writer Henry Lawson
\index{Lawson, H.}
spent a year in the Yanco area.  Between July 1912 and April 1914, 622
farms covering 24\,000\,acres were allotted on perpetual lease.  At
Leeton a butter factory was opened in 1913, followed by a cannery in
1914.  Besides dairying on the larger holdings and fruit grown on
horticultural blocks, pigs and fat lambs were raised and production
was also diversified by growing tobacco and broom
millet.\fn{\citet{kelly1988}; \citet[pt.\,1]{langfordsmith1966};
\citet[ch.\,14]{lloyd1988}.}

Irrigators in the MIA were required to pay water rates sufficient `to
meet interest on the cost and upkeep of the storage dam and all main
channels for conveyance of water to the main parts of the settlement,
and also the cost of distribution of the water'. A concession provided
that `half water rates are charged for the first year, increasing
until full rates are paid in the sixth year.'\fn{L.\,A.\,B.~Wade
(1914), BAAS Hdbk NSW pp.\,147--48.}

Within a few years there was widespread dissatisfaction with the
results obtained by settlers.  The problem was most serious for those
on the non-horticultural blocks and arose from the failure to secure
yields of lucerne comparable with those which had been reported
earlier by Samuel McCaughey and provided the basis for farm size.
Inquiries were made in 1915 by A.\,C. Carmichael, a Royal
Commissioner, and again in 1916 by Judge W.~Bevan, leading to a
revision of farm size.  Another grievance was the claim that the rent
charged to landholders was based on too high a valuation of the
land.\fn{\citet[p.\,212]{lloyd1988}.}

Irrigation in the MIA was needed to supplement the annual rainfall,
about 16\,inches. on average, particularly during the long dry summers
to support production of fodder crops and fruit.  Initially it was
expected that the basic supply to allotments would be equivalent to a
further 12\,inches. with provision for additional water for particular
crops. Irrigation brought problems in 1917 at Yanco when serious
waterlogging \index{waterlogging} occurred and was the beginning of a
long-held view among settlers that it was the consequence of seepage
from the earthen channels.  However, this experience followed
exceptionally high rainfall in 1916 and could be connected in one
locality with previous irrigation over some years by Samuel McCaughey
on land later incorporated in the new irrigation settlement.
Regulation of water delivery to allotments was the responsibility of
water bailiffs and although some Dethridge meters
\index{Dethridge meter} were made available
in 1911 soon after their invention in Victoria they were apparently
not used before 1920.\fn{L.\,A.\,B.~Wade (1914), BAAS Hdbk NSW,
p.\,147; Information by H.\,N.~England in Appendix~A of 1945 CSIR
report by J.\,A.~Prescott and others on waterlogging in Yanco no.\,1
Irrigation Area; K.~Murley (1967), The Dethridge Meter, SRWSC.}

A major development began during World War I for settlement of war
veterans in the MIA.  Plans were made for 1500 blocks to be available
but little more than 800 were occupied in the early 1920s.  Experience
with this scheme was disheartening and led later to several official
enquiries.\fn{\citet[p.\,215]{lloyd1988};
\citet[p.\,71]{davidson1969}.}

The Curlwaa Irrigation Area \index{irrigation!area!Curlwaa} was
established on 10\,550\,acres under the Wentworth Irrigation Act of
1890 and in 1912 it came under the control of the Water Conservation
and Irrigation Commission (WCIC). \index{WCIC} Its location is on
low-level alluvial land between Tuckers Creek \index{creek!Tuckers}
and the Murray River. Much of the land was unsuitable for irrigation
and in 1916 only 1356\,acres were held on irrigation leases by 67
settlers.  Their blocks ranged from five to 37\,acres and each lessee
was entitled to receive 30\,inches. of water per acre annually by
pumping from the Murray River.  Irrigation was then used mainly for
fruit growing.\fn{CSIR Bull.\ 107, 1937.}

The Hay Irrigation Area \index{irrigation!area!Hay} had an area of
3840\,acres, with almost 1000 acres held under irrigation leases by 81
settlers in 1916. Water was pumped from the Murrumbidgee River to give
each settler a right to 24\,inches of water annually for each acre
incurring the water rate.  Dairying was the main occupation in
1916.\fn{Ann.\ Rept to June 1916 of WCIC, NSW.}

By 1920 the area irrigated in New South Wales included 34\,000\,acres
in settlements controlled by the WCIC and perhaps 50\,000\,acres
undertaken by those licenced by the government to divert water for
irrigation from various streams in the State.  More than 1200 licences
were in force by 1918 for diversions from streams.  The Murray River
was the major supplier for the private diverters, whose pumps in
1919--20 were estimated as delivering up to 40\,000\,acre-feet in New
South Wales.  Although details of irrigation by private licensees are
not shown in annual reports of the WCIC, the areas irrigated in 1920
from all streams in the State are reported elsewhere as totalling more
than 50\,000\,acres, mostly supplied by the Murray, Murrumbidgee,
Lachlan, Hunter, and Namoi.  Accordingly, irrigation by private
diverters was by 1920 much more widespread than that undertaken by
government enterprise.\fn{NSW WCIC annual reports 1918, 1919, 1920;
NSW PP Joint vol.\,4 1920, R.~Murray Commission, NSW Rept to 30 June
1920;
\citet[pp.\,71--81]{hallows1996}.}

\section*{South Australia}

Entry by the South Australian government into irrigation development
was a consequence of policy in favour of closer settlement, subject to
the Crown Lands Act 1897 passed during the term of the Kingston
Government when Peter Gillen and Laurence O'Loughlin had terms as
Commissioner of Crown Lands.  Among the areas chosen for closer
settlement were the swamps along the Murray River downstream from
Mannum.  Earlier efforts at their reclamation included the protection
of a large swamp from flooding by an embankment erected in 1881 by Sir
William Jervois \index{Jervois, W.} who acquired the Wellington
pastoral property during his term as Governor of the colony.
Irrigation was being undertaken later on swamps after their
reclamation by a few landowners including J.~Cowan \index{Cowan, J.}
at Glen Lossie (700\,acres) and H.\,W.~Morphett \index{Morphett, H.W.}
at Woods Point (650\,acres), who along with many others gave evidence
in 1902 to the Interstate Royal Commission on the River Murray aboard
a paddle steamer travelling along the river.\fn{VicPP no.\,35 of
1902--03 vol.\,3.}

The successful reclamation and irrigation of swamps on the lower
Murray led to a proposal for similar treatment for the freshwater Lake
Albert near the mouth of the Murray River.  The scheme was urged by
D.\,J.~Gordon \index{Gordon, D.J.} in the press and in a government
publication with the claim that it would provide an additional
40\,000\,acres of land.  Reference to the proposal was made in 1910 in
the report by Stuart Murray, \index{Murray, S.} the Victorian
engineer, to the South Australian government on the utilization of the
waters of the Murray River. There is no indication of further official
attention to this scheme before 1920.  From 1904 the
Surveyor-General's office became responsible for reclamation and
subdivision of several swamps near Murray Bridge and the Department of
Agriculture established an experimental farm on the Mobilong swamp
\index{swamp!Mobilong} in
1906 with Samuel McIntosh \index{McIntosh, S.} as manager.  He later
became government nominee and chairman of boards of management
controlling these swamps.

Their reclamation relied on the provision of levee banks that
extend\-ed along the river between high land at each end of the
bow-shaped swamp.  These banks were made high enough to give
protection against floods, so to cope with those of the magnitude
known in 1870, a height of almost eight feet was necessary at Murray
Bridge \index{Murray Bridge, SA} instead of the lower embankment used
by Sir William Jervois.  Material for the banks was taken from the
adjacent highland and from the swamps, whose highly organic or clayey
soils contracted on drying and enabled river water to leak into the
swamps.  This defect was corrected when embankments were properly
cored.  Sluice gates installed in the levee banks could be used to
assist drainage during periods of low river level or for irrigation
when the river rose adequately. The Surveyor-General's report for
1909--10 refers to the difficulty of reclamation on certain swamps
where the water let in by settlers who opened sluice gates took days
to remove by pumping.  The main use of the swamps reclaimed for closer
settlement was dairying.  Over the period 1904 to 1909 the government
undertook reclamation of the swamps known as Mobilong, Burdett, Long
Flat, Monteith and Mypolonga.  By 1920 their combined area of
3800\,acres provided 113 holdings varying in size up to
50\,acres.\fn{D.\,J.~Gordon, (1908), Handbook of South Australia; SAPP
29 of 1910, S.~Murray's report; J.\,K.~Taylor \& H.\,G.~Poole (1932),
Report on the Soils of the Bed of Lake Albert, SA, J.\,CSIR
\textbf{5}, pp.\,129--130; S.~McIntosh (1912), J.\,Agric.\ SA
\textbf{15}, Irrigation and Reclamation, March p.\,812; A.\,J.~Perkins
(1903), J.\,Agric.\ SA, \textbf{6}.}

\index{swamp!Mobilong}\index{swamp!Burdett}\index{swamp!Long
Flat}\index{swamp!Monteith}\index{swamp!Mypolonga}

The Irrigation and Reclamation Lands Act of 1908 enabled the
government to begin extension of irrigation along the Murray upstream
from Morgan.  \index{Morgan, SA} The legislation provided for
establishment of government irrigation areas containing individual
allotments under perpetual leasehold tenure, with 50\,acres as the
maximum irrigable area to be held by one person.  The government
proceeded to use its powers in two ways.  Irrigation still being
carried on at remnants of village settlements became the basis of
irrigation areas under its control.  The Waikerie Irrigation Area
\index{irrigation!area!Waikerie}  was
thus established in 1909 with 66 holdings including 42 new ones on
adjacent highland where water supply channels were made and supplied
by improved pumping facilities.  Similarly the King\-ston and Moorook
Irrigation Areas provided for irrigators once involved in village
settlements at these localities; and the Cadell Irrigation Area (1916)
was established on land briefly occupied in the 1890s by the New Era
village settlement.\fn{B.\,J.~Menzies \& P.\,N.~Gray (1983),
Irrigation and Settlement in the South Australian Riverland, Tech.\
Paper Dept.\ Agric.\ SA no.\,7, p.\,191; SAPP 86 of 1913, p.\,4;
\citet[p.\,56]{mack1994}; T.\,J.~Marshall \& N.\,J.~King (1932), CSIR
Bull.\ no.\,62.}

\index{irrigation!area!Cadell}\index{irrigation!area!Kingston}
\index{irrigation!area!Moorook}

The other path taken by the government was to irrigate virgin land
near the river.  Extensive areas of lowland then held under pastoral
leasehold were available north of the river between Renmark and
King\-ston.  One of these near Lake Bonney \index{lake!Bonney} had
been under consideration from the 1880s for private or government
irrigation.  A start was made, however, at a river-bank locality now
known as Berri \index{Berri, SA} which had once been considered for a
village settlement.  By 1908 when the new enterprise began, a moderate
area of lowland flood plain accommodated a wood-lot to supply
steamboats together with irrigated production of fruit and vegetables
for the passing traffic.  This segment of the flood plain is backed by
sandy slopes rising to an undulating plain then covered by mallee
vegetation.

Attention was first given to the lower ground; it compared with land
irrigated successfully at Renmark and some village settlements by
means of centrifugal pumps suitable for the low lifts involved.  The
survey work led to a start in 1909 on the construction of irrigation
channels.  Pumping of water began in 1910 when the Berri Irrigation
Area was proclaimed. \index{irrigation!area!Berri} The irrigated area
spread northwest and west from Berri township before and during the
1914--18 war. Preparations were made for the irrigation of further
allotments at a distance from the river; these were reserved for
returned servicemen.  The provision of additional irrigation blocks
involved clearing and subdivision of the higher country covered by
mallee vegetation before construction of water channels.  One
complication was the need to exclude certain tracts not irrigable by
gravitation from open channels at the highest levels commensurate with
pumping equipment to be installed; another was the requirement that
subdivision of the commandable area should provide allotments
equivalent in value rather than size.  This latter consideration was
important in view of the variations represented by the pine-clad sandy
slopes facing the river and by the pattern of east-west sandy ridges
alternating with loamy swales.\fn{\citet[p.\,4]{andison1953};
B.\,J.~Menzies \& P.\,N.~Gray, (1983), p.\,221.}

Another irrigation settlement started by the South Australian
government involved a large tract of land west of Berri used by John
Chambers \index{Chambers, J.} as part of his vast Cobdogla
\index{Cobdogla, SA} pastoral run; it included extensive flood plains
along the river backed by higher land with mallee scrub.  Surveyors
moved to this area in 1911 after finishing their work at Berri; they
were followed by gangs clearing the land and making channels.
Topographic surveys showed that tens of thousands of acres, mainly in
the scrub-covered highland rising more than 100\,feet above the river,
were available for irrigation.  The first effort was to make use of
the flood plain between the river and Lake Bonney; it could be
supplied by a pump lift of 30 to 40\,feet and earthen channels. The
intention was to raise fodder crops and lucerne there as the basis for
dairying.  The Cobdogla Irrigation Area
\index{irrigation!area!Cobdogla} was proclaimed in 1916 and land in
this lower part, known as the Cobdogla Division, became available in
1918 for allotment to settlers. Next year the Cobdogla township was
proclaimed.\fn{\citet[p.\,30]{woolmer1973}.}

Attention then turned to development of neighbouring areas for
occupation later by ex-servicemen.  One portion known as the Nookamka
Division was provided with a town known firstly as Lake Bonney South
and later as Barmera. \index{Barmera, SA} Another formed the Loveday
Division in which the Loveday township appeared later.  There was
ample scope for extension of irrigation to more than 10\,000\,acres in
the Weigall Division at the south and the McIntosh Division lying
north of Lake Bonney.  \index{lake!Bonney} The good returns available
in the war period for dried fruit supported optimism for expansion of
irrigation in the Cobdogla Irrigation Area for vines and fruit.

The Verran government installed in June 1910 took a very positive
attitude to use of the Murray River for both navigation and
irrigation.  Crawford Vaughan, \index{Vaughan, C.}  the new
Commissioner of Crown Lands gave attention to irrigation by creating
the Irrigation and Reclamation Works department with Samuel McIntosh
as director and staff drawn from the Surveyor-General's office.
McIntosh \index{McIntosh, S.} then had almost twenty years of
experience along the river, including work at Renmark, advising the
village settlements, managing an irrigation company near Waikerie,
\index{Waikerie, SA} and
helping closer settlement of the reclaimed swamps near Murray
Bridge.\index{Murray Bridge, SA}

McIntosh had returned from an overseas tour in 1911 with two important
objectives realised eventually at Cobdogla.  The volume of water
expected to be required there for irrigation warranted the
installation of a powerful and economic pumping plant; the best
available at the time appeared to be the distinctive Humphrey Internal
Combustion Pump. \index{pump!Humphrey} An order was placed for two
units but delivery was frustrated by the world war and installation
was not made before 1920.  His other idea was to follow the example of
some Californian irrigation colonies and provide a pressurised piped
supply for irrigation.  McIntosh learnt of the recent invention in
South Australia by W.\,R.~Hume \index{Hume, W.R.} of concrete pipes
manufactured by recourse to a centrifugal method; he found that these
pipes had been installed in 1911 for irrigation on the riverside
property known as Murray Heights near Ramco.  The Hume pipes
\index{Hume pipes} were used by the government only in the Loveday
Division, but not before 1920.

Besides the development of the Berri and Cobdogla areas, the
government decided to extend settlement at Renmark by the development
of its Block~E and creation under government control of the new
Chaffey Irrigation Area \index{irrigation!area!Chaffey} having the
single Ral Ral Division.

The settlers at the various irrigation settlements were required to
pay annual rates for the use of irrigation water.  The rate in each
irrigation settlement was fixed by reference to the unimproved value
of the land and included a proportion of interest on the cost of the
works and of their maintenance and management.  The South Australian
settlements on the Murray involved pumping from the river but had no
requirement for storage of water.\fn{SA Irrigation and Reclamation Act
no.\,1178 of 1914, Section 32.}

Although other irrigation schemes were considered for action by the
government, the only one not dependent on supplies from the Murray
River was the Pekina scheme \index{irrigation!scheme!Pekina} in the
southern Flinders Ranges, not far from the township of
Orroroo. \index{Orroroo, SA} In this district there was interest in
establishing a reservoir in the hills to hold water drained by Pekina
Creek \index{creek!Pekina} so as to irrigate land on a nearby plain
where the average annual rainfall is 13\,inches.  The scheme involved
an embankment 70\,feet high for a reservoir with a capacity of 220
million gallons to be delivered through more than seven miles of
underground main to the irrigation area of 429\,acres divided into 45
blocks, each with an annual allowance of 12\,inches of water and
equipped with a water meter.  The technical development of irrigation
was undertaken by the engineering staff of the Public Works
Department, and an experimental farm undertaken by the Ministry of
Agriculture was managed by Samuel McIntosh.  Construction began in
1908 and was completed in 1910, when lucerne and fodder crops were
sown with the object of helping the dairying industry.  The Pekina
scheme was developed under the Water Conservation Act of 1886 by the
Public Works Dept and was later leased to the municipal authority of
the district.\fn{S.~McIntosh (1912), J.\,Agric.\ SA;
\citet[p.\,56]{hammerton1986}.}

A feature of irrigation development in South Australia before 1920 was
the responsibility of the Commissioner for Crown Lands.  Until 1910
the Surveyor-General's office provided survey and engineering staff
for the reclamation and irrigation works along the Murray river.  When
the Irrigation and Reclamation Department was set up in 1910, it
required detachment of staff from the Surveyor-General's department,
with both departments responsible to the Commissioner of Crown Lands.
G.\,W.~Goyder, \index{Goyder, G.\,W.} the Surveyor-General for 25
years up to his retirement in 1894, had been involved with many
aspects of land utilisation in South Australia, including areas
adjacent to the Murray River.  Long after his departure, the
Surveyor-General's office continued to have responsibility for
irrigation development to an extent much greater than in other States.

Irrigation in South Australia relied on government support for
settlements to be provided with water from the River Murray.  Private
diversions from that stream probably began after the collapse of
several village settlements of the 1890s, and by 1912 several farms
were irrigating orchards, vineyards or fodder crops along the river.
However even when private diversions were regularised by the passage
of the Control of Waters Act in 1919, there were less than 1000\,acres so
irrigated in 1921. Private irrigation using groundwater was a minor
activity on the Adelaide Plains before 1920.\fn{S.~McIntosh (1912),
J.\,Agric.\ SA, March, p.\,810; S.~McIntosh in SAPP 32 of 1921, cited
in Menzies \& Gray, 1983, p.\,201.}

\section*{Western Australia}

The earliest irrigation undertaken by the government was the
installation in 1908 of an irrigation system to grow lucerne on the
Brunswick State Farm near Bunbury. \index{Bunbury, WA} The next
development was prompted by the need for irrigation of citrus groves
near the Harvey River \index{river!Harvey} south of Perth. A scheme
for this area involving construction of a dam on the Harvey River was
proposed in 1912 when the government first attempted to gain
legislation known as the Rights in Water and Irrigation Act 1914. Work
started in 1915 to provide storage of 520 million gallons for
irrigation of up to 3000\,acres at the foot of the Darling Range.  The
Harvey Irrigation District \index{irrigation!scheme!Harvey} which
opened in 1916, after work by the Public Works Department and the
Water Supply Department, was the sole entry by the Western Australian
government for irrigation development before 1920.  Hugh Oldham,
\index{Oldham, H.} who had worked earlier at Renmark and Mildura, was
an engineer of the Water Supply Department who referred to the Harvey
scheme in 1913 and in 1918 to the Royal Commission on Agricultural
Industries of the South-west.\fn{WA V\&P 1917--18 Paper 15, 2nd Prog.\
Rept RC~Agric.\ Industries; WAPD 1912 vol.\,43, p.\,1853, 1913
vol.\,449; H.~Oldham (1913), WA V\&P 1917--18, vol.\,2, Paper 15, 2nd
Prog.\ Rept RC~Agric.\ Industries.}

\section*{Queensland}

Irrigation of sugar cane \index{sugar cane} by private companies
continued to be a feature of agriculture in Queensland following its
use on the Burdekin River \index{river!Burdekin} delta at
Ayr. \index{Ayr, Qld} Efforts to introduce irrigation to the Bundaberg
\index{Bundaberg, Qld} district began in 1889 when John Cran
\index{Cran, J.} of the Millaquin refinery negotiated with the Chaffey
Brothers for irrigation of land at Woongarra. That scheme was
abandoned but success attended other moves to irrigate sugar cane in
the district after the arrival in 1900 of Dr~Maxwell, the recently
appointed director of the Sugar Experiment Stations who had experience
of irrigation in the Hawaiian sugar plantations.  The Bingera
plantation of Gibson and Howes began to use irrigation in 1903, with
water pumped from the Burnett River,
\index{river!Burnett} and in the same year Young Brothers were
supplying their Fairymead plantation with underground water in the
same way as on the Burdekin delta.  A revival of interest in
irrigation of sugar cane in the Woongarra area occurred in 1903 with
an unsuccessful proposal to pump water from the Elliott River to
3000\,acres.\fn{\textsl{Brisbane Evening Observer}, 6 Apr.\ 1889;
H.\,E.\,A.~Eklund (1923), Qld Agric.\ J.}

Queensland governments gave attention to the underground water
resour\-ces both of the inland regions and in coastal areas where they
might be exploited for sugar cane irrigation, but no direct government
involvement in irrigation settlements was undertaken before 1920.
However, after a visit by Elwood Mead \index{Mead, E.} to Queensland
in 1909, the government followed the recommendations in his report for
legislation known as the Rights in Water and Water Conservation and
Utilisation Act of 1910, which provided state control of water
resources.\fn{Qld J\&P 1910 vol.\,3, Rept on Water Resources of
Queensland, by E.~Mead; Qld PD vol.\,107, 29 Nov.\ 1910, p.\,2356;
\citet{hallows1996}.}

By 1920 the total area reported as irrigated in Queensland rose to
more than 10\,000\,acres, mostly for production of sugar cane in the
Ayr and Bundaberg districts.  In the Cunnamulla, Hungerford,
Rockhampton, Bowen and Townsville districts irrigation involved only a
few hundred acres.  No irrigation works were constructed by the
government before 1920.\fn{Qld Agr.\ J. 1924, Irrigation Statistics,
pp.\,289--308.}

\section*{Tasmania}

The main concern of Tasmanian governments with irrigation during the
early years of the new century continued to be with the use of waters
held in the adjacent Lakes Sorell \index{lake!Sorell} and Crescent.
\index{lake!Crescent} Earlier efforts by the government to divert
sufficient water from these lakes to irrigate land in the Midlands had
failed, leaving the Clyde Water Trust as the sole beneficiary of water
in Lake Crescent\,---\,the source of the Clyde
River. \index{river!Clyde} Then several outbreaks of typhoid fever
along that river during the early 1890s led to criticism of the
Trust's control of water and to renewed consideration of sharing lake
water between landholders in the Clyde valley and those in the
Tunbridge district of the Midlands.\fn{\citet[pp.\,152--54]{masoncox1994}.}

The government then legislated to recast the Clyde Water Trust,
\index{trust!water!Clyde} by the
Clyde Water Act of 1898, and to form the Midland Water Trust, under
the Midland Water Act of 1898.  The latter trust would control water
supply, in excess of the Clyde Trust's quota, for irrigation and other
purposes in the Midlands.  When the Midland Trust agreed in 1901 to
supply water to a company for generation of electricity to manufacture
calcium carbide, the Minister for Land and Works intervened and began
negotiations with another company for a hydro-electric
\index{hydro-electricity} scheme based on
use of the lake waters.  One consequence was further
legislation\,---\,the Lakes Sorell and Crescent Conservation Act
1901\,---\,which safeguarded the daily entitlement of water for the
Clyde valley, made the delivery of water to the Midlands conditional
on an adequate supply, and gave the Minister authority to carry out
conservation works at the Lakes.\fn{\citet[p.\,151]{masoncox1994}.}

Subsequent gaugings of rainfall and lake discharge indicated an
inadequate supply of water during dry years, with the result that
proposed conservation works were abandoned, irrigation development by
the Midlands Water Trust \index{trust!water!Midlands} was frustrated,
and the Clyde Water Trust continued to be an important irrigation
authority.\fn{TasPP no.\,46 of 1902, Rept by Rahbek.}

Hydro-electric power generation began before 1900 and was undertaken
mainly by mining companies.  A major project based on the Great Lake
\index{lake!Great} was begun in 1911 by a company requiring power for
processing zinc ore, and its first operation in 1916 came after sale
of the company to the government and creation of its hydro-electric
department directed by J.\,H.~Butters, \index{Butters, J.H.} formerly
chief engineer of the company.  Coincidentally, low rainfall over
several of these years led to a revival of interest in irrigation and
proposals for its integration with hydro-electic power generation.
One example was the scheme suggested by W.\,E.~Shoobridge
\index{Shoobridge, W.E.} for use of
waters held in Arthurs Lakes \index{Arthurs Lakes, Tas.} in the
central highlands to generate power and provide irrigation in the Lake
River
\index{river!Lake} valley.  As with earlier proposals concerning Lakes
Sorell and Crescent, Shoobridge's suggestion failed because of
uncertainty over rainfall and lake resources, as indicated in reports
by visiting irrigation engineers. In 1919 he called unsuccessfully for
legislation giving full control of Tasmanian water resources to the
State.  By 1920 the hydro-electric department was heavily involved in
collecting data on water resources and their use for power generation,
while public control of irrigation was limited to the Clyde Water
Trust and the Macquarie Water Trust \index{trust!water!Macquarie} set
up in 1892 for management of \index{lake!Tooms} Tooms
Lake.\fn{\citet[p.\,170]{masoncox1994}; TasPP nos.\,11 of 1916, 28 of
1917.}\index{trust!water!Clyde}


\closure
Promotion of irrigation by governments became decisive in Victoria,
New South Wales, and South Australia by 1910, a few years earlier than
similar development in Western Australia.  Progress was made in
Queensland without state control, while in Tasmania state involvement
with water resources concerned only hydro-electric power generation.

Commercial failure of the irrigation settlements that were launched by
the Chaffeys, the fiasco produced by the Victorian irrigation trusts,
and the experience of widespread drought during 1902 led Victorian
governments to accept greater responsibility for irrigation.  The
first step was the creation of the SRWSC, the second was the
appointment in 1907 of Elwood Mead as its chairman, a position he used
to establish several new irrigation settlements.

Special government authorities to deal with irrigation were
established later in New South Wales and South Australia to launch and
control major irrigation settlements dependent on supplies from the
River Murray or its tributaries.  Engineers and surveyors were the
mainstay of this work which included the provision of irrigation farms
for returned servicemen after World War~I.  Numerous agricultural and
horticultural problems confronting settlers new to irrigation soon
taxed the resources of the government departments of agriculture.

%\section*{References}
%1. J.M.Powell, Watering The Garden State,1989, p.121.
%2. S.Murray 1889, quoted by L.R.East in Aqua vol.9, 1958, p.248.
%3.  J.M.Powell 1989, Fig.35,p.148.
%4.  J.M.Powell 1989 p.150 re W.S.Manifold \& VicPD vol.110
%      30/8/1905, p.1247-8.
%5.  S.Murray, Irrigation In Victoria, A Report for the Hon. Geo. Graham, 1892, 
%      L.R.East Vic Hist. Mag. 1967 p.182 and A.Deakin, Yearbook Of Australia, 
%      1892, p.9
%6.  C.G.McCoy, The Supply Of Water For Irrigation In Victoria From 1881
%       to 1981, 1988 p.54 \& Ann.Rept SRWSC 1906-07.
%7.  VicPP Annual Rept SRWSC 1914-15, also  E.Mead 1914 , BAAS Hdbk
%      Vic, p.266. 
%8.  L.R.East, Vic.Hist.Mag, 1967 vol.38, p.205, \& R.F.McNab, 
%     The Early Settlement And Water Supply Of The Wimmera And 
%     Mallee, 1944.
%9.  Vic Water Act 1905,1909, and 1915.
%10.  Annual rept. SRWSC 1908, \& J.Rutherford in J.M.Powell(ed.) 
%       The Making Of Rural Australia, 1974,p.126.
%11. S.Whitehead, BAAS Hdbk Vic1914, p.95.
%12. CSIR Bull. No.123, Merbein Rept.
%13. E.Mead , BAAS Hdbk Vic 1914 p.265
%14. S.Wells, Paddle Steamers To Cornucopia 1986 p.180; E.Mead BAAS 
%      Hdbk Vic1914; A.J.McIntyre, Sunraysia, 1948, \& CSIR Bull. 123, 1939, 
%      Merbein  Rept .
%15. E.Mead 1914.
%16. CSIR Bull. No 73, 1933.
%17. D.C.S.Sissons in ADB vol.12,p.163,  Lesley Scholes, A History Of
%      The Shire Of Swan Hill, 1989, p.117, \& The Australian Encyclopedia,
%      1927, article on rice.
%18. T.Smith, 1916, J.Agric Vic. vol.14 p.493. 
%19. C.G.McCoy 1988 p.16.
%20. E.Mead 1914 p.265.
%21. Aqua Jan 69, K.N.James, Werribee, 1985 p.86, 90, 97.
%22. R.Broome, Arriving, 1984, p.131.
%23. R.Broome, 1984,  p.133
%24. Lesley Scholes 1989 p.116.
%25. Helen Coulson, Echuca-Moama On The Murray, 1995, \& J.Trevaks,
%      Shepparton, A Successful Experiment In Settling Jews On The Land
%      In Australia, Australian Jewish almanac, 1937, \& E.Mead 1914 p.265.
%26. Lesley Scholes, 1989 p.124.
%27. Melbourne Argus, 20/3/1913, p.15.
%28. VicPP R.C.Closer Settlement 1916, final rept and Argus 20/3/1913.
%29. Jenny Keating, The Drought Walked Through, 1992, p.99, \& H.W.Forster,
%     Waranga, 1965  p.97.
%30. C.G.McCoy,1988, p.14, \& A.J. Tisdall, Irrigation In Victoria, in
%       Australian Academy of Science, Water Resources Use And
%       Management, 1964.
%31. J.N.Churchyard, Aqua 1960 v.11(12), pp.259-268.
%32. J.N.Churchyard Aqua 1960 \& CSIR Bull. 137.
%33. E.Mead 1914.
%34. VicPP No 32 of 1925, \& Lesley Scholes, 1989.
%35. I.Meacham  Aqua , 1961,p.184.
%36. C.G.McCoy,1988,  p.54-55.
%37. E.Mead 1914 \&Ann.Rept.SRWSC 1918.
%38. G.J.Evatt , Public Administration 1938, vol 1(2), pp.16-31.
%39. C.J.Lloyd, Either Drought Or Plenty, 1988, p.183.
%40. NSW Rept R.C.Water Conserv.1885-87 \& NSW LA V\&P 1897,vol.5 
%      Rept.on the prospects of irrigation  in NSW by Col.F.J.Home. 
%41. C.J. Lloyd , 1988, pp.178-80, 182-3.
%42. C.J.Lloyd, 1988,  pp.124-5.
%43. G.J.Evatt, Public Administration, vol 1(2),1938, p.16.
%44. VicPP 35 of 1902-03.
%45. Patricia McCaughey,  Samuel McCaughey, A Biography, 1955.
%46. J.A.Gibson 1962, Proc.Hay Historical Soc.,
%47. NSW V\&P 1906 vol.5.Standing Comm.Public Works,  Murrumbidgee 
%      Rept \& MoE, MoE p.96. 
%48. Patricia McCaughey 1955 \& C.J.Lloyd 1988.
%49. NSW LA V\&P 1903 vol.1, \&  C.H.Munro, Australian Water 
%      Resources And Their Development 1974 p.135 re McKinney as the real 
%      planner.
%50. C.J.Lloyd, 1988 p.187.
%51. R.T.McKay, The Utilisation Of The Murrumbidgee Waters, 
%     Agr.Gaz.NSW., vol. 18, 1907,p.103.      
%52. NSW PP V\&P Session 1906, Parl.Standing Comm Public Works,Rept. \&
%      MoE. 
%53. NSW PP 1906  vol.5. 
%54. W.H.Williamson, A Century Of Scientific Progress, Royal Soc.
%      New South Wales, 1968 Ch. 2, p.63.
%55. H.J.Frith \& G.Sawer(eds), The Murray Waters, 1974 p.127.
%56. K.Jeffcoat, More Precious Than Gold, 1988 , p.114.
%57. B.M.Kelly, From Wilderness To Eden, 1988.
%58. C.J.Lloyd, 1988.
%59. T.Langford -Smith \& J.Rutherford, Water and Land, 1966 
%      Part 1 \& C.J.Lloyd, 1988, Ch. 14.
%60. L.A.B.Wade, 1914 BAAS Hdbk NSW p.147-48.
%61. C.J.Lloyd,1988,  p.212.
%62. L.A.B.Wade 1914, BAAS Hdbk NSW,  p.147.
%63. Information by H.N.England in Appendix A of 1945 CSIR report by
%      J.A.Prescott and others on waterlogging in Yanco No.1 Irrigation Area.
%64. K.Murley, The Dethridge Meter, SRWSC 1967.
%65. C.J.Lloyd,1988,  p215, \& B.R.Davidson,Australia Wet Or Dry?,1969 p.71.
%66. CSIR Bull. 107,1937.
%67. Ann.Rept to June 1916 of WCIC, NSW. 
%68. NSW WCIC annual reports 1918,1919,1920.
%69. NSW PP Joint Vol 4 1920, R.Murray Commission , NSW 
%      report to 30/6/1920.
%70. P.J.Hallows \& D.G.Thompson , The History Of Irrigation In
%      Australia, 1996(?), pp.71-81.
%71. VicPP No 35 of 1902/03 vol 3.
%72. D.J.Gordon, Handbook Of South Australia, 1908. 
%73. SAPP 29 of 1910, S.Murray's report.
%74. J.K.Taylor \& H.G.Poole, Report On The Soils Of The Bed Of Lake 
%      Albert, South Australia, J. CSIR 1932 vol.5,pp129-130.
%75. S. McIntosh J.Agric SA vol. 15 Irrigation And Reclamation, March 1912 
%      p.812, \& A.J.Perkins J.Agric.SA ,vol.6, 1903.
%76. B.J.Menzies \& P.N.Gray, Irrigation And Settlement In The South 
%      Australian Riverland, Tech.Paper Dept. Agric. S.A.No.7, 1983, p.191.
%77. SAPP 86 of 1913, p.4. 
%78. D.Mack, The Village Settlements On The River Murray In South Australia 
%      1894-1909, 1994,  p.56 and T.J.Marshall \& N.J.King, CSIR Bull. No.62,
%      1932.
%79. Mrs L.M.Andison, Berri, Hub Of The Upper Murray,1953 p.4.
%80. B.J.Menzies \& P.N.Gray, 1983,p.221.
%81. G.Woolmer, The Barmera Story, 1973 p.30.
%82. SAIrrigation and Reclamation Act no 1178 of 1914, Section 32. 
%83. S.McIntosh J.Agric.SA 1912. 
%84. Marianne Hammerton, Water South Australia, 1986, p.56.
%85. S.McIntosh J.Agric SA March 1912 , p.810.
%86. S.McIntosh in SAPP32 of 1921, cited in Menzies \& Gray , 1983, p.201.
%87. WA V \& P 1917-18 Paper 15, 2nd Prog.Rept. R.C. Agric. Industries.
%88. WAPD 1912 Vol.43, p.1853, \& WAPD 1913 vol.449.
%89. H.Oldham 1913, WA V \& P 1917-18, vol 2, Paper 15, 2nd Prog.Rept 
%      R.C.Agric Industries.
%90. Brisbane Evening Observer, 6 April 1889. 
%91. H.E.A.Eklund Qld Agric.J.1923. 
%92. H.E.A.Eklund 1923.
%93. Qld J \& P 1910 vol 3, Rept.on . . . water resources of Queensland,
%      by E.Mead,\& Qld PD vol. 107, 29/11/1910, p.2356, \& P.J.Hallows \& 
%      D.C.Thompson 1996(?).
%94. Qld Agr.J. 1924, irrigation statistics, pp.289-308.
%95. Margaret Mason-Cox, Lifeblood Of A Colony, 1993,pp.152-54.
%96. Margaret Mason-Cox, 1993 p.151. 
%97. TasPP No.46 of 1902, Rept by Rahbek.
%98. Margaret Mason-Cox, 1993, p.170.
%99. TasPP No.11 of 1916 \& No.28 of 1917.

