% $Id$
%
% To make this, latex selectbibgen. Then edit in items from
% selectbibgen.bbl by hand, stripping the natbib-like stuff.  Ugly.

\def\refitem{\medskip\noindent\hangindent=2em}

\fancyhead[RE]{\small \sffamily Select Bibliography}
\chapter{Select Bibliography}

\refitem
J.\,A. Alexander, {\em The Life of {George} {Chaffey}: A Story of Irrigation
  Beginnings in {California} and {Australia}\/} (Melbourne: Macmillan, 1928).

\refitem
B.\,R. Davidson, {\em Australia Wet or Dry?\/} (Melbourne University Press,
  1969).

\refitem
A. Deakin, 'Irrigation in {Australia}', in {\em Yearbook of Australia\/}
  (Sydney, 1892).

\refitem
H.\,J. Frith, \& G. Sawer eds., {\em {The} {Murray} Waters\/}
  (Sydney: Angus \& Robertson, 1974).

\refitem
P.\,J. Hallows, \& D.\,G. Thompson, {\em The History of Irrigation in
  {Australia}\/} (Mildura: Australian National Committee on Irrigation
  and Drainage, 1996).

\refitem
M. Hammerton, {\em Water {South} {Australia}\/} (Adelaide: Wakefield
Press, 1986).

\refitem
E. Hill, {\em Water into Gold\/}, 3rd edn., Reprint, Angus \&
  Robertson 1965 (Melbourne: Robertson \& Mullins, 1937).

\refitem
C.\,J. Lloyd, {\em Either Drought or Plenty\/} (Sydney: Dept.\ Water
  Resources, NSW, 1988).

\refitem
D. Mack, {\em The Village Settlements on the {River} {Murray} in
  {South} {Australia} 1894--1909\/} (Adelaide: David Mack, 1994).

\refitem
C.\,S. Martin, {\em Irrigation and Closer Settlement in the {Shepparton}
  District 1836--1906\/} (Melbourne University Press, 1955).

\refitem
M. Mason-Cox, {\em Lifeblood of a Colony: A History of Irrigation in
  {Tasmania}\/} (Hobart: Rivers and Water Supply Commission, 1994).

\refitem
C.\,G. McCoy, {\em The Supply of Water for Irrigation in {Victoria}
  from 1881 to 1981\/} (Melbourne: Rural Water Commission of Victoria,
  1988).

\refitem
J.\,M. Powell, {\em Watering the Garden State: Water, Land and
  Community in {Victoria}\/} (Sydney: Allen \& Unwin, 1989).

\refitem
S. Priestley, {\em Echuca: A Centenary History\/} (Brisbane:
Jacaranda, 1965).

